### 2014-04-16 paper/main.tex
Added a few more metrics, cleaned up the oracle (which I don't think I will be useing) updated eval and chart.R a lot.

OUTLINE

1. Introduction
  [story]
    New user experiance is important
      new users are needed to maintine usage of a site as older users leave the system
      all users are new users at some time
      new user experiance may set the expectation for how the system behaves in the future. (Citation of this may be important)

2. Background
3. Methodology
4. Experiments %Not a fan of this title.
  4.1 RMSE by profile size
  4.2 Soft metrics
  4.3 Value of implicit information
5. conclusions

### 2014-04-16 lenskit/src/eval/chart.R

@@ -6,9 +6,11 @@ library("ggplot2")
 all.data <- read.csv("eval-user.csv")
 
 serr = function(d) {qnorm(0.975)*sd(d)/sqrt(length(d))}
+nonMetrics = c("Algorithm","Retain", "DataSet")
+metrics = setdiff(names(all.data),nonMetrics)
 
-for (metric in c("Coverage","RMSE","MAE","nDCG","Information.ByUser", "TopN.nDCG", "TopN.ActualLength", "TopN.avgPop", "TopN.diversity", "TopN.precision", "TopN.recall", "TopN.Independent.Recall")) {
-  plot = ggplot(all.data, aes_string(x="DataSet", y=metric, color="Algorithm"))+stat_summary(fun.y=mean, geom="line") + stat_summary(fun.data = mean_cl_boot, geom="errorbar")
+for (metric in metrics) {
+  plot = ggplot(all.data, aes_string(x="Retain", y=metric, color="Algorithm"))+stat_summary(fun.y=mean, geom="line") + stat_summary(fun.data = mean_cl_normal, geom="errorbar")
   filename = paste(metric,".pdf",sep="")
   cat("Outputting to",filename,"\n")
   pdf(filename, width=11, height=8.5)
@@ -18,7 +20,7 @@ for (metric in c("Coverage","RMSE","MAE","nDCG","Information.ByUser", "TopN.nDCG
 
 
 dat <- read.csv("eval-results.csv")
-plot = ggplot(dat, aes(x=DataSet, y=TopN.pop.entropy, color=Algorithm))+geom_point()+stat_summary(fun.y=mean, geom="line")
-pdf("topN.pop.entropy.pdf", width=11, height=8.5)
+plot = ggplot(dat, aes(x=Retain, y=TopN.Entropy, color=Algorithm))+geom_point()+stat_summary(fun.y=mean, geom="line")
+pdf("topN.entropy.pdf", width=11, height=8.5)
 print(plot)
 dev.off()

### 2014-04-16 lenskit/src/main/java/TopNDiversityMetric.java

@@ -1,40 +1,22 @@
-/*
- * LensKit, an open source recommender systems toolkit.
- * Copyright 2010-2013 Regents of the University of Minnesota and contributors
- * Work on LensKit has been funded by the National Science Foundation under
- * grants IIS 05-34939, 08-08692, 08-12148, and 10-17697.
- *
- * This program is free software; you can redistribute it and/or modify
- * it under the terms of the GNU Lesser General Public License as
- * published by the Free Software Foundation; either version 2.1 of the
- * License, or (at your option) any later version.
- *
- * This program is distributed in the hope that it will be useful, but WITHOUT
- * ANY WARRANTY; without even the implied warranty of MERCHANTABILITY or FITNESS
- * FOR A PARTICULAR PURPOSE. See the GNU General Public License for more
- * details.
- *
- * You should have received a copy of the GNU General Public License along with
- * this program; if not, write to the Free Software Foundation, Inc., 51
- * Franklin Street, Fifth Floor, Boston, MA 02110-1301, USA.
- */
 
 import com.google.common.base.Preconditions;
-import com.google.common.collect.ImmutableList;
+import org.grouplens.lenskit.Recommender;
+import org.grouplens.lenskit.collections.CollectionUtils;
 import org.grouplens.lenskit.core.LenskitRecommender;
 import org.grouplens.lenskit.eval.Attributed;
 import org.grouplens.lenskit.eval.data.traintest.TTDataSet;
-import org.grouplens.lenskit.eval.metrics.AbstractTestUserMetric;
-import org.grouplens.lenskit.eval.metrics.TestUserMetricAccumulator;
+import org.grouplens.lenskit.eval.metrics.AbstractMetric;
+import org.grouplens.lenskit.eval.metrics.ResultColumn;
 import org.grouplens.lenskit.eval.metrics.topn.ItemSelector;
 import org.grouplens.lenskit.eval.metrics.topn.TopNMetricBuilder;
 import org.grouplens.lenskit.eval.traintest.TestUser;
 import org.grouplens.lenskit.knn.item.ItemSimilarity;
 import org.grouplens.lenskit.knn.item.model.ItemItemBuildContext;
 import org.grouplens.lenskit.scored.ScoredId;
+import org.grouplens.lenskit.util.statistics.MeanAccumulator;
 import org.grouplens.lenskit.vectors.SparseVector;
 
-import javax.annotation.Nonnull;
[…]

### 2014-04-16 lenskit/src/main/java/coldstartrecommendation/DownsampleDataSet.java

@@ -1,0 +1,164 @@
+package coldstartrecommendation;
+
+
+import com.google.common.base.Preconditions;
+import com.google.common.collect.Lists;
+import com.google.common.hash.HashCode;
+import com.google.common.io.Closer;
+import it.unimi.dsi.fastutil.longs.*;
+import org.grouplens.lenskit.collections.CollectionUtils;
+import org.grouplens.lenskit.cursors.Cursor;
+import org.grouplens.lenskit.cursors.Cursors;
+import org.grouplens.lenskit.data.dao.EventDAO;
+import org.grouplens.lenskit.data.dao.UserDAO;
+import org.grouplens.lenskit.data.dao.UserEventDAO;
+import org.grouplens.lenskit.data.event.Event;
+import org.grouplens.lenskit.data.event.Rating;
+import org.grouplens.lenskit.data.history.UserHistory;
+import org.grouplens.lenskit.data.pref.Preference;
+import org.grouplens.lenskit.eval.AbstractTask;
+import org.grouplens.lenskit.eval.TaskExecutionException;
+import org.grouplens.lenskit.eval.data.CSVDataSourceBuilder;
+import org.grouplens.lenskit.eval.data.DataSource;
+import org.grouplens.lenskit.eval.data.crossfold.*;
+import org.grouplens.lenskit.eval.data.traintest.GenericTTDataBuilder;
+import org.grouplens.lenskit.eval.data.traintest.TTDataSet;
+import org.grouplens.lenskit.util.io.Describer;
+import org.grouplens.lenskit.util.io.Descriptions;
+import org.grouplens.lenskit.util.io.HashDescriptionWriter;
+import org.grouplens.lenskit.util.io.UpToDateChecker;
+import org.grouplens.lenskit.util.table.writer.CSVWriter;
+import org.grouplens.lenskit.util.table.writer.TableWriter;
+import org.slf4j.Logger;
+import org.slf4j.LoggerFactory;
+
+import java.io.File;
+import java.io.IOException;
+import java.util.*;
+
+/**
+ * A task that takes a crossfold and downsamples each user in each training set.
+ * For now this only supports downsampling to a constant size, in the future, 
+ * it should support arbitrary downsamplers.
+ */
+public class DownsampleDataSet extends AbstractTask<List<TTDataSet>> {
+    private static final Logger logger = LoggerFactory.getLogger(DownsampleDataSet.class);
+
+    List<TTDataSet> sources;
+    private int retain = 0;
[…]

### 2014-04-16 lenskit/src/main/java/coldstartrecommendation/ItemScorerCoveragePredictMetric.java

@@ -1,0 +1,106 @@
+package coldstartrecommendation;
+
+import org.grouplens.lenskit.ItemScorer;
+import org.grouplens.lenskit.Recommender;
+import org.grouplens.lenskit.eval.Attributed;
+import org.grouplens.lenskit.eval.data.traintest.TTDataSet;
+import org.grouplens.lenskit.eval.metrics.AbstractMetric;
+import org.grouplens.lenskit.eval.metrics.ResultColumn;
+import org.grouplens.lenskit.eval.traintest.TestUser;
+import org.grouplens.lenskit.vectors.SparseVector;
+import org.grouplens.lenskit.vectors.VectorEntry;
+import org.slf4j.Logger;
+import org.slf4j.LoggerFactory;
+
+/**
+ * Simple evaluator that records user, rating and prediction counts and computes
+ * recommender coverage over the queried items.
+ *
+ * @author <a href="http://www.grouplens.org">GroupLens Research</a>
+ */
+public class ItemScorerCoveragePredictMetric extends AbstractMetric<ItemScorerCoveragePredictMetric.Context, ItemScorerCoveragePredictMetric.AggregateCoverage, ItemScorerCoveragePredictMetric.Coverage> {
+    private static final Logger logger = LoggerFactory.getLogger(ItemScorerCoveragePredictMetric.class);
+
+    public ItemScorerCoveragePredictMetric() {
+        super(AggregateCoverage.class, Coverage.class);
+    }
+
+    @Override
+    public Context createContext(Attributed algo, TTDataSet ds, Recommender rec) {
+        return new Context(rec.getItemScorer());
+    }
+
+    @Override
+    public Coverage doMeasureUser(TestUser user, Context context) {
+        SparseVector ratings = user.getTestRatings();
+        SparseVector predictions = context.is.score(user.getUserId(), user.getTestHistory().itemSet());
+        
+        if (predictions == null) {
+            return null;
+        }
+        int n = 0;
+        int good = 0;
+        for (VectorEntry e : ratings.fast()) {
+            n += 1;
+            if (predictions.containsKey(e.getKey())) {
+                good += 1;
+            }
+        }
+        context.addUser(n, good);
+        return new Coverage(n, good);
+    }
+
+    @Override
+    protected AggregateCoverage getTypedResults(Context context) {
+        return new AggregateCoverage(context.nusers, context.npreds, context.ngood);
+    }
+
+    public static class Coverage {
[…]

### 2014-04-16 lenskit/src/main/java/coldstartrecommendation/OracleItemScorer.java

@@ -1,13 +1,13 @@
 package coldstartrecommendation;
 
-import it.unimi.dsi.fastutil.longs.LongArraySet;
-import it.unimi.dsi.fastutil.longs.LongOpenHashSet;
-import it.unimi.dsi.fastutil.longs.LongSet;
-import it.unimi.dsi.fastutil.longs.LongSortedSet;
+import it.unimi.dsi.fastutil.longs.*;
+import org.apache.commons.lang3.tuple.ImmutablePair;
+import org.apache.commons.lang3.tuple.Pair;
 import org.grouplens.lenskit.RatingPredictor;
 import org.grouplens.lenskit.baseline.MeanDamping;
 import org.grouplens.lenskit.basic.AbstractItemScorer;
 import org.grouplens.lenskit.basic.AbstractRatingPredictor;
+import org.grouplens.lenskit.collections.CollectionUtils;
 import org.grouplens.lenskit.data.dao.ItemEventDAO;
 import org.grouplens.lenskit.data.dao.UserDAO;
 import org.grouplens.lenskit.data.dao.UserEventDAO;
@@ -15,12 +15,16 @@ import org.grouplens.lenskit.data.event.Rating;
 import org.grouplens.lenskit.data.history.UserHistory;
 import org.grouplens.lenskit.eval.data.DataSource;
 import org.grouplens.lenskit.eval.data.traintest.TTDataSet;
+import org.grouplens.lenskit.eval.script.BuiltBy;
 import org.grouplens.lenskit.vectors.MutableSparseVector;
 import org.grouplens.lenskit.vectors.VectorEntry;
 
 import javax.annotation.Nonnull;
 import javax.inject.Inject;
+import java.util.Collection;
+import java.util.HashSet;
 import java.util.List;
+import java.util.Map;
 
 /**
  * TODO: document me
@@ -29,30 +33,47 @@ public class OracleItemScorer extends AbstractItemScorer {
 
     private final DataSource data;
     private final UserEventDAO ud;
+    Long2ObjectOpenHashMap<HashSet<Pair<Long, Double>>> userSets;
+    
     @Inject
     OracleItemScorer(DataSource data, UserEventDAO ud) {
         this.data = data;
         this.ud = ud;
+
+        userSets = new Long2ObjectOpenHashMap<HashSet<Pair<Long, Double>>>();
+        
+        for (Rating r : data.getEventDAO().streamEvents(Rating.class).fast()) {
+            long user = r.getUserId();
[…]

### 2014-04-16 lenskit/src/main/java/coldstartrecommendation/TopNRMSEMetric.java

@@ -1,0 +1,161 @@
+package coldstartrecommendation;
+
+import org.grouplens.lenskit.Recommender;
+import org.grouplens.lenskit.collections.CollectionUtils;
+import org.grouplens.lenskit.eval.Attributed;
+import org.grouplens.lenskit.eval.data.traintest.TTDataSet;
+import org.grouplens.lenskit.eval.metrics.AbstractMetric;
+import org.grouplens.lenskit.eval.metrics.ResultColumn;
+import org.grouplens.lenskit.eval.metrics.topn.ItemSelector;
+import org.grouplens.lenskit.eval.metrics.topn.TopNMetricBuilder;
+import org.grouplens.lenskit.eval.traintest.TestUser;
+import org.grouplens.lenskit.scored.ScoredId;
+import org.grouplens.lenskit.util.statistics.MeanAccumulator;
+import org.grouplens.lenskit.vectors.SparseVector;
+import org.grouplens.lenskit.vectors.VectorEntry;
+
+import java.util.List;
+
+import static java.lang.Math.sqrt;
+
+/**
+ * computes RMSE for just those items returned in a topN recommendation task.
+ */
+public class TopNRMSEMetric extends AbstractMetric<TopNRMSEMetric.Context, TopNRMSEMetric.AggregateResult, TopNRMSEMetric.UserResult> {
+    private final String suffix;
+    private final int listSize;
+    private final ItemSelector candidates;
+    private final ItemSelector exclude;
+
+    public TopNRMSEMetric(String sfx, int listSize, ItemSelector candidates, ItemSelector exclude) {
+        super(AggregateResult.class, UserResult.class);
+        suffix = sfx;
+        this.listSize = listSize;
+        this.candidates = candidates;
+        this.exclude = exclude;
+    }
+
+    @Override
+    protected String getSuffix() {
+        return suffix;
+    }
+
+    @Override
+    public Context createContext(Attributed algo, TTDataSet ds, Recommender rec) {
+        return new Context();
+    }
+
+    @Override
+    public UserResult doMeasureUser(TestUser user, Context context) {
+        List<ScoredId> recs = user.getRecommendations(listSize, candidates, exclude);
+        SparseVector ratings = user.getTestRatings();
+        SparseVector predictions = user.getPredictions();
+        if (predictions == null || recs == null) {
+            return null;
+        }
+        
+        double sse = 0;
+        int n = 0;
+        for(ScoredId s : CollectionUtils.fast(recs)) {
[…]

### 2014-04-17 lenskit/src/eval/chart.R
small tweaks, and removeing some cruft

@@ -6,7 +6,8 @@ library("ggplot2")
 all.data <- read.csv("eval-user.csv")
 
 serr = function(d) {qnorm(0.975)*sd(d)/sqrt(length(d))}
-nonMetrics = c("Algorithm","Retain", "DataSet")
+nonMetrics = c("Algorithm","Retain", "DataSet", "User", "Partition", "NGood", "TestEvents", "TrainEvents", "NAttempted", "RatingEntropy")
+
 metrics = setdiff(names(all.data),nonMetrics)
 
 for (metric in metrics) {

### 2014-04-17 lenskit/src/main/java/coldstartrecommendation/TopNMapMetric.java

@@ -1,0 +1,157 @@
+package coldstartrecommendation;
+
+import com.google.common.base.Preconditions;
+import it.unimi.dsi.fastutil.longs.LongSet;
+import org.grouplens.lenskit.Recommender;
+import org.grouplens.lenskit.collections.CollectionUtils;
+import org.grouplens.lenskit.core.LenskitRecommender;
+import org.grouplens.lenskit.eval.Attributed;
+import org.grouplens.lenskit.eval.data.traintest.TTDataSet;
+import org.grouplens.lenskit.eval.metrics.AbstractMetric;
+import org.grouplens.lenskit.eval.metrics.ResultColumn;
+import org.grouplens.lenskit.eval.metrics.topn.ItemSelector;
+import org.grouplens.lenskit.eval.metrics.topn.ItemSelectors;
+import org.grouplens.lenskit.eval.metrics.topn.TopNMetricBuilder;
+import org.grouplens.lenskit.eval.traintest.TestUser;
+import org.grouplens.lenskit.knn.item.ItemSimilarity;
+import org.grouplens.lenskit.knn.item.model.ItemItemBuildContext;
+import org.grouplens.lenskit.scored.ScoredId;
+import org.grouplens.lenskit.util.statistics.MeanAccumulator;
+import org.grouplens.lenskit.vectors.SparseVector;
+
+import javax.annotation.Nullable;
+import java.util.List;
+
+/**
+ * Metric that measures how diverse the items in the TopN list are.
+ * 
+ * To use this metric ensure that you have a reasonable itemSimilarityMetric configured in each algorithm
+ * 
+ * Example configuration (add this to your existing algorithm configuration)
+ * <pre>
+ * root (ItemSimilarityMetric)
+ * within (ItemSimilarityMetric) {
+ *     bind VectorSimilarity to CosineVectorSimilarity
+ *     bind UserVectorNormalizer to BaselineSubtractingUserVectorNormalizer
+ *     within (UserVectorNormalizer) {
+ *         bind (BaselineScorer, ItemScorer) to ItemMeanRatingItemScorer
+ *         set MeanDamping to 5.0d
+ *     }
+ * }
+ * </pre>
+ * 
+ * I also recommend enabling model sharing and cacheing between algorithms to make this much more efficient.
+ * 
+ * This computes the average disimilarity (-1 * similarity) of all pairs of items. 
+ * 
[…]

### 2014-04-17 lenskit/src/main/java/coldstartrecommendation/TopNRMSEMetric.java

@@ -133,26 +133,6 @@ public class TopNRMSEMetric extends AbstractMetric<TopNRMSEMetric.Context, TopNR
      * @author <a href="http://www.grouplens.org">GroupLens Research</a>
      */
     public static class Builder extends TopNMetricBuilder<Builder, TopNRMSEMetric> {
-        private String suffix;
-
-        /**
-         * Get the column suffix for this metric.
-         * @return The column suffix.
-         */
-        public String getSuffix() {
-            return suffix;
-        }
-
-        /**
-         * Set the column suffix for this metric.
-         * @param l The column suffix
-         * @return The builder (for chaining).
-         */
-        public Builder setSuffix(String l) {
-            suffix = l;
-            return this;
-        }
-
         @Override
         public TopNRMSEMetric build() {
             return new TopNRMSEMetric(suffix, listSize, candidates, exclude);

### 2014-04-21 paper/main.tex
set up to start writing

@@
-OUTLINE
+\documentclass[letterpaper]{sig-alternate}
 
-1. Introduction
-  [story]
-    New user experiance is important
-      new users are needed to maintine usage of a site as older users leave the system
-      all users are new users at some time
-      new user experiance may set the expectation for how the system behaves in the future. (Citation of this may be important)
+\pdfpagewidth=8.5in
+\pdfpageheight=11in
 
-2. Background
-3. Methodology
-4. Experiments %Not a fan of this title.
-  4.1 RMSE by profile size
-  4.2 Soft metrics
-  4.3 Value of implicit information
-5. conclusions
+%\usepackage[factor=250,spacing=true]{microtype}
+
+\begin{document}
+\conferenceinfo{RecSys '14}{October 6--10, 2014, Foster City, Silicon Valley, USA}
+%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
+%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
+
+\title{TODO: title}
+
+\numberofauthors{2}
+
+\author {
+\alignauthor
+Daniel Kluver\\
+\affaddr{GroupLens Research}\\
+\affaddr{Department of Computer Science and Engineering}\\
+\affaddr{University of Minnesota}\\
+\affaddr{Minneapolis, MN 55455 USA}\\
+\email{kluver@cs.umn.edu}
+\alignauthor
+Joseph A. Konstan\\
+\affaddr{GroupLens Research}\\
+\affaddr{Department of Computer Science and Engineering}\\
+\affaddr{University of Minnesota}\\
+\affaddr{Minneapolis, MN 55455 USA}\\
+\email{konstan@cs.umn.edu}
+}
+
+\maketitle
+\begin{abstract}
+
+TODO: abstract
+
+\end{abstract}
+
+%TODO: update these things.
+% A category with the (minimum) three required fields
+\category{H.4}{Information Systems Applications}{Miscellaneous}
+%A category including the fourth, optional field follows...
+\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]
+
+%TODO: update these things.
+\terms{Theory}
+
+%TODO: update these things.
+\keywords{ACM proceedings, \LaTeX, text tagging}
+
+\section{Introduction}
+\section{Methodology}
+\section{Results}
+\section{Conclusions}
+
+\section{Acknowledgements}
+ \cite{harper2005economic}
+ TODO:
+
+\bibliographystyle{abbrv}
+\bibliography{resources}  % sigproc.bib is the name of the Bibliography in this case
+
+\end{document}

### 2014-04-21 lenskit/src/main/java/coldstartrecommendation/TopNRMSEMetric.java

@@ -54,6 +54,7 @@ public class TopNRMSEMetric extends AbstractMetric<TopNRMSEMetric.Context, TopNR
             return null;
         }
         
+        double ratSum = 0;
         double sse = 0;
         int n = 0;
         for(ScoredId s : CollectionUtils.fast(recs)) {
@@ -62,13 +63,15 @@ public class TopNRMSEMetric extends AbstractMetric<TopNRMSEMetric.Context, TopNR
             if (Double.isNaN(rating) || Double.isNaN(prediction)) {
                 continue;
             }
+            ratSum += rating;
             double err = prediction - rating;
             sse += err * err;
             n++;
         }
         if (n > 0) {
             double rmse = sqrt(sse / n);
-            return context.addUser(n, rmse);
+            double avgRat = ratSum /n;
+            return context.addUser(n, rmse, avgRat);
         } else {
             return null;
         }
@@ -85,23 +88,20 @@ public class TopNRMSEMetric extends AbstractMetric<TopNRMSEMetric.Context, TopNR
 
         @ResultColumn("TopN.RMSE")
         public final double score;
-
-        public UserResult(double len, double rmse) {
+        
+        @ResultColumn("TopN.Ave.Rat")
+        public final double averageRating;
+        
+        public UserResult(double len, double rmse, double avgRat) {
             count = len;
             score = rmse;
+            averageRating = avgRat;
         }
     }
     
-    public static class AggregateResult {
-        @ResultColumn("TopN.RMSE.seenItems")
-        public final double count;
-
-        @ResultColumn("TopN.RMSE")
-        public final double score;
-
-        public AggregateResult(double len, double rmse) {
-            count = len;
-            score = rmse;
+    public static class AggregateResult extends UserResult {
+        public AggregateResult(double len, double rmse, double avgRat) {
+            super(len, rmse, avgRat);
         }
     }
     
@@ -109,19 +109,21 @@ public class TopNRMSEMetric extends AbstractMetric<TopNRMSEMetric.Context, TopNR
         double nUsers = 0;
         double nItems = 0;
         double rmseSum = 0;
+        double avgRatSum = 0;
         private AggregateResult result;
 
-        public UserResult addUser(int n, double rmse) {
+        public UserResult addUser(int n, double rmse, double avgRat) {
             nUsers += 1;
             nItems += n;
             rmseSum += rmse;
-            return new UserResult(n, rmse);
+            avgRatSum += avgRat;
[…]

### 2014-04-21 paper/main.tex
some work

@@
 \section{Results}
+
+\begin{figure}
+\centering
+\includegraphics[width=\columnwidth]{../lenskit/target/analysis/RMSE.pdf}
+\caption{caption}
+\label{fig:rmse}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\columnwidth]{../lenskit/target/analysis/nDCG.pdf}
+\caption{caption}
+\label{fig:ndcg}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\columnwidth]{../lenskit/target/analysis/MAP.pdf}
+\caption{caption}
+\label{fig:map}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\columnwidth]{../lenskit/target/analysis/Precision.pdf}
+\caption{caption}
+\label{fig:precision}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\columnwidth]{../lenskit/target/analysis/Precision_fallout.pdf}
+\caption{caption}
+\label{fig:fallout}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\columnwidth]{../lenskit/target/analysis/TopN_RMSE_seenItems.pdf}
+\caption{caption}
+\label{fig:topn.rmse.seen}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\columnwidth]{../lenskit/target/analysis/TopN_RMSE.pdf}
+\caption{caption}
+\label{fig:topn.rmse}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\columnwidth]{../lenskit/target/analysis/TopN_Ave_Rat.pdf}
+\caption{caption}
+\label{fig:topn.avg.rat}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\columnwidth]{../lenskit/target/analysis/TopN_MeanPopularity.pdf}
+\caption{caption}
+\label{fig:pop}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\columnwidth]{../lenskit/target/analysis/diversity.pdf}
+\caption{caption}
+\label{fig:diversity}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\columnwidth]{../lenskit/target/analysis/topN_entropy.pdf}
+\caption{caption}
+\label{fig:spread}
+\end{figure}
+
 \section{Conclusions}

### 2014-04-21 lenskit/src/eval/chart.R

@@ -1,27 +1,26 @@
 # Create a chart comparing the algorithms
 
-cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
 library("ggplot2")
 
 all.data <- read.csv("eval-user.csv")
 
-serr = function(d) {qnorm(0.975)*sd(d)/sqrt(length(d))}
 nonMetrics = c("Algorithm","Retain", "DataSet", "User", "Partition", "NGood", "TestEvents", "TrainEvents", "NAttempted", "RatingEntropy")
 
 metrics = setdiff(names(all.data),nonMetrics)
 
 for (metric in metrics) {
-  plot = ggplot(all.data, aes_string(x="Retain", y=metric, color="Algorithm"))+stat_summary(fun.y=mean, geom="line") + stat_summary(fun.data = mean_cl_normal, geom="errorbar")
+  plot = ggplot(all.data, aes_string(x="Retain", y=metric, color="Algorithm", shape="Algorithm"))+stat_summary(fun.y=mean, geom="line")+stat_summary(fun.y=mean, geom="point") + stat_summary(fun.data = mean_cl_normal, geom="errorbar")
+  metric = gsub('\\.', '_', metric)
   filename = paste(metric,".pdf",sep="")
   cat("Outputting to",filename,"\n")
-  pdf(filename, width=11, height=8.5)
+  pdf(filename, width=5.5, height=4.25)
   print(plot)
   dev.off()
 }
 
 
 dat <- read.csv("eval-results.csv")
-plot = ggplot(dat, aes(x=Retain, y=TopN.Entropy, color=Algorithm))+geom_point()+stat_summary(fun.y=mean, geom="line")
-pdf("topN.entropy.pdf", width=11, height=8.5)
+plot = ggplot(dat, aes(x=Retain, y=TopN.Entropy, color=Algorithm, shape=Algorithm))+geom_point()+stat_summary(fun.y=mean, geom="line")
+pdf("topN_entropy.pdf", width=5.5, height=4.5)
 print(plot)
 dev.off()

### 2014-04-22 paper/main.tex
work on results section

--- 2014-04-21 paper/main.tex
@@
 \section{Introduction}
+  % describe new user experience in recommender systems.
+  When someone first interacts with any intelligent system, the system knows nothing about them.
+TODO:
+
+  % new user experience is important, and here is why.
+    % talk about first impressions and user trust
+  The new user experience is a very important part of any recommender system.
+  The first interaction with a system is the only chance the system gets to make a first impression on the user.
+  This first impression can be very important in structuring a user's expectations for how the system will behave, and determining how much the user trusts the system's recommendations.
+  If the system behaves poorly when the user is new, then the user may leave and not come back.
+  This is especially a concern with recommender systems which rely on large amounts of user information to make personalization decisions.
+  When users are new the system will not have much information to personalize the user experience which may lead the user to leave the system.
+
+  % Past work on cold start has focused on… (how long should this section be?)
+    % ways to sort items to get to most value out of early ratings
+    % research looking at user experience and other forms of data collection
+
+  Past work has referred to the new user experience issue as the \emph{user cold start} problem.
+  This work has normally focused on two
+TODO:
+
+  % How an algorithm behaves the first time a user sees it is important (first impression)
+    % Past work looks carefully at issues X, Y, and Z, but doesn’t take a systematic look at the behavior of algorithms for users with few ratings.
+  % Our work sits at the intersection of two important fields of recommender system research: New user experiance, and comparing algorithms.
+While there has been much work comparing various algorithms, there has been little formal work comparing algorithms based on their performance for new users.
+Therefore our work focuses on the behavior of recommender algorithms across different amounts of user data.
+(subsection might need re-ordering)
+  % RQs
+    % RQ1 How do the behavior of different algorithms change with the number of ratings?
+      % Explain that there are a lot of different ways to measure algorithm behavior.
+        % Predictive accuracy, commonly measured with MAE or RMSE
+          % coverage
+        % Value of recommendations, tradidionally done with metrics from information retrieval such as precision, recall, MAP, etc. These algorithms try to measure ...
+          % cite the John/Joe paper (RMSE is not enough)
+          % cite the paper cited by 10 is enough
+          % topN RMSE
+        % There are many properties of recommenders that are (find words here)
+          % popularity
+          % diversity
+          % spread
+      % RQ1A  How does algorithm behavior change with respect to predictive accuracy?
+      % RQ1B  How does algorithm behavior change with respect to TopN recommender performance?
+      % RQ1C How does algorithm behavior change with respect to other metrics such as popularity and diversity?
+
+    % We look at three families of recommender algorithms (user based KNN, item based KNN, SVD)
+    % we compare on several dimensions
+    % we suggest a methodology for performing these types of computations
+
 \section{Methodology}
+  % (high level description/restatement of approach) To answer our research questions we will various metrics of recommenders in an offline setting simulations users as they join the system.
+    % maybe hit some limitations
+    % motiviate _why do a simulation_ (a lot of metrics we want to try)
+  % temporal replay evaluation 
+    % similar technique, harder to do than ours, and with a different goal.
+  % SOMEWHERE AROUND HERE talk about drnner and adaptive bootstrapping and 10 isenough paper's ad hoc plots by datapoint size.
+  % algorithms (discuss tuning of algorithms per algorithm?)
+    % item mean baseline
+    % user adjusted baseline (@michael, is there a cite for this?)
+    % user-user
+    % item-item
+    % SVD
+  % metrics (expected / good behavior for each metric (I.E. this should go down, any reasonable algorithm should always beet baseline)) 
+    % accuracy metrics 
+      % RMSE
+    % recommendation performance metrics (list subject to change)
+      % ndcg
+      % MAP@N (not yet measured)
+      % precision@N
+      % fallout @ N
+      % RMSE of recommended items
+      % (note that we looked at MRR and recall, but found them to behave the same).
+    % other properties
+      % recommendation average popularity
+      % recommendation diversity
+      % spread (entropy of recommendations) (related to recommendation coverage)
+
+  % dataset
+    % # user
+    % # items
+    % distributions of # ratings per user (important: min 20)
+  % subsampling strategy
+    % describe normal crossfold solution
+    % describe issues with solution
+    % describe our crossfold then subsample solution.
+
 \section{Results}
+
+  Using the methodology described in \cite{sec:methodology} we compute a value for each metric on each of the five algorithms.
+  We plot these results against the number of items retained.
+  These plots allow us to compare the algorithms against each other, and also, to understand how each algorithm's performance changes with more data.
+  Each plot will have error bars showing the 95\% confidence intervals around the average per use value for the metric.
+
+\subsection{Prediction accuracy}
+  % Prediction accuracy
+    % RMSE plot
+
+TODO: rmse plot
+
+    % point out error bars etc.
+    % explain / interpret plot
+  Figure \ref{fig:rmse} shows how accurate our algorithms are as measured by RMSE.
+  As would be expected the accuracy of our item baseline doesn't change as the amount of information changes.
+  The User-User and svd algorithms both perform as expected, starting out with the same accuracy as the item baseline, and then monotonically improving.
+  User-User does show a small increase of error with only one rating, but the change is quite small and unlikely to be noticed by users.
+  Interestingly, User-User and svd both perform about as well as our smarter user item baseline. %TODO: explain this better
+
+      % explain strangeness of Item-Item
+  Item-Item performs quite poorly in our cold start setting.
+  Unlike the other algorithms Item-Item actually performs worse as it gets more ratings for the first few ratings.
+  The explanation for this has to do with the coverage of the algorithm.
+  Item-Item, unlike the other two algorithms, has relatively low coverage at cold start, needing around 8 ratings to reach 90 percent coverage. %TODO: check this
+
+      % possibly coverage plot
+      % true crossing point?
+  % Recommendation quality metrics
+    % NDCG
+    % MAP, precision, recall plots 
+    % explain results from MAP and precision
+    % explain recall results
+    % recall results, strangeness and popularity
+    % popularity graph
+    % TopN average rating / topn rmse
+    % interpret graph and reason about what it means about quality
+  % Other recommender metrics
+    % diversity plot
+    % talk about popularity and diversity
+      % algos don't cross, and seem more straight, this seems to be more of a system wide choice than other metrics.***
+      % relatively the same thing as each other.
+      % upward diversity trend in baselines - seen items removed from recs => must increase diversity
+      % we don't know best, but user user is probably too bad, and avg rating is probably too high SVD and Item-Item probably in a good zone.
+      % user study would be needed to find out.
+    % spread plot
+      % by and far item-item has the best spread
+      % user-user has good spread, but interestingly the spread gets worse
+        % not sure why, perhaps as more ratings there is a regression to the mean sort of trend where more ratings means mroe neighbors available, means more ratings.
+        % either way we don't see this as a good thing...
+      % svd
+        % spread seems a little low. like before we don't _know_, but it looks like svd might focus too much on a smaller list of items.
 \section{Conclusions}
+  % summary of results in a big old table.
+    % the accuracy of svd and user-user both seemed good. item-item seemed to have trouble especially with very new users.
+    % svd did quite well on our traditional topN metrics, item-item did OK, and user-user did bad., but there are flaws with these metrics.
+    % on our topN RMSE item-item (again) seemed to have issues, and user-user did quite well, but the low hit rate is troubleing. 
+    % again svd seems the winner here.
+    % on our fluffy metrics item-item and svd seemed to do well. I might argue that item-item seemed to do better (spread, and higher throughout).
+    % user-user seemed to do poor
+  % what this means
+    % user-user seems to be able to give accurate predictions, but seems to focus on overly obscure items in recommendation.
+    % item-item seems to give not-bad recommendations, but its predictions have issues
+    % based only on these results svd seems to be the best performer and is our recommendation to new systems.
+  % other takeaways
+    % regularization
+    % if you use a switching recommender, this methodology can help you choose when to switch.
+    % non-traditional metrics can tell you more about what is going on, these seem to be more properties of the class of algo than of the data, so its worth looking at these when choosing algorithms, to find one with properties that are good for your task.
+
+  % limitations and opprotunities
+    % dataset limitations
+      % only one domain
+      % no quitters in our dataset
+    % metrics limitations
+      % many metrics require user data to help us understand what users want from a recommender
+      % no well known ``goodness'' metric for topN.
+    % algorithm limitations
+      % only covered three common families,
+      % different algos tuned specifically for cold start.
+      % algos that use extra data about users or about items.
+
+  % despite the limitations we were able to show differences between algorithms in a cold start situation.
 

### 2014-04-22 paper/main.tex
more results work

@@
       % explain strangeness of Item-Item
-  Item-Item performs quite poorly in our cold start setting.
-  Unlike the other algorithms Item-Item actually performs worse as it gets more ratings for the first few ratings.
-  The explanation for this has to do with the coverage of the algorithm.
-  Item-Item, unlike the other two algorithms, has relatively low coverage at cold start, needing around 8 ratings to reach 90 percent coverage. %TODO: check this
+  Item-Item performs quite poorly in the cold start setting.
+  Until we have around 20 ratings the RMSE of Item-Item is worse than the item baseline.
+  More interestingly, unlike the other algorithms Item-Item actually appears to performs worse as it gets more ratings for the first few ratings.
 
+  The curve exhibited by Item-Item's RMSE is caused by two trends.
+  First as the system revives more ratings, predictions from Item-Item are more accurate, leading to a downward trend.
+  Secondly, as the system revives more ratings, Item-Item can make predictions for more items, meaning the fewer predictions are serviced by the baseline.
+  Because the baseline is more accurate, this leads to an upward trend.
+  Overall this leads to curved performance.
+  If we were to compute RMSE only over the items where item-item can make a prediction we would see a curve similar to the other algorithms, but shifted up.
       % possibly coverage plot
       % true crossing point?
+
+\subsection {Recommendatin quality metrics}
   % Recommendation quality metrics
     % NDCG
-    % MAP, precision, recall plots 
+  Figure \ref{fig:ndcg} shows the algorithm's performance on the nDCG metric.
+  The algorithm's behavior is essentially equivalent to the RMSE behavior.
+  Again, User-User has a minor dip on very new users, but then tracks SVD in uniformly improving quite well from there.
+  Likewise Item-Item exhibits the same dipped accuracy, although it doesn't catch up the baseline as fast.
+  Unlike previous metrics, the item baseline, and the user-item baseline have the same performance.
+  This is because nDCG is a ranking metric, not an accuracy metric.
+  Because the user item baseline orders items the same way as the item baseline, both will have the same accuracy on any ranking metrics.
+
+    % MAP, precision, fallout plots 
     % explain results from MAP and precision
@@
     % recall results, strangeness and popularity
-    % popularity graph
+
+  Figure \ref{fig:map} shows MAP@20, precision@20, and fallout@20.
+  All three metrics show essentially the same trend, User-User gets the lowest score, the baselines get the highest score.
+  On both MAP@20 and precision@20 svd gets a higher score than Item-Item throughout, performing almost as highly as the baseline.
+  On fallout@20 Item-Item performs higher than svd and the baseline, but decreases faster than bath, ending with a lower score than either. 
+  We also tried recall@20 and mrr@20 and found the same trend.
+
+  This result is rather strange as MAP and precision are metrics where high scores are good, where fallout is a metric where high scores are bad.
+  This leads to an issue, these metrics contradict each other in terms of the quality of the recommendations from our algorithm.
+  One possible reason for this is that these topN metrics are known to be biased by popular items (TODO: cite bellogin).
+  Looking at the average popularity (figure \ref{fig:pop})of recommended items, we see that this seems to explain most of the variance in these metrics.
+  Therefore we seek a different metric to analyze the quality of the recommendations.
+
+  Figure \ref{fig:topN.rmse}
     % TopN average rating / topn rmse
@@
   % Other recommender metrics
+    % popularity graph
     % diversity plot

### 2014-04-23 paper/main.tex
more work on results.

@@
 
-  This result is rather strange as MAP and precision are metrics where high scores are good, where fallout is a metric where high scores are bad.
+  This result is rather strange as MAP and precision are metrics where high scores imply better performance, whereas fallout is a metric where high scores imply poor performance.
   This leads to an issue, these metrics contradict each other in terms of the quality of the recommendations from our algorithm.
-  One possible reason for this is that these topN metrics are known to be biased by popular items (TODO: cite bellogin).
-  Looking at the average popularity (figure \ref{fig:pop})of recommended items, we see that this seems to explain most of the variance in these metrics.
-  Therefore we seek a different metric to analyze the quality of the recommendations.
+  One possible reason for this is that these topN metrics are known to be biased toward recommendations that have more popular items \cite{bellogin}.
+  Looking at the average popularity (figure \ref{fig:pop}) of recommended items, we see that the average popularity of the recommendations closely replicates the previous three plots.
+  Therefore we seek a different, less biased metric to analyze the quality of the recommendations.
 
-  Figure \ref{fig:topN.rmse}
   % TopN average rating / topn rmse
     % interpret graph and reason about what it means about quality
+  The first graph in figure \ref{fig:topN.rmse} shows the average number of items in the top 20 recommendations for each algorithm that were also in the test set.
+  The average number of test set items in recommendations shows a similar trend to the MAP, precision, and recall metrics, user-user getting by far the fewest seen items, and the baseline getting the most.
+  %TODO: talk about trust building here!
+
+  The second graph in figure \ref{fig:topN.rmse} shows the average rating for those items in the top 20 recommendations for each algorithm.
+  The first thing to note about this graph is that all of these algorithms are doing a reasonable job of recommending items the user might like to watch, with all algorithms averaging above 4 stars (implying that the recommended items are, at least on average, enjoyable).
+  SVD performs as expected, starting at close to the same average rating as the baselines, and then slowing improving as more ratings are entered.
+  Item-Item performs poorly, not only performing worse than the baseline, but showing worse performance as more ratings are added.
+  This is quite possibly a similar trend to the one we observed with RMSE and nDCG, where Item-Item's performance decreased as coverage increased leading to the curved performance at cold start.
+
+  During testing with other values of N for this metric and the next, we found that User-User's behavior compared to the other algorithms changed significantly with different values of N.
+  This is probably due to the very low rate of having seen items in the recommendations, which makes its value much less stable, than the other algorithms.
+  Because of this behavior we will draw no conclusion from User-User's performance on this metric, to avoid accidentally attributing behavior to it, that was instead formed by our choice of 20 as N.
+
+  The third graph in figure \ref{fig:topN.rmse} shows the RMSE our algorithms got on items in the top 20 recommendations for each algorithm.
+  Not surprisingly, this figure shows the same trends as the last one meaning that the algorithms whose recommendations were (on average) most likable, were those whose recommendations were most accurate.
+  Again, SVD performs quite well, Item-Item shows poor performance, and, due to sensitivity of User-User to our choice of N, we make no conclusions about User-User.
+
+\subsection{Other recommender metrics}
   % Other recommender metrics
+  Figure \ref{fig:pop} shows the
+
     % popularity graph
@@
 \section{Acknowledgements}
- \cite{harper2005economic}
  TODO:

### 2014-04-24 paper/main.tex
finished extreemly rough draft of results section

@@
   % Other recommender metrics
-  Figure \ref{fig:pop} shows the
+    % popularity graph
+  The first graph in figure \ref{fig:pop} shows the average popularity of recommended items by each algorithm.
+  %TODO: re-write this
+  Across all datapoint sizes the ranking of algorithms is the same, with the baselines recommending the most popular movies, followed by svd, which recommends less popular movies, then Item-Item, which recommends much less popular movies, leaving User-User to recommend the least popular movies.
+  Unfortunately, we don't know of any guideline for understanding how popular users desire their recommendations to be.
+  That in mind, we expect that users will feel the baseline's popularity (around 1000) is too high, and User-User's popularity (around 50) is too low.
+  SVD and Item-Item both seem to perform well, with SVD generating more popular recommendations, than Item-Item, and both algorithms showing a significant trend to make less popular predictions as we know more about the user.
 
-    % popularity graph
     % diversity plot
-    % talk about popularity and diversity
-      % algos don't cross, and seem more straight, this seems to be more of a system wide choice than other metrics.***
-      % relatively the same thing as each other.
-      % upward diversity trend in baselines - seen items removed from recs => must increase diversity
-      % we don't know best, but user user is probably too bad, and avg rating is probably too high SVD and Item-Item probably in a good zone.
-      % user study would be needed to find out.
+  The diversity graph in \ref{fig:pop} shows a similar trend to the popularity trend with the baseline algorithms producing the least diverse list of recommendations and User-User providing by far the most diverse list of recommendations.
+  For the first few ratings SVD is able to provide more diverse recommendations than Item-Item.
+  However, Item-Item seems to grow in diversity faster than SVD in response to more data, so after 4 ratings Item-Item is providing slightly more diverse recommendations than SVD.
+  Overall, Item-Item and SVD are providing recommendations that, at least in an offline context, appear to have quite comparable diversity.
+
   % spread plot
-      % by and far item-item has the best spread
-      % user-user has good spread, but interestingly the spread gets worse
-        % not sure why, perhaps as more ratings there is a regression to the mean sort of trend where more ratings means mroe neighbors available, means more ratings.
-        % either way we don't see this as a good thing...
-      % svd
-        % spread seems a little low. like before we don't _know_, but it looks like svd might focus too much on a smaller list of items.
+  The graph in \ref{fig:spread} shows the spread of the recommendations from each algorithm.
+  Because this metric can only be computed over a group of people, confidence intervals over the per-user average are not provided, instead the specific spread values computed in each five folds are shown as points, with lines connecting the value from the five metrics.
+  %TODO: make this point earier and better.
+  Like the diversity and popularity metrics we see that the values take a much larger range, with different algorithms seeming to converge to different amounts spread.
+  The baseline algorithms, which recommend every user the same list of recommendations (unless the user has rated one of those items) have the lowest spread.
+  SVD has the second lowest spread, starting with a spread very close to the baselines and then slowly growing more spread as it gets more ratings.
+  Like popularity, we don't know exactly how much spread is best for a recommender system, but we can guess that we want more spread from our recommendations than baseline, and all other things being equal, more spread is probably better.
+  Based on this intuition it seems that SVD might have focus on too few items in its recommendations.
+
+  User-User and Item-Item both have significantly more spread than the other algorithms.
+  They show interestingly opposite trends, with User-User providing more spread at one rating, and then decreasing, and Item-Item giving less spread and then increasing.
+  By 8 ratings, we see that Item-Item has by far the most spread recommendations.
+  It is strange that User-User would get less spread as it learns more about users.
+  This implies that the more it learns about users, the less spread out its recommendations are.
+  This could be some form of a regression towards the mean, where as it learns more about users its recommendations become less personalized to the users individual quirks.
+  More research will be needed to understand this trend.
+
 \section{Conclusions}

### 2014-04-24 paper/main.tex
done with results, figures tenatively added, after next run will commit images and redirect

@@
   % Prediction accuracy
+\begin{figure*}
+  \centering
+  \includegraphics[width=\linewidth]{/export/scratch/kluver/workspace/coldstartrecommendation/lenskit/output/ekstrandTuning/accuracy.pdf}
+  \caption{TODO}
+  \label{fig:rmse}
+  \label{fig:ndcg}
+\end{figure*}
     % RMSE plot
-
-TODO: rmse plot
-
     % point out error bars etc.
@@
 
+\begin{figure*}[ht!]
+  \centering
+  \includegraphics[width=1.1\linewidth]{/export/scratch/kluver/workspace/coldstartrecommendation/lenskit/output/ekstrandTuning/TopNPrecision.pdf}
+  \caption{TODO}
+  \label{fig:map}
+\end{figure*}
+
   Figure \ref{fig:map} shows MAP@20, precision@20, and fallout@20.
@@
     % interpret graph and reason about what it means about quality
+
+\begin{figure*}[ht!]
+  \centering
+  \includegraphics[width=1.1\linewidth]{/export/scratch/kluver/workspace/coldstartrecommendation/lenskit/output/ekstrandTuning/rmse_20.pdf}
+  \caption{TODO}
+  \label{fig:topN.rmse}
+\end{figure*}
+
   The first graph in figure \ref{fig:topN.rmse} shows the average number of items in the top 20 recommendations for each algorithm that were also in the test set.
@@
   % Other recommender metrics
+
+\begin{figure*}[ht!]
+  \centering
+  \includegraphics[width=\linewidth]{/export/scratch/kluver/workspace/coldstartrecommendation/lenskit/output/ekstrandTuning/popdiv.pdf}
+  \caption{TODO}
+  \label{fig:pop}
+\end{figure*}
+
     % popularity graph
@@
   Overall, Item-Item and SVD are providing recommendations that, at least in an offline context, appear to have quite comparable diversity.
+
+\begin{figure}[ht!]
+  \centering
+  \includegraphics[width=1.1\columnwidth]{/export/scratch/kluver/workspace/coldstartrecommendation/lenskit/output/ekstrandTuning/topN_entropy.pdf}
+  \caption{TODO}
+  \label{fig:spread}
+\end{figure}
 

### 2014-04-24 lenskit/src/eval/chart.R

@@ -9,18 +9,78 @@ nonMetrics = c("Algorithm","Retain", "DataSet", "User", "Partition", "NGood", "T
 metrics = setdiff(names(all.data),nonMetrics)
 
 for (metric in metrics) {
-  plot = ggplot(all.data, aes_string(x="Retain", y=metric, color="Algorithm", shape="Algorithm"))+stat_summary(fun.y=mean, geom="line")+stat_summary(fun.y=mean, geom="point") + stat_summary(fun.data = mean_cl_normal, geom="errorbar")
+  plot = ggplot(all.data, aes_string(x="Retain", y=metric, color="Algorithm", shape="Algorithm"))+stat_summary(fun.y=mean, geom="line")+stat_summary(fun.y=mean, geom="point") + stat_summary(fun.data = mean_cl_normal, geom="errorbar")+scale_x_continuous(breaks=4*(0:5))
   metric = gsub('\\.', '_', metric)
   filename = paste(metric,".pdf",sep="")
   cat("Outputting to",filename,"\n")
-  pdf(filename, width=5.5, height=4.25)
+  pdf(filename, width=4, height=3)
   print(plot)
   dev.off()
 }
 
 
 dat <- read.csv("eval-results.csv")
-plot = ggplot(dat, aes(x=Retain, y=TopN.Entropy, color=Algorithm, shape=Algorithm))+geom_point()+stat_summary(fun.y=mean, geom="line")
-pdf("topN_entropy.pdf", width=5.5, height=4.5)
+plot = ggplot(dat, aes(x=Retain, y=TopN.Entropy, color=Algorithm, shape=Algorithm))+geom_point()+stat_summary(fun.y=mean, geom="line")+ylab("Spread")+scale_x_continuous(breaks=4*(0:5))
+pdf("topN_entropy.pdf", width=6, height=4)
+print(plot)
+dev.off()
+
+library(reshape2)
+
+all.data.melt = melt(all.data, c("Algorithm", "Retain", "User"), metrics)
+melt.mean = aggregate(value~Algorithm+Retain+variable, data=all.data.melt, mean)
+melt.var = aggregate(value~Algorithm+Retain+variable, data=all.data.melt, var)
+names(melt.var)<- c("Algorithm","Retain","variable","var")
+melt.length = aggregate(value~Algorithm+Retain+variable, data=all.data.melt, length)
[…]

### 2014-04-25 paper/main.tex
done working for the day

--- 2014-04-24 paper/main.tex
@@
 \section{Methodology}
+\label{sec:methodology}
   % (high level description/restatement of approach) To answer our research questions we will various metrics of recommenders in an offline setting simulations users as they join the system.
@@
     % motiviate _why do a simulation_ (a lot of metrics we want to try)
-  % temporal replay evaluation 
-    % similar technique, harder to do than ours, and with a different goal.
-  % SOMEWHERE AROUND HERE talk about drnner and adaptive bootstrapping and 10 isenough paper's ad hoc plots by datapoint size.
+
+  To answer our research questions we will evaluate algorithms in an offline simulation designed to simulate users as they join the system.
+  We will do this by running separate instances of the standard train and test methodology, each one designed so that test users have some small number of ratings.
+  We will then plot metric results against the number of retained items to understand both how the algorithms compare on these metrics, and how the algorithm performance changes with the number of ratings provided.
+  This, or a similar technique, has been used by previous authors \cite{DrennerInitialExperiance, TenIsEnough, AdaptiveBootstrap}, however, these normally are used with only one algorithm or metric, and normally just to make the point that prediction accuracy improves as the number of ratings grows.
+  We will use this technique to develop a much more nuanced understanding of the differences between algorithms, and how they behave for users with few ratings.
+
+  \subsection*{Algorithms}
   % algorithms (discuss tuning of algorithms per algorithm?)
-    % item mean baseline
-    % user adjusted baseline (@michael, is there a cite for this?)
-    % user-user
-    % item-item
-    % SVD
+  % motivation for algorithm selection
+  % list algorithms
+
+  We seek to develop an understanding of how a range of different standard algorithms perform at for new users.
+  Rather than picking advanced algorithms that are tuned for new user performance, we decided to first understand how representatives of three common types of algorithms perform.
+  Therefore we will compare User-User, Item-Item, and Funk's SVD algorithms against two simple baseline algorithms.
+  More information about these algorithms can be found in table \ref{tbl:algo}
+  Code for our evaluation, using the Lenskit evaluation framework \cite{lenskit} is available at \url{https://bitbucket.org/kluver/coldstartrecommendation}.
+
+  \begin{table}
+    \centering
+    \begin{tabular}{|p{6em}|p{18em}|}
+      \hline
+      Algorithm        & Algorithm description                                                                                     \\\hline
+      User-User        & User based nearest neighbor collaborative filtering \cite{resnick1994grouplens}                           \\\hline
+      Item-Item        & Item based nearest neighbor collaborative filtering \cite{sarwar2001item}                                 \\\hline
+      SVD              & Simon Funk's SVD based collaborative filtering approach \cite{funk_netflix_2006}                          \\\hline
+      ItemBaseline     & Item's average rating, slight mean damping is applied                                                     \\\hline
+      UserItem\-Baseline & ItemBasline adjusted by the user's average offset from the ItemBaseline. slight mean damping is applied \\\hline
+    \end{tabular}
+    \caption{Summary of algorithms used in our research.}
+    \label{tbl:algo}
+  \end{table}
+
+  % evaluation and code to run these algortihms available at.
+  % citations and lenskit info
+  %need new topic sentence.
+
+  Each algorithm was tuned using the values from Ekstrand's 2012 lenskit evaluation \cite{ekstrand2012recommenders}.
+  Sine this evaluation was on a slightly different dataset (MovieLens 10M), we performed only minor tuning from these parameters.
+  Most notably we found the tuning of the damping parameter used in the baseline recommenders to be too large, which limited accuracy at cold start.
+  Overall, we did not find our qualitative results to be highly sensitive to minor tuning changes, although specific values may change slightly.
+
+  \subsection*{Metrics}
+
+  % motivate many metrics.
   % metrics (expected / good behavior for each metric (I.E. this should go down, any reasonable algorithm should always beet baseline)) 
@@
     % distributions of # ratings per user (important: min 20)
+
+  % temporal replay evaluation 
+    % similar technique, harder to do than ours, and with a different goal.
+
+  %TODO: restructure this, I can't have my cake and eat  too. This needs ot be refocused to explicitly focus on similarities and differences.
+  %One related approach to the one we take here is to evaluate recommender systems temporally.
+  %Various authors have taken this approach \cite{KorenTemporal, LathiaTemporal, burke2010evaluating} to evaluating recommender systems to understand how recommenders change with time.
+  %This work often features similar evaluations to ours, focusing on time (life of user or life of system) as opposed to number of ratings.
+  %One major difference between this approach and the one we take is that using ratings in the order that they were given can expose algorithms to biases from the system they were collected in.
+  %For example, in the MovieLens system users enter their first 15 ratings in a different interface, which does not give them choice over which items to rate.
+  %This is a forced situation which produces qualitatively different ratings than normal ratings.
+  %For this work we want to avoid this effect and therefor must use randomization.
+  %TODO: SOMEWHERE AROUND HERE THIS GOES!
+
   % subsampling strategy
@@
 
-  Using the methodology described in \cite{sec:methodology} we compute a value for each metric on each of the five algorithms.
+  Using the methodology described in section \ref{sec:methodology} we compute a value for each metric on each of the five algorithms.
   We plot these results against the number of items retained.
@@
   % Prediction accuracy
+\begin{figure*}
+  \centering
+
+  \includegraphics[width=\linewidth]{../lenskit/output/ekstrandTuned20/accuracy.pdf}
+  \caption{TODO}
+  \label{fig:rmse}
+  \label{fig:ndcg}
+\end{figure*}
     % RMSE plot
-
-TODO: rmse plot
-
     % point out error bars etc.
@@
 
+\begin{figure*}[ht!]
+  \centering
+  \includegraphics[width=1.1\linewidth]{../lenskit/output/ekstrandTuned20/TopNPrecision.pdf}
+  \caption{TODO}
+  \label{fig:map}
+\end{figure*}
+
   Figure \ref{fig:map} shows MAP@20, precision@20, and fallout@20.
@@
     % interpret graph and reason about what it means about quality
+
+\begin{figure*}[ht!]
+  \centering
+  \includegraphics[width=1.1\linewidth]{../lenskit/output/ekstrandTuned20/rmse_20.pdf}
+  \caption{TODO}
+  \label{fig:topN.rmse}
+\end{figure*}
+
   The first graph in figure \ref{fig:topN.rmse} shows the average number of items in the top 20 recommendations for each algorithm that were also in the test set.
@@
   % Other recommender metrics
+
+\begin{figure*}[ht!]
+  \centering
+  \includegraphics[width=\linewidth]{../lenskit/output/ekstrandTuned20/popdiv.pdf}
+  \caption{TODO}
+  \label{fig:pop}
+\end{figure*}
+
     % popularity graph
@@
   Overall, Item-Item and SVD are providing recommendations that, at least in an offline context, appear to have quite comparable diversity.
+
+\begin{figure}[ht!]
+  \centering
+  \includegraphics[width=1.1\columnwidth]{../lenskit/output/ekstrandTuned20/topN_entropy.pdf}
+  \caption{TODO}
+  \label{fig:spread}
+\end{figure}
 

### 2014-04-30 paper/main.tex
wrote some intro stuff I guess. left an alt approach in comments, will want to clean up later

@@
 \documentclass[letterpaper]{sig-alternate}
+
+\usepackage{hyperref}
+%  \multicolumn{2}{|c|}{Team sheet} \\
 
@@
 (subsection might need re-ordering)
+
+  Our work sits at the intersection of two important traditions of recommender systems research: new user experience, and offline algorithm comparison.
+  Our goal is to understand how different algorithms behave for new users, especially how their behavior changes as new users add ratings into the system.
+  While there has been much work comparing algorithms in an offline setting, very little of it has explored what are the best algorithms for users with very few ratings.
+  Furthermore, we do not seek to find the algorithm that scores best on some metric of choice, instead, we seek to form a complete understanding of the differences between algorithms for cold start use.
+
+  We structure our work around the following research questions.
+  \begin{itemize}
+    \item RQ1: How do algorithms behave for users with very few ratings?
+    \item RQ2: How does algorithm behavior change with increasing ratings for new users?
+  \end{itemize}
+  Unfortunately, algorithm behavior is a broad and hard to define subject.
+  We will look at three distinct ways to quantify the behavior of a recommender system, each of which may lead to multiple metrics.
+%  One of the most traditional ways to measure the behavior of a recommendation system is to measure its performance on the rating task.
+%  This is normally done with metrics such as RMSE or MAE.
+%  The accuracy of the algorithm may be important at cold start as users my judge the recommender on how well it predicts items they have not yet rated.
+%  This early judgment may be very important to the user.
+
+%  Recently, work in the recommender system community has been shifting away from prediction accuracy to start looking at how well the system ranks and recommends items.
+%  This change was, in part prompted by work criticizing the practice of measuring only the RMSE of recommender systems \cite{TODO:rmse considered harmful...}
+%  The quality of the ranking provided by the recommender has normally been measured with information retrial metrics such as nDCG.
+%  The quality of the recommendations has normally been measured by looking at which sort of items end up being recommended with metrics such as precision@N, recall@N, fallout@N etc.
+%  We will also look at some simpler metrics along with these such as simply evaluating the ratings and predictions that end up in top 20 recommendations.
+
+%  Finally, we find that
+
 % RQs
@@
           % spread
+
+  Prediction accuracy, as measured by RMSE or MAE is one of the classic techniques in recommended system evaluation.
+  The idea behind these evaluations is to measure the accuracy of predictions on withheld recommendations.
+  If the algorithm can successfully predict the withheld ratings, then we expect it will accurately predict future ratings as well.
+  Because users may build trust in the system by looking at the predictions of items, we are interested in knowing which algorithms are most accurate, and which show the most notable change in accuracy as ratings are entered.
+
+  More recently, work as shifted away from accuracy metrics in favor of metrics that evaluate a recommender systems ability to rank, or recommend.
+  % TODO: cite johnjoe here?
+  These algorithms try to evaluate how useful the recommendations provided by the system are to the user.
+  This is commonly done by looking either at how withheld items are ranked with metrics such as nDCG, or by looking at if good or bad withheld items tend to be recommended positively by the recommender.
+
+  Prior work \cite{ziegler and martijn} has shown that users are sensitive to properties of recommendations that are not as trivially good or bad as those above.
+  We will look at three properties that can be measured in an offline context, the popularity, diversity, and spread of recommendations.
+
+  We expect that different algorithms will recommend items that are more or less popular.
+  This difference could explain feelings that a list isn't personal enough (too popular), or is too novel / obscure (not popular enough).
+
+  Past work \cite{ziegler and martijn} has shown that users tend to prefer recommendation lists that are more diverse.
+  <TODO: skim this past work again and report some details/results.>
+
+  Finally, we expect that algorithms with a good understand of its users will be able to recommend different items to different users.
+  To measure this we will look at how well the recommender spreads its attention of many items.
+  We will do this by computing the Shannon's entropy of the items recommended to each user.
+  If the recommender system recommends the same items to each user, then the entropy will be low.
+  Conversely, if the recommender recommends different items to each user, then the entropy will high.
+  In practice, a recommender may focus on both a shared core of popular or well reviewed items, and more personal items, which would lead to middle values.
+
+  Based on this list of possible metrics, we will clarify our research questions with three sub questions:
+  \begin{\itemize}
+  \item How do algorithms behave with respect to prediction accuracy?
+  \item How do algorithms behave with respect to ranking items appropriately?
+  \item How do algorithms behave as measured by other metrics such as popularity and diversity?
+  \end{\itemize}
+
 % RQ1A  How does algorithm behavior change with respect to predictive accuracy?
@@
 
-  Using the methodology described in section \ref{sec:methodology} we compute a value for each metric on each of the five algorithms.
-  We plot these results against the number of items retained.
-  These plots allow us to compare the algorithms against each other, and also, to understand how each algorithm's performance changes with more data.
-  Each plot will have error bars showing the 95\% confidence intervals around the average per use value for the metric.
+  Using the methodology described in section \ref{sec:methodology} we generate plots for each metric showing how the five algorithms perform for between 0 and 19 ratings.
+  Where appropriate, we will compute metrics independently for each user, and plot the average user value, with a 95\% confidence interval around this mean.
+  Where this cannot be done we will compute the metric once for each fold, and plot the values per fold, with a line intersecting the average.
 
@@
   \includegraphics[width=\linewidth]{../lenskit/output/ekstrandTuned20/accuracy.pdf}
-  \caption{TODO}
+  \caption{Left: RMSE, Right: nDCG}
   \label{fig:rmse}

### 2014-05-01 paper/main.tex
some work I guess

@@
 \section{Introduction}
-  % describe new user experience in recommender systems.
-  When someone first interacts with any intelligent system, the system knows nothing about them.
-TODO:
+  % describe new user experience in in recommender systems.
+  One of the important problems in recommender systems is the new user experiance.
 
@@
     % talk about first impressions and user trust
+  How a recommender system handels new users can change how these users interact with the recommender system.
+
+%%%%%%%%%%%%%%%%%
   The new user experience is a very important part of any recommender system.
@@
   When users are new the system will not have much information to personalize the user experience which may lead the user to leave the system.
+%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-  Past work has referred to the new user experience issue as the \emph{user cold start} problem.
-  This work has normally focused on two
-TODO:
+  % past work tends to focus on the algorithm side, the algorithm side of this problem is cold start.
+    % cold start algorithms is a larger problem
+    % cold start algorithms tend to use extra information to help imrpove eary performance.
+    % we only look at user cold start as it is ubiquitous.
+  Past work on new user experiance in recommender systems has tended to focus on the algorithmic cold start problem.
+
+  % One common new user strategy is to have a special data collection mode.
+  % collect ratings fast to get over a precived ``bad period''
+  % work on this focuses on how to collect useful ratings.
+  % (framework from decision tree work)
+  The best studied strategy for new users is to have users go through a new user survey.
+
+  % One promising approach is to let user control interaction
+  % user or system driven control.
+  % promising, we don't know much about follow up.
+  One promissing approach to new users is to have users control the new user survey.
 

### 2014-05-03 paper/main.tex
some work on outlineing and small eval changes

@@
 \usepackage{hyperref}
-%  \multicolumn{2}{|c|}{Team sheet} \\
 
@@
     % talk about first impressions and user trust
+    % How an algorithm behaves the first time a user sees it is important (first impression)
   How a recommender system handels new users can change how these users interact with the recommender system.
-
-%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%
   The new user experience is a very important part of any recommender system.
@@
 %%%%%%%%%%%%%%%%%%%%%%
-
-  % Past work on cold start has focused on… (how long should this section be?)
-    % ways to sort items to get to most value out of early ratings
-    % research looking at user experience and other forms of data collection
 
@@
   % (framework from decision tree work)
-  The best studied strategy for new users is to have users go through a new user survey.
+  The best studied strategy for new users is to have users go through a new user rating survey.
 
@@
   % promising, we don't know much about follow up.
-  One promissing approach to new users is to have users control the new user survey.
+  One promissing approach to new users is to have users control the new user rating survey.
 
-  % How an algorithm behaves the first time a user sees it is important (first impression)
+  % bridge from past work to our work.
+    % point out holes (evaluating standard algorithms for little to no ratings, wholeistic evaluation)
+    % point out that other approaches than smart object ordering exist, and we need to be able to compare.
+    % cite the comparison work at CHI.
+    % we propose to evaluate 
+  While past work has covered the new user rating survey as a solution to new user cold start, much less work has been done comparing other strategies.
+
   % Past work looks carefully at issues X, Y, and Z, but doesn’t take a systematic look at the behavior of algorithms for users with few ratings.
@@
   Our work sits at the intersection of two important traditions of recommender systems research: new user experience, and offline algorithm comparison.
+%%%%%%%%%%%%%%%%%%%%%%
   Our goal is to understand how different algorithms behave for new users, especially how their behavior changes as new users add ratings into the system.
@@
   Furthermore, we do not seek to find the algorithm that scores best on some metric of choice, instead, we seek to form a complete understanding of the differences between algorithms for cold start use.
+%%%%%%%%%%%%%%%%%%%%%%
 
@@
   We will look at three distinct ways to quantify the behavior of a recommender system, each of which may lead to multiple metrics.
-%  One of the most traditional ways to measure the behavior of a recommendation system is to measure its performance on the rating task.
-%  This is normally done with metrics such as RMSE or MAE.
-%  The accuracy of the algorithm may be important at cold start as users my judge the recommender on how well it predicts items they have not yet rated.
-%  This early judgment may be very important to the user.
-
-%  Recently, work in the recommender system community has been shifting away from prediction accuracy to start looking at how well the system ranks and recommends items.
-%  This change was, in part prompted by work criticizing the practice of measuring only the RMSE of recommender systems \cite{TODO:rmse considered harmful...}
-%  The quality of the ranking provided by the recommender has normally been measured with information retrial metrics such as nDCG.
-%  The quality of the recommendations has normally been measured by looking at which sort of items end up being recommended with metrics such as precision@N, recall@N, fallout@N etc.
-%  We will also look at some simpler metrics along with these such as simply evaluating the ratings and predictions that end up in top 20 recommendations.
-
-%  Finally, we find that
 
@@
   Prediction accuracy, as measured by RMSE or MAE is one of the classic techniques in recommended system evaluation.
+%%%%%%%%%%%%%%%%%%%%%%
   The idea behind these evaluations is to measure the accuracy of predictions on withheld recommendations.
@@
   Because users may build trust in the system by looking at the predictions of items, we are interested in knowing which algorithms are most accurate, and which show the most notable change in accuracy as ratings are entered.
+%%%%%%%%%%%%%%%%%%%%%%
 
-  More recently, work as shifted away from accuracy metrics in favor of metrics that evaluate a recommender systems ability to rank, or recommend.
+  More recently, the research community has shifted away from accuracy metrics in favor of metrics that evaluate a recommender system's ability to rank or recommend.
+%%%%%%%%%%%%%%%%%%%%%%
   % TODO: cite johnjoe here?
@@
   This is commonly done by looking either at how withheld items are ranked with metrics such as nDCG, or by looking at if good or bad withheld items tend to be recommended positively by the recommender.
+%%%%%%%%%%%%%%%%%%%%%%
 
-  Prior work \cite{ziegler and martijn} has shown that users are sensitive to properties of recommendations that are not as trivially good or bad as those above.
-  We will look at three properties that can be measured in an offline context, the popularity, diversity, and spread of recommendations.
+  Prior work has shown that users are sensitive to properties of a recommender other than simply how well it performs.
+%%%%%%%%%%%%%%%%%%%%%%
+  These properties, such as the diversity of the output recommendations, can effect how the user interacts with the recommender and drastically change the user experience \cite{ziegler and martijn}.
+  Unfortunately, there is no way to know all such interesting metrics
+  We will look at three such metrics, the popularity, diversity, and spread of the recommendations.
+%%%%%%%%%%%%%%%%%%%%%%
 
-  We expect that different algorithms will recommend items that are more or less popular.
-  This difference could explain feelings that a list isn't personal enough (too popular), or is too novel / obscure (not popular enough).
+  The popularity of the items recommended to a user can have an impact on the user's satisfaction.
+%%%%%%%%%%%%%%%%%%%%%%
+  If the user is only recommended items that are very popular, then the recommender is likely failing to make novel and interesting recommendations.
+  In this case the user may feel that the recommender is not helpful in finding new and interesting things to watch.
+  However, if a new user is only recommended items that are not well known, it is unlikely that the user will have heard of the items recommended.
+  This could cause the user to not trust the system, as they have no way of comparing the recommendations against their personal knowledge.
+%%%%%%%%%%%%%%%%%%%%%%
 
   Past work \cite{ziegler and martijn} has shown that users tend to prefer recommendation lists that are more diverse.
+%%%%%%%%%%%%%%%%%%%%%%
   <TODO: skim this past work again and report some details/results.>
+%%%%%%%%%%%%%%%%%%%%%%
 
   Finally, we expect that algorithms with a good understand of its users will be able to recommend different items to different users.
+%%%%%%%%%%%%%%%%%%%%%%
   To measure this we will look at how well the recommender spreads its attention of many items.
@@
   In practice, a recommender may focus on both a shared core of popular or well reviewed items, and more personal items, which would lead to middle values.
+%%%%%%%%%%%%%%%%%%%%%%
 
   Based on this list of possible metrics, we will clarify our research questions with three sub questions:
-  \begin{\itemize}
+  \begin{itemize}
   \item How do algorithms behave with respect to prediction accuracy?
@@
   \item How do algorithms behave as measured by other metrics such as popularity and diversity?
-  \end{\itemize}
+  \end{itemize}
 
@@
   To answer our research questions we will evaluate algorithms in an offline simulation designed to simulate users as they join the system.
+%%%%%%%%%%%%%%%%%%%%%%
   We will do this by running separate instances of the standard train and test methodology, each one designed so that test users have some small number of ratings.
@@
   We will use this technique to develop a much more nuanced understanding of the differences between algorithms, and how they behave for users with few ratings.
+%%%%%%%%%%%%%%%%%%%%%%
 
@@
   We seek to develop an understanding of how a range of different standard algorithms perform at for new users.
+%%%%%%%%%%%%%%%%%%%%%%
   Rather than picking advanced algorithms that are tuned for new user performance, we decided to first understand how representatives of three common types of algorithms perform.
@@
   Code for our evaluation, using the Lenskit evaluation framework \cite{lenskit} is available at \url{https://bitbucket.org/kluver/coldstartrecommendation}.
+%%%%%%%%%%%%%%%%%%%%%%
 
@@
   Each algorithm was tuned using the values from Ekstrand's 2012 lenskit evaluation \cite{ekstrand2012recommenders}.
-  Sine this evaluation was on a slightly different dataset (MovieLens 10M), we performed only minor tuning from these parameters.
-  Most notably we found the tuning of the damping parameter used in the baseline recommenders to be too large, which limited accuracy at cold start.
-  Overall, we did not find our qualitative results to be highly sensitive to minor tuning changes, although specific values may change slightly.
+%%%%%%%%%%%%%%%%%%%%%%
+  Since this evaluation was on a slightly different dataset (MovieLens 10M), we performed only minor tuning from these parameters.
+  Most notably we found the tuning of the damping parameter used in the baseline recommenders to be too large for this dataset.
+  As the damping can have a large effect on baseline quality for new users, we adjusted to a more reasonable damping for this dataset.
+  Overall, we did not find our results to be highly sensitive to minor tuning changes, although specific values may change slightly.
+%%%%%%%%%%%%%%%%%%%%%%
 
   \subsection*{Metrics}
+  We will discuss a total of eleven metrics.
+%%%%%%%%%%%%%%%%%%%%%%
+  Some of these metrics will prove to be redundant, and future work should be able to use a much smaller list.
+  The metrics are listed in table \ref{tbl:metrics}.
+  \begin{table}[ht!]
+    \centering
+    \begin{tabular}{|p{7em}|p{16em}|}
+      \hline
+      \multicolumn{2}{|c|}{{\bf Accuracy Metrics}} \\\hline
+      RMSE                   & Standard prediction accuracy metric. \\\hline
+      \hline
+      \multicolumn{2}{|c|}{{\bf Recommendation Metrics}} \\\hline
+      nDCG                   & Standard ranking quality evaluation.\\\hline
+      Precision@20           & Count of recommended items in the test set and rated 4.0 or higher.\\\hline
+      MAP@20                 & Average of precision at each rank one through twenty.\\\hline
+      Fallout@20             & Count of recommended items in the test set and rated 2.0 or lower.\\\hline
+      SeenItems@20           & Count of recommended items in the test set.\\\hline
+      MeanRating@20          & Average rating of recommended items that are in the test set.\\\hline
+      RMSE@20                & RMSE computed only over those items that were recommended.\\\hline
+      \hline
+      \multicolumn{2}{|c|}{{\bf Other Metrics}} \\\hline
+      Average\-Popularity@20 & The average number of users in the training set that have seen recommended items.\\\hline
+      AILS@20                & The average pairwise similarity between recommended items. \\\hline
+      Spread@20              & The entropy of the distribution of recommended items for users in the test set.\\\hline
+    \end{tabular}
+    \caption{TODO:}
+    \label{tbl:metrics}
+  \end{table}
+
+  All metrics evaluating recommendations (all but RMSE and nDCG) are based on the top 20 predicted items for the user, excluding items in the training set.
+  The AILS@20 metric is based on the ILS metric introduced by ziegler et al \cite{ziegler}.
+  AILS@N is simply a rescaling of ILS to be scale free with regards to N, this was done to make the metric easier to interpret, by putting it on the same scale as similarity values.
+  The choice of 20 items for the evaluations was arbitrary, testing with other values showed similar results, except on one metric.
+  This issue will be addressed later in the results section.
+
+%This is simply a rescaling of the ILS metric from \cite{ziegler}.
+Similarity is computed by subtracting the average item rating from the item vectors for each item and then computing the cosine similarity.\\\hline
 
@@
 
+  \subsection*{Dataset}
   % dataset
@@
 
-  % temporal replay evaluation 
-    % similar technique, harder to do than ours, and with a different goal.
+  \subsection*{Evaluation Procedure}
 

### 2014-05-03 lenskit/src/eval/chart.R

@@ -50,7 +50,7 @@ data.melt$variable[data.melt$variable == "TopN.RMSE.seenItems"] = "SeenItems@20"
 data.melt$variable[data.melt$variable == "TopN.RMSE"] = "RMSE@20"
 data.melt$variable[data.melt$variable == "TopN.Ave.Rat"] = "MeanRating@20"
 data.melt$variable[data.melt$variable == "TopN.MeanPopularity"] = "AveragePopularity@20"
-data.melt$variable[data.melt$variable == "diversity"] = "Diversity@20"
+data.melt$variable[data.melt$variable == "diversity"] = "AILS@20"
 data.melt$variable = ordered(data.melt$variable, levels = c("Coverage", "RMSE", "nDCG", "Precision@20", "MAP@20", "Fallout@20", "SeenItems@20", "MeanRating@20", "RMSE@20", "AveragePopularity@20", "Diversity@20", "TestTime","MAE","MutualInformation","PredicitonEntropy","ItemScorer.NAttempted","ItemScorer.NGood","ItemScorer.Coverage","TopN.ActualLength", "Recall.fallout", "Recall"))
 
 # make plots
@@ -78,7 +78,7 @@ print(plot)
 dev.off()
 
 
-vars = c("AveragePopularity@20", "Diversity@20")
+vars = c("AveragePopularity@20", "AILS@20")
 data.melt.sub = subset(data.melt, variable %in% vars)
 plot = ggplot(data.melt.sub, aes(x=Retain, y=value, color=Algorithm, shape=Algorithm, ymin=lb, ymax=ub))+geom_point()+geom_line()+geom_errorbar()+scale_x_continuous(breaks=4*(0:5))+facet_wrap(~variable, scales="free_y", nrow=1)
 pdf("popdiv.pdf", width=8.5, height=3.5)

### 2014-05-03 lenskit/src/main/java/TopNDiversityMetric.java

@@ -97,7 +97,7 @@ public class TopNDiversityMetric extends AbstractMetric<MeanAccumulator, TopNDiv
                     continue;
                 }
                 SparseVector v2 = context.itemVector(i2);
-                simSum -= sim.similarity(i1,v1,i2,v2);
+                simSum += sim.similarity(i1,v1,i2,v2);
             }
         }
 

### 2014-05-04 paper/main.tex
outlined or written first two sections, secction three is written, but needs extensive review, possible re-outlining.

@@
     % cite the comparison work at CHI.
-    % we propose to evaluate 
   While past work has covered the new user rating survey as a solution to new user cold start, much less work has been done comparing other strategies.
+
+  % Temporal evaluation
+    % similar in-so-much as it can provide similar answers (growth over time), 
+    % different in-so-much as it tends to be used for just system growth, but it could be used for user growth.
+  Some researchers have taken to perfforming temporal evaluations of recommender systems, looking at accuracy metrics as a function of the lifetime of a system.
+
+  %TODO: restructure this, I can't have my cake and eat  too. This needs ot be refocused to explicitly focus on similarities and differences.
+  %One related approach to the one we take here is to evaluate recommender systems temporally.
+  %Various authors have taken this approach \cite{KorenTemporal, LathiaTemporal, burke2010evaluating} to evaluating recommender systems to understand how recommenders change with time.
+  %This work often features similar evaluations to ours, focusing on time (life of user or life of system) as opposed to number of ratings.
+  %One major difference between this approach and the one we take is that using ratings in the order that they were given can expose algorithms to biases from the system they were collected in.
+  %For example, in the MovieLens system users enter their first 15 ratings in a different interface, which does not give them choice over which items to rate.
+  %This is a forced situation which produces qualitatively different ratings than normal ratings.
+  %For this work we want to avoid this effect and therefor must use randomization.
+  %TODO: SOMEWHERE AROUND HERE THIS GOES!
+
+  % inspired by this work...
 
@@
       AILS@20                & The average pairwise similarity between recommended items. \\\hline
-      Spread@20              & The entropy of the distribution of recommended items for users in the test set.\\\hline
+      Spread@20              & The Shannon's entropy of the distribution of recommended items for users in the test set.\\\hline
     \end{tabular}
@@
   All metrics evaluating recommendations (all but RMSE and nDCG) are based on the top 20 predicted items for the user, excluding items in the training set.
-  The AILS@20 metric is based on the ILS metric introduced by ziegler et al \cite{ziegler}.
-  AILS@N is simply a rescaling of ILS to be scale free with regards to N, this was done to make the metric easier to interpret, by putting it on the same scale as similarity values.
   The choice of 20 items for the evaluations was arbitrary, testing with other values showed similar results, except on one metric.
@@
 
-%This is simply a rescaling of the ILS metric from \cite{ziegler}.
-Similarity is computed by subtracting the average item rating from the item vectors for each item and then computing the cosine similarity.\\\hline
+  % Provide computation details for AILS@N
+  Most of our metrics are standard techniques, or easily commutable, therefore we will only comment on the AILS, and spread metrics.
+%%%%%%%%%%%%%%%%%%%%%%%%%
+  The AILS@20 metric is based on the ILS metric introduced by ziegler et al \cite{ziegler}.
+  AILS@N is simply a rescaling of ILS to be scale free with regards to N, this was done to make the metric easier to interpret, by putting it on the same scale as similarity values.
 
-  % motivate many metrics.
-  % metrics (expected / good behavior for each metric (I.E. this should go down, any reasonable algorithm should always beet baseline)) 
-    % accuracy metrics 
-      % RMSE
-    % recommendation performance metrics (list subject to change)
-      % ndcg
-      % MAP@N (not yet measured)
-      % precision@N
-      % fallout @ N
-      % RMSE of recommended items
-      % (note that we looked at MRR and recall, but found them to behave the same).
-    % other properties
-      % recommendation average popularity
-      % recommendation diversity
-      % spread (entropy of recommendations) (related to recommendation coverage)
+  The spread metric is computed by computing recommendations for each test user.
+  Let $C(I)$ be the count of how many times item $I$ showed up in the recommendations.
+  Let $P(I) = C(I) / \sum_I C(I)$ be the probability that a random recommendation is item $I$.
+  Finally we define spread as the Shannon's entropy of $P(I)$, $Spread = -\sum_I P(I) \log(P(I))$.
+%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
     % distributions of # ratings per user (important: min 20)
+    % establish range of evaluation.
+  For this evaluation we will use the MovieLens 1M dataset.
+
+  % movielens system
+      % silly cold start
+      % randomization
+  The MovieLens system uses a new user rating survey to allow users to enter the system.
 
   \subsection*{Evaluation Procedure}
-
-  %TODO: restructure this, I can't have my cake and eat  too. This needs ot be refocused to explicitly focus on similarities and differences.
-  %One related approach to the one we take here is to evaluate recommender systems temporally.
-  %Various authors have taken this approach \cite{KorenTemporal, LathiaTemporal, burke2010evaluating} to evaluating recommender systems to understand how recommenders change with time.
-  %This work often features similar evaluations to ours, focusing on time (life of user or life of system) as opposed to number of ratings.
-  %One major difference between this approach and the one we take is that using ratings in the order that they were given can expose algorithms to biases from the system they were collected in.
-  %For example, in the MovieLens system users enter their first 15 ratings in a different interface, which does not give them choice over which items to rate.
-  %This is a forced situation which produces qualitatively different ratings than normal ratings.
-  %For this work we want to avoid this effect and therefor must use randomization.
-  %TODO: SOMEWHERE AROUND HERE THIS GOES!
-
   % subsampling strategy
   % describe normal crossfold solution
+  The natural way to perform this evaluation is with a modified user crossfolding strategy.
+
   % describe issues with solution
+  Unfortunately, we found that this strategy can causes biases in our evaluation.
+
   % describe our crossfold then subsample solution.
+  Our solution to this bais is to downsample an existing crossfold.
 

### 2014-05-05 paper/main.tex
just now noticed that I misconfigured Item-Item. fixed and prepped for re-run

@@
+%TODO: fix citations.
+%TODO: review results
+%TODO: outline conclusions
+%TODO: write introduction
+%TODO: write methodology
+%TODO: re-write results
+%TODO: write conclusions
+%TODO: review extensively
 \documentclass[letterpaper]{sig-alternate}
@@
   % citations and lenskit info
-  %need new topic sentence.
 
-  Each algorithm was tuned using the values from Ekstrand's 2012 lenskit evaluation \cite{ekstrand2012recommenders}.
+  Both nearest neighbor algorithms, User-User and Item-Item subtract the UserItemBaseline score from the ratings before performing similarity calcuations
+  %TODO: verify this claim.
+
+  Each algorithm was configured and tuned similarly to Ekstrand's 2012 lenskit evaluation \cite{ekstrand2012recommenders}.
 %%%%%%%%%%%%%%%%%%%%%%
@@
 
+  % Provide computation details for spread
   The spread metric is computed by computing recommendations for each test user.
+%%%%%%%%%%%%%%%%%%%%%%%%%%  
   Let $C(I)$ be the count of how many times item $I$ showed up in the recommendations.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%
+
+  % provide expectations.
+  We expect that the item baseline algorithm will have approximately constant performance on all metrics.
+  Since the Item Baseline is such a simple algorithm we expect that any reasonable algorithm should be able to do better on it on the accuracy on recommendation metrics.
+  Furthermore, we expect the improvement over the item baseline to be gradual, with larger improvements only happening with more data.
+
+  Many of the algorithms are configured with 
+  We expect that our intelligent algorithms, SVD, Item-Item, and User-User should be able to beat the item user baseline after
 
@@
   % describe our crossfold then subsample solution.
-  Our solution to this bais is to downsample an existing crossfold.
+  Our solution to this bais is to subsample an existing crossfold.
 
@@
 
-  Using the methodology described in section \ref{sec:methodology} we generate plots for each metric showing how the five algorithms perform for between 0 and 19 ratings.
+  Using the methodology described in section \ref{sec:methodology} we generate plots for each metric showing how the five algorithms perform new users.
   Where appropriate, we will compute metrics independently for each user, and plot the average user value, with a 95\% confidence interval around this mean.
-  Where this cannot be done we will compute the metric once for each fold, and plot the values per fold, with a line intersecting the average.
+  Where this cannot be done we will compute the metric once for each of the five folds, and plot the values for each fold, with a lines through the average.
 

### 2014-05-05 lenskit/src/eval/chart.R

@@ -51,7 +51,7 @@ data.melt$variable[data.melt$variable == "TopN.RMSE"] = "RMSE@20"
 data.melt$variable[data.melt$variable == "TopN.Ave.Rat"] = "MeanRating@20"
 data.melt$variable[data.melt$variable == "TopN.MeanPopularity"] = "AveragePopularity@20"
 data.melt$variable[data.melt$variable == "diversity"] = "AILS@20"
-data.melt$variable = ordered(data.melt$variable, levels = c("Coverage", "RMSE", "nDCG", "Precision@20", "MAP@20", "Fallout@20", "SeenItems@20", "MeanRating@20", "RMSE@20", "AveragePopularity@20", "Diversity@20", "TestTime","MAE","MutualInformation","PredicitonEntropy","ItemScorer.NAttempted","ItemScorer.NGood","ItemScorer.Coverage","TopN.ActualLength", "Recall.fallout", "Recall"))
+data.melt$variable = ordered(data.melt$variable, levels = c("Coverage", "RMSE", "nDCG", "Precision@20", "MAP@20", "Fallout@20", "SeenItems@20", "MeanRating@20", "RMSE@20", "AveragePopularity@20", "AILS@20", "TestTime","MAE","MutualInformation","PredicitonEntropy","ItemScorer.NAttempted","ItemScorer.NGood","ItemScorer.Coverage","TopN.ActualLength", "Recall.fallout", "Recall"))
 
 # make plots
 vars = c("RMSE", "nDCG")

### 2014-05-05 paper/main.tex
prepped for re-re-re-run

@@
 
+  %TODO: verify these claims.
+  %TODO: check for examples of how to write this
+  % algorithm is configured as such.
+  %TODO: topic sentence.
+
   % evaluation and code to run these algortihms available at.
   % citations and lenskit info
-
-  Both nearest neighbor algorithms, User-User and Item-Item subtract the UserItemBaseline score from the ratings before performing similarity calcuations
-  %TODO: verify this claim.
-
-  Each algorithm was configured and tuned similarly to Ekstrand's 2012 lenskit evaluation \cite{ekstrand2012recommenders}.
+  Each algorithm was configured and tuned similarly to Ekstrand's 2012 lenskit based evaluation \cite{ekstrand2012recommenders}.
 %%%%%%%%%%%%%%%%%%%%%%

### 2014-05-05 paper/main.tex
minor work
some very small amout of work.

@@
   % describe new user experience in in recommender systems.
-  One of the important problems in recommender systems is the new user experiance.
+  One of the important problems in recommender systems is the new user experience.
+  Most recommenders only have one chance to impress new users.
+  If the user doesn't like the system, the user will leave.
+  Even if the user stays, how the recommender system behaves for these new users can set the user's expectations for how the system will work in the future.
 
@@
     % How an algorithm behaves the first time a user sees it is important (first impression)
-  How a recommender system handels new users can change how these users interact with the recommender system.
+  The user's first impression of the system can be very important in teaching the user to trust the system.
+  If the recommendation system does not perform well, the user will not trust its recommendations.
+  User trust in recommenders is a complex issue.
+  %TODO: re-skim the trust article.
+
 %%%%%%%%%%%%%%%%%%%%%%
-  The new user experience is a very important part of any recommender system.
-  The first interaction with a system is the only chance the system gets to make a first impression on the user.
-  This first impression can be very important in structuring a user's expectations for how the system will behave, and determining how much the user trusts the system's recommendations.
-  If the system behaves poorly when the user is new, then the user may leave and not come back.
   This is especially a concern with recommender systems which rely on large amounts of user information to make personalization decisions.
@@
   %TODO: check for examples of how to write this
+  %TODO: the Item-Item algo is ``better'' with just the item mean, not the user item mean. (change is pretty small). 
+  %TODO: exstrand's User-User only subtracts the user mean. I havn't tested this, he might have.
   % algorithm is configured as such.
@@
   % citations and lenskit info
-  Each algorithm was configured and tuned similarly to Ekstrand's 2012 lenskit based evaluation \cite{ekstrand2012recommenders}.
+  Each algorithm was tuned similarly to Ekstrand's 2012 lenskit based evaluation \cite{ekstrand2012recommenders}.
 %%%%%%%%%%%%%%%%%%%%%%

### 2014-05-05 paper/main.tex

--- 2014-05-05 paper/main.tex
@@
-%TODO: fix citations.
+%TODO: diversity config matches item-item config.
 %TODO: review results
@@
 
-\pdfpagewidth=8.5in
-\pdfpageheight=11in
+\pdfpagewidth = 8.5in
+\pdfpageheight = 11in
 
@@
 \section{Introduction}
-  % describe new user experience in in recommender systems.
-  One of the important problems in recommender systems is the new user experiance.
 
-  % new user experience is important, and here is why.
-    % talk about first impressions and user trust
-    % How an algorithm behaves the first time a user sees it is important (first impression)
-  How a recommender system handels new users can change how these users interact with the recommender system.
-%%%%%%%%%%%%%%%%%%%%%%
-  The new user experience is a very important part of any recommender system.
-  The first interaction with a system is the only chance the system gets to make a first impression on the user.
-  This first impression can be very important in structuring a user's expectations for how the system will behave, and determining how much the user trusts the system's recommendations.
-  If the system behaves poorly when the user is new, then the user may leave and not come back.
-  This is especially a concern with recommender systems which rely on large amounts of user information to make personalization decisions.
-  When users are new the system will not have much information to personalize the user experience which may lead the user to leave the system.
-%%%%%%%%%%%%%%%%%%%%%%
+  % new user experiance is only chance to make all important first impression
+  One of the important problems in recommender systems is the new user experience.
+  Every potential user of the system must at some point be a new user.
+  If the system cannot provide a good user experience, then the user is likely to leave and not return.
+  Even if the user does choose to enter the system, the user's first impression of the system can have a lasting impact on how the user interacts with the system.
+  The user's first impression can shape how the user expects the system to behave in future interactions, fundamentally changing how the user perceives the system.
+
+% one important trait is trust.
+    % trust is complex
+    % matching user's experiance is probably important.
+  One important part of how the user interacts with a recommender system is how much the user trusts the recommender system.
+  The more the user trusts a recommender system, the more the user can take advantage of it.
+  If the user doesn't trust a recommender system, the system can be of no value to the user.
+  Building trust with the user is therefore an important part of the new user experience.
+  Past work has shown that one of the strongest factors in the formation of trust is how much the system's recommendations match the user's knowledge \cite{wangAttributionOfTrust}.
+  For new users this can easily take the form of recommending items that the user has already seen.
+  %TODO: I kinda of want to cut this. I want to talk about this theme, but I don't actually know where it fits in.
 
   % past work tends to focus on the algorithm side, the algorithm side of this problem is cold start.
+    % SMAL DATA TENDS TO IMPLY BAD RESULTS.
     % cold start algorithms is a larger problem
@@
     % we only look at user cold start as it is ubiquitous.
-  Past work on new user experiance in recommender systems has tended to focus on the algorithmic cold start problem.
+  Past work on new user experience in recommender systems has tended to focus on the algorithmic cold start problem.
+  The cold start problem refers to the problem that no algorithm can make useful recommendations with very limited amounts of information.
+  While this applies to both new users and new items we will only discuss research on the new user cold start problem.
+  Most solutions to the algorithmic cold start problem require some source of extra information about new users.
 
@@
   The best studied strategy for new users is to have users go through a new user rating survey.
+  Systems that take this practice do not allow their users to receive recommendations until some fixed number of ratings have been entered.
+  The idea is that, after enough ratings have been entered, the predictions will be `good enough'.
+  This idea has been explored by many authors, most consider a system in which the user is presented with a list of items, and asked to rate items until they have enough ratings.
+  For a good discussion of prior work in this direction see \cite{adaptiveBootstrapping}.
+  Generally, this work strives to strike a balance between the amount of effort the user must put in before getting enough ratings, and the usefulness of each rating to the user.
+  This balance is complicated by how effort effects the user's perception of value \cite{TenIsEnough}.
 
@@
   % promising, we don't know much about follow up.
-  One promissing approach to new users is to have users control the new user rating survey.
+  One promising approach to the new user rating survey is to have users control which movies they rate.
+  This was pioneered by McNee et al., who looked at comparisons between user driven and system driven new user surveys \cite{mcneeInterfaces}.
+  They found that users who were in control of which items to rate during a new user survey rated fewer movies, but that the profiles they created were more useful to the system in predicting their future behavior.
+  Interestingly, while these users took more time to create their profiles, they reported that the sign up was short at the same rate as the system driven condition.
+  While these results are quite promising, to our knowledge no follow up work has been done to explore these issues further.
+  This goes to show, however, that the new user experience is a complicated issue and novel strategies may yet prove to be the best solution.
 
@@
     % point out that other approaches than smart object ordering exist, and we need to be able to compare.
-    % cite the comparison work at CHI.
+
   While past work has covered the new user rating survey as a solution to new user cold start, much less work has been done comparing other strategies.
+  Most of this work has been focused on strategies for how to get ratings from users.
+  Novel interfaces may elicit information other than ratings such as item comparisons \cite{Loepp}.
+  Furthermore, very little work has carefully considered how best to use the information we receive.
+  Most authors simply assert that more ratings are better, but for new users we may see interesting performance gains by simply choosing algorithms that perform well for new users.
 
-  % Temporal evaluation
-    % similar in-so-much as it can provide similar answers (growth over time), 
-    % different in-so-much as it tends to be used for just system growth, but it could be used for user growth.
-  Some researchers have taken to perfforming temporal evaluations of recommender systems, looking at accuracy metrics as a function of the lifetime of a system.
-
-  %TODO: restructure this, I can't have my cake and eat  too. This needs ot be refocused to explicitly focus on similarities and differences.
-  %One related approach to the one we take here is to evaluate recommender systems temporally.
-  %Various authors have taken this approach \cite{KorenTemporal, LathiaTemporal, burke2010evaluating} to evaluating recommender systems to understand how recommenders change with time.
-  %This work often features similar evaluations to ours, focusing on time (life of user or life of system) as opposed to number of ratings.
-  %One major difference between this approach and the one we take is that using ratings in the order that they were given can expose algorithms to biases from the system they were collected in.
-  %For example, in the MovieLens system users enter their first 15 ratings in a different interface, which does not give them choice over which items to rate.
-  %This is a forced situation which produces qualitatively different ratings than normal ratings.
-  %For this work we want to avoid this effect and therefor must use randomization.
-  %TODO: SOMEWHERE AROUND HERE THIS GOES!
-
-  % inspired by this work...
-
-  % Past work looks carefully at issues X, Y, and Z, but doesn’t take a systematic look at the behavior of algorithms for users with few ratings.
-  % Our work sits at the intersection of two important fields of recommender system research: New user experiance, and comparing algorithms.
-While there has been much work comparing various algorithms, there has been little formal work comparing algorithms based on their performance for new users.
-Therefore our work focuses on the behavior of recommender algorithms across different amounts of user data.
-(subsection might need re-ordering)
-
-  Our work sits at the intersection of two important traditions of recommender systems research: new user experience, and offline algorithm comparison.
-%%%%%%%%%%%%%%%%%%%%%%
+  We propose to carefully compare the performance of standard algorithms at recommending for new users.
   Our goal is to understand how different algorithms behave for new users, especially how their behavior changes as new users add ratings into the system.
@@
   Furthermore, we do not seek to find the algorithm that scores best on some metric of choice, instead, we seek to form a complete understanding of the differences between algorithms for cold start use.
-%%%%%%%%%%%%%%%%%%%%%%
+
+  Our approach is related to temporal evaluations of recommender systems \cite{KorenTemporal, LathiaTemporal, burke2010evaluating}.
+  Temporal evaluations tend to evaluate recommender systems as a function of time, seeing how they performed at different points in their lifetime.
+  This approach is well suited for understanding the effect of changes to the system, or user base of a recommender.
+  Unfortunately, this approach is less well suited towards comparing algorithms, as it necessarily co-mingles the effect of a recommender on user navigation with the quality of the recommender making it hard to consider algorithms that were not deployed with the technique.
+  Regardless, we consider this technique an inspiration for our evaluation.
+  %TODO: I kind of want to cut this paragraph
+  %TODO: Our work sits at the intersection of two important traditions of recommender systems research: new user experience, and offline algorithm comparison.
 
@@
   Prediction accuracy, as measured by RMSE or MAE is one of the classic techniques in recommended system evaluation.
-%%%%%%%%%%%%%%%%%%%%%%
   The idea behind these evaluations is to measure the accuracy of predictions on withheld recommendations.
@@
   Because users may build trust in the system by looking at the predictions of items, we are interested in knowing which algorithms are most accurate, and which show the most notable change in accuracy as ratings are entered.
-%%%%%%%%%%%%%%%%%%%%%%
 
   More recently, the research community has shifted away from accuracy metrics in favor of metrics that evaluate a recommender system's ability to rank or recommend.
-%%%%%%%%%%%%%%%%%%%%%%
-  % TODO: cite johnjoe here?
   These algorithms try to evaluate how useful the recommendations provided by the system are to the user.
   This is commonly done by looking either at how withheld items are ranked with metrics such as nDCG, or by looking at if good or bad withheld items tend to be recommended positively by the recommender.
-%%%%%%%%%%%%%%%%%%%%%%
 
   Prior work has shown that users are sensitive to properties of a recommender other than simply how well it performs.
-%%%%%%%%%%%%%%%%%%%%%%
-  These properties, such as the diversity of the output recommendations, can effect how the user interacts with the recommender and drastically change the user experience \cite{ziegler and martijn}.
+  These properties, such as the diversity of the output recommendations, can effect how the user interacts with the recommender and drastically change the user experience \cite{zieglerDiversity, martijnDiversity}.
   Unfortunately, there is no way to know all such interesting metrics
   We will look at three such metrics, the popularity, diversity, and spread of the recommendations.
-%%%%%%%%%%%%%%%%%%%%%%
 
   The popularity of the items recommended to a user can have an impact on the user's satisfaction.
-%%%%%%%%%%%%%%%%%%%%%%
   If the user is only recommended items that are very popular, then the recommender is likely failing to make novel and interesting recommendations.
@@
   This could cause the user to not trust the system, as they have no way of comparing the recommendations against their personal knowledge.
-%%%%%%%%%%%%%%%%%%%%%%
 
-  Past work \cite{ziegler and martijn} has shown that users tend to prefer recommendation lists that are more diverse.
-%%%%%%%%%%%%%%%%%%%%%%
-  <TODO: skim this past work again and report some details/results.>
-%%%%%%%%%%%%%%%%%%%%%%
+  Past work \cite{zieglerDiversity, martijnDiversity} has shown that users tend to prefer recommendation lists that are more diverse.
+  The belief is that by providing more diverse lists we are giving the user a larger range of items to choose from.
+  This can make choosing an item easier as the user is less likely to have to compare very similar options.
 
   Finally, we expect that algorithms with a good understand of its users will be able to recommend different items to different users.
-%%%%%%%%%%%%%%%%%%%%%%
   To measure this we will look at how well the recommender spreads its attention of many items.
@@
   In practice, a recommender may focus on both a shared core of popular or well reviewed items, and more personal items, which would lead to middle values.
-%%%%%%%%%%%%%%%%%%%%%%
 
@@
   \end{itemize}
-
-% RQ1A  How does algorithm behavior change with respect to predictive accuracy?
-      % RQ1B  How does algorithm behavior change with respect to TopN recommender performance?
-      % RQ1C How does algorithm behavior change with respect to other metrics such as popularity and diversity?
 
@@
   We seek to develop an understanding of how a range of different standard algorithms perform at for new users.
-%%%%%%%%%%%%%%%%%%%%%%
   Rather than picking advanced algorithms that are tuned for new user performance, we decided to first understand how representatives of three common types of algorithms perform.
@@
   Code for our evaluation, using the Lenskit evaluation framework \cite{lenskit} is available at \url{https://bitbucket.org/kluver/coldstartrecommendation}.
-%%%%%%%%%%%%%%%%%%%%%%
 
@@
       Algorithm          & Algorithm description \\\hline
-      User-User        & User based nearest neighbor collaborative filtering \cite{resnick1994grouplens}                           \\\hline
-      Item-Item        & Item based nearest neighbor collaborative filtering \cite{sarwar2001item}                                 \\\hline
-      SVD              & Simon Funk's SVD based collaborative filtering approach \cite{funk_netflix_2006}                          \\\hline
-      ItemBaseline     & Item's average rating, slight mean damping is applied                                                     \\\hline
-      UserItem\-Baseline & ItemBasline adjusted by the user's average offset from the ItemBaseline. slight mean damping is applied \\\hline
+      ItemBaseline       & Item's average rating, with mild bayesian damping towards the global mean \cite{funk_netflix_2006}. \\\hline
+      UserItem\-Baseline & ItemBasline adjusted by the user's average offset from the ItemBaseline. Mild mean damping is applied\cite{funk_netflix_2006}. \\\hline
+      User-User          & User based nearest neighbor collaborative filtering \cite{resnick1994grouplens} with a neighborhood size of 30 and ratings normalized by subtracting the UserItemBaseline score.
+\\\hline
+      Item-Item          & Item based nearest neighbor collaborative filtering \cite{sarwar2001item} with a neighborhood size of 30 and ratings normalized by subtracting the ItemBaseline score.
+\\\hline
+      SVD                & Simon Funk's SVD based collaborative filtering approach \cite{funk_netflix_2006} with 30 features and 150 training iterations per feature. \\\hline
     \end{tabular}
-    \caption{Summary of algorithms used in our research.}
+    \caption{Summary of algorithms.}
     \label{tbl:algo}
@@
 
-  %TODO: verify these claims.
-  %TODO: check for examples of how to write this
-  % algorithm is configured as such.
-  %TODO: topic sentence.
-
   % evaluation and code to run these algortihms available at.
   % citations and lenskit info
-  Each algorithm was configured and tuned similarly to Ekstrand's 2012 lenskit based evaluation \cite{ekstrand2012recommenders}.
-%%%%%%%%%%%%%%%%%%%%%%
-  Since this evaluation was on a slightly different dataset (MovieLens 10M), we performed only minor tuning from these parameters.
-  Most notably we found the tuning of the damping parameter used in the baseline recommenders to be too large for this dataset.
-  As the damping can have a large effect on baseline quality for new users, we adjusted to a more reasonable damping for this dataset.
-  Overall, we did not find our results to be highly sensitive to minor tuning changes, although specific values may change slightly.
-%%%%%%%%%%%%%%%%%%%%%%
+  Each algorithm was configured and tuned similarly to Ekstrand's 2012 evaluation \cite{ekstrand2012recommenders}.
+  Since this evaluation was on a slightly different dataset (MovieLens 10M), we performed minor tuning from these parameters.
+  Most notable, we found that the previous damping parameter used in the baselines was too large.
+  We found that a value of 5 gave much better results for new user recommendations.
+  During tuning we found that, among the configurations that performed well, our key results are consistent.
+  %TODO: carefully run this wording by joe.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%
-  The AILS@20 metric is based on the ILS metric introduced by ziegler et al \cite{ziegler}.
+  The AILS@20 metric is based on the ILS metric introduced by ziegler et al \cite{zieglerDiversity}.
   AILS@N is simply a rescaling of ILS to be scale free with regards to N, this was done to make the metric easier to interpret, by putting it on the same scale as similarity values.
+  We used the same item similarity metric for AILS@20 as we used in the Item-Item recommender.
 
@@
       % algos that use extra data about users or about items.
+    % cite the comparison work at CHI.
 

### 2014-05-05 paper/main.tex
improved introduction, I think.

@@
   One of the important problems in recommender systems is the new user experience.
-  Every potential user of the system must at some point be a new user.
-  If the system cannot provide a good user experience, then the user is likely to leave and not return.
+  If the system cannot provide a good user experience, then new users are likely to leave and not return.
   Even if the user does choose to enter the system, the user's first impression of the system can have a lasting impact on how the user interacts with the system.
@@
     % matching user's experiance is probably important.
-  One important part of how the user interacts with a recommender system is how much the user trusts the recommender system.
-  The more the user trusts a recommender system, the more the user can take advantage of it.
-  If the user doesn't trust a recommender system, the system can be of no value to the user.
-  Building trust with the user is therefore an important part of the new user experience.
-  Past work has shown that one of the strongest factors in the formation of trust is how much the system's recommendations match the user's knowledge \cite{wangAttributionOfTrust}.
-  For new users this can easily take the form of recommending items that the user has already seen.
+  %One important part of how the user interacts with a recommender system is how much the user trusts the recommender system.
+  %The more the user trusts a recommender system, the more the user can take advantage of it.
+  %If the user doesn't trust a recommender system, the system can be of no value to the user.
+  %Building trust with the user is therefore an important part of the new user experience.
+  %Past work has shown that one of the strongest factors in the formation of trust is how much the system's recommendations match the user's knowledge \cite{wangAttributionOfTrust}.
+  %For new users this can easily take the form of recommending items that the user has already seen.
   %TODO: I kinda of want to cut this. I want to talk about this theme, but I don't actually know where it fits in.
@@
     % we only look at user cold start as it is ubiquitous.
-  Past work on new user experience in recommender systems has tended to focus on the algorithmic cold start problem.
-  The cold start problem refers to the problem that no algorithm can make useful recommendations with very limited amounts of information.
-  While this applies to both new users and new items we will only discuss research on the new user cold start problem.
-  Most solutions to the algorithmic cold start problem require some source of extra information about new users.
+  Past work on new user experience in recommender systems has tended to focus on the `cold start problem'.
+  Cold start refers to the problem that no algorithm can make useful recommendations with very limited amounts of information, therefore new users and items must have worse recommendations.
+  We will only discuss the user side of the cold start problem in this work.
+  Most solutions to the cold start problem require some source of extra information about new users.
 
@@
   Systems that take this practice do not allow their users to receive recommendations until some fixed number of ratings have been entered.
-  The idea is that, after enough ratings have been entered, the predictions will be `good enough'.
+  The idea is that after enough ratings have been entered, the predictions will be `good enough'.
   This idea has been explored by many authors, most consider a system in which the user is presented with a list of items, and asked to rate items until they have enough ratings.
   For a good discussion of prior work in this direction see \cite{adaptiveBootstrapping}.
-  Generally, this work strives to strike a balance between the amount of effort the user must put in before getting enough ratings, and the usefulness of each rating to the user.
+  Generally, this work strives to strike a balance between the effort user needed to get enough ratings, and the usefulness of each rating to the system.
   This balance is complicated by how effort effects the user's perception of value \cite{TenIsEnough}.
@@
   This was pioneered by McNee et al., who looked at comparisons between user driven and system driven new user surveys \cite{mcneeInterfaces}.
-  They found that users who were in control of which items to rate during a new user survey rated fewer movies, but that the profiles they created were more useful to the system in predicting their future behavior.
-  Interestingly, while these users took more time to create their profiles, they reported that the sign up was short at the same rate as the system driven condition.
+  They found that users who were in control of the items they rated made fewer ratings, but that the profiles they created were more useful in predicting the user's future ratings.
+  Interestingly, while these users took more time to create their profiles, they were just as likely as other users to report that the sign up was short.
   While these results are quite promising, to our knowledge no follow up work has been done to explore these issues further.
@@
   Most of this work has been focused on strategies for how to get ratings from users.
-  Novel interfaces may elicit information other than ratings such as item comparisons \cite{Loepp}.
-  Furthermore, very little work has carefully considered how best to use the information we receive.
+  Very little work has carefully considered how best to use the information we receive.
   Most authors simply assert that more ratings are better, but for new users we may see interesting performance gains by simply choosing algorithms that perform well for new users.
 
-  We propose to carefully compare the performance of standard algorithms at recommending for new users.
-  Our goal is to understand how different algorithms behave for new users, especially how their behavior changes as new users add ratings into the system.
+  We propose to combine two traditions of recommender system research, offline algorithm analysis and new user experience research.
+  We seek to compare algorithms based on their performance for new users.
+  Our goal is to understand the differences between algorithms for new users, especially how these algorithms respond to different amounts of information.
   While there has been much work comparing algorithms in an offline setting, very little of it has explored what are the best algorithms for users with very few ratings.
-  Furthermore, we do not seek to find the algorithm that scores best on some metric of choice, instead, we seek to form a complete understanding of the differences between algorithms for cold start use.
+  %Furthermore, we do not seek to find the algorithm that scores best on some metric of choice, instead, we seek to form a complete understanding of the differences between algorithms for cold start use.
 
-  Our approach is related to temporal evaluations of recommender systems \cite{KorenTemporal, LathiaTemporal, burke2010evaluating}.
-  Temporal evaluations tend to evaluate recommender systems as a function of time, seeing how they performed at different points in their lifetime.
-  This approach is well suited for understanding the effect of changes to the system, or user base of a recommender.
-  Unfortunately, this approach is less well suited towards comparing algorithms, as it necessarily co-mingles the effect of a recommender on user navigation with the quality of the recommender making it hard to consider algorithms that were not deployed with the technique.
-  Regardless, we consider this technique an inspiration for our evaluation.
+  %Our approach is related to temporal evaluations of recommender systems \cite{KorenTemporal, LathiaTemporal, burke2010evaluating}.
+  %Temporal evaluations tend to evaluate recommender systems as a function of time, seeing how they performed at different points in their lifetime.
+  %This approach is well suited for understanding the effect of changes to the system, or user base of a recommender.
+  %Unfortunately, this approach is less well suited towards comparing algorithms, as it necessarily co-mingles the effect of a recommender on user navigation with the quality of the recommender making it hard to consider algorithms that were not deployed with the technique.
+  %Regardless, we consider this technique an inspiration for our evaluation.
   %TODO: I kind of want to cut this paragraph
-  %TODO: Our work sits at the intersection of two important traditions of recommender systems research: new user experience, and offline algorithm comparison.
+  %TODO: !!! Our work sits at the intersection of two important traditions of recommender systems research: new user experience, and offline algorithm comparison.
 
@@
   Unfortunately, algorithm behavior is a broad and hard to define subject.
-  We will look at three distinct ways to quantify the behavior of a recommender system, each of which may lead to multiple metrics.
+  We will look at three distinct sub questions to help capture recommender system behavior.
+  Each subquestion may represent several actual metrics in our evaluation.
+  %We will look at three distinct ways to quantify the behavior of a recommender system. each of which will lead to multiple metrics.
 
@@
 
-  Prediction accuracy, as measured by RMSE or MAE is one of the classic techniques in recommended system evaluation.
-  The idea behind these evaluations is to measure the accuracy of predictions on withheld recommendations.
-  If the algorithm can successfully predict the withheld ratings, then we expect it will accurately predict future ratings as well.
-  Because users may build trust in the system by looking at the predictions of items, we are interested in knowing which algorithms are most accurate, and which show the most notable change in accuracy as ratings are entered.
-
-  More recently, the research community has shifted away from accuracy metrics in favor of metrics that evaluate a recommender system's ability to rank or recommend.
-  These algorithms try to evaluate how useful the recommendations provided by the system are to the user.
-  This is commonly done by looking either at how withheld items are ranked with metrics such as nDCG, or by looking at if good or bad withheld items tend to be recommended positively by the recommender.
-
-  Prior work has shown that users are sensitive to properties of a recommender other than simply how well it performs.
-  These properties, such as the diversity of the output recommendations, can effect how the user interacts with the recommender and drastically change the user experience \cite{zieglerDiversity, martijnDiversity}.
-  Unfortunately, there is no way to know all such interesting metrics
-  We will look at three such metrics, the popularity, diversity, and spread of the recommendations.
-
-  The popularity of the items recommended to a user can have an impact on the user's satisfaction.
-  If the user is only recommended items that are very popular, then the recommender is likely failing to make novel and interesting recommendations.
-  In this case the user may feel that the recommender is not helpful in finding new and interesting things to watch.
-  However, if a new user is only recommended items that are not well known, it is unlikely that the user will have heard of the items recommended.
-  This could cause the user to not trust the system, as they have no way of comparing the recommendations against their personal knowledge.
-
-  Past work \cite{zieglerDiversity, martijnDiversity} has shown that users tend to prefer recommendation lists that are more diverse.
-  The belief is that by providing more diverse lists we are giving the user a larger range of items to choose from.
-  This can make choosing an item easier as the user is less likely to have to compare very similar options.
-
-  Finally, we expect that algorithms with a good understand of its users will be able to recommend different items to different users.
-  To measure this we will look at how well the recommender spreads its attention of many items.
-  We will do this by computing the Shannon's entropy of the items recommended to each user.
-  If the recommender system recommends the same items to each user, then the entropy will be low.
-  Conversely, if the recommender recommends different items to each user, then the entropy will high.
-  In practice, a recommender may focus on both a shared core of popular or well reviewed items, and more personal items, which would lead to middle values.
-
-  Based on this list of possible metrics, we will clarify our research questions with three sub questions:
   \begin{itemize}
   \item How do algorithms behave with respect to prediction accuracy?
-  \item How do algorithms behave with respect to ranking items appropriately?
+  \item How do algorithms behave with respect to ranking and recommending?
   \item How do algorithms behave as measured by other metrics such as popularity and diversity?
@@
 
-    % We look at three families of recommender algorithms (user based KNN, item based KNN, SVD)
-    % we compare on several dimensions
-    % we suggest a methodology for performing these types of computations
+  We will answer these questions by developing an evaluation framework that can be used to understand how algorithms perform for new users.
+  This framework will compare algorithms on eleven different metrics to allow a more complete understanding of the differences between algorithms.
+  We will then use this framework to study how three common algorithms behave for new users.
+  Finally, we will draw conclusions from our evaluation and outline directions for future work.
 
@@
   This issue will be addressed later in the results section.
+
+  Prediction accuracy, as measured by RMSE or MAE is one of the classic techniques in recommended system evaluation.
+  The idea behind these evaluations is to measure the accuracy of predictions on withheld recommendations.
+  If the algorithm can successfully predict the withheld ratings, then we expect it will accurately predict future ratings as well.
+  Because users may build trust in the system by looking at the predictions of items, we are interested in knowing which algorithms are most accurate, and which show the most notable change in accuracy as ratings are entered.
+
+  More recently, the research community has shifted away from accuracy metrics in favor of metrics that evaluate a recommender system's ability to rank or recommend.
+  These algorithms try to evaluate how useful the recommendations provided by the system are to the user.
+  This is commonly done by looking either at how withheld items are ranked with metrics such as nDCG, or by looking at if good or bad withheld items tend to be recommended positively by the recommender.
+
+  Prior work has shown that users are sensitive to properties of a recommender other than simply how well it performs.
+  These properties, such as the diversity of the output recommendations, can effect how the user interacts with the recommender and drastically change the user experience \cite{zieglerDiversity, martijnDiversity}.
+  Unfortunately, there is no way to know all such interesting metrics
+  We will look at three such metrics, the popularity, diversity, and spread of the recommendations.
+
+  The popularity of the items recommended to a user can have an impact on the user's satisfaction.
+  If the user is only recommended items that are very popular, then the recommender is likely failing to make novel and interesting recommendations.
+  In this case the user may feel that the recommender is not helpful in finding new and interesting things to watch.
+  However, if a new user is only recommended items that are not well known, it is unlikely that the user will have heard of the items recommended.
+  This could cause the user to not trust the system, as they have no way of comparing the recommendations against their personal knowledge.
+
+  Past work \cite{zieglerDiversity, martijnDiversity} has shown that users tend to prefer recommendation lists that are more diverse.
+  The belief is that by providing more diverse lists we are giving the user a larger range of items to choose from.
+  This can make choosing an item easier as the user is less likely to have to compare very similar options.
+
+  Finally, we expect that algorithms with a good understand of its users will be able to recommend different items to different users.
+  To measure this we will look at how well the recommender spreads its attention of many items.
+  We will do this by computing the Shannon's entropy of the items recommended to each user.
+  If the recommender system recommends the same items to each user, then the entropy will be low.
+  Conversely, if the recommender recommends different items to each user, then the entropy will high.
+  In practice, a recommender may focus on both a shared core of popular or well reviewed items, and more personal items, which would lead to middle values.
+
+  Based on this list of possible metrics, we will clarify our research questions with three sub questions:
 

### 2014-05-06 paper/main.tex
wrote short section on temporal eval, definitly more to that than I had given credit

@@
 
-  %Our approach is related to temporal evaluations of recommender systems \cite{KorenTemporal, LathiaTemporal, burke2010evaluating}.
-  %Temporal evaluations tend to evaluate recommender systems as a function of time, seeing how they performed at different points in their lifetime.
-  %This approach is well suited for understanding the effect of changes to the system, or user base of a recommender.
-  %Unfortunately, this approach is less well suited towards comparing algorithms, as it necessarily co-mingles the effect of a recommender on user navigation with the quality of the recommender making it hard to consider algorithms that were not deployed with the technique.
-  %Regardless, we consider this technique an inspiration for our evaluation.
-  %TODO: I kind of want to cut this paragraph
-  %TODO: !!! Our work sits at the intersection of two important traditions of recommender systems research: new user experience, and offline algorithm comparison.
-
-  We structure our work around the following research questions.
+  We structure our work around the following research question.
   \begin{itemize}
-    \item RQ1: How do algorithms behave for users with very few ratings?
-    \item RQ2: How does algorithm behavior change with increasing ratings for new users?
+    \item RQ1: How do algorithms behave for users with few ratings?
   \end{itemize}
@@
   We will look at three distinct sub questions to help capture recommender system behavior.
-  Each subquestion may represent several actual metrics in our evaluation.
+  Each sub question may represent several actual metrics in our evaluation.
   %We will look at three distinct ways to quantify the behavior of a recommender system. each of which will lead to multiple metrics.
@@
   \begin{itemize}
-  \item How do algorithms behave with respect to prediction accuracy?
-  \item How do algorithms behave with respect to ranking and recommending?
-  \item How do algorithms behave as measured by other metrics such as popularity and diversity?
+  \item RQ1.a: How do algorithms behave with respect to prediction accuracy?
+  \item RQ1.b: How do algorithms behave with respect to ranking and recommending?
+  \item RQ1.c: How do algorithms behave as measured by other metrics such as popularity and diversity?
   \end{itemize}
@@
   To answer our research questions we will evaluate algorithms in an offline simulation designed to simulate users as they join the system.
-%%%%%%%%%%%%%%%%%%%%%%
-  We will do this by running separate instances of the standard train and test methodology, each one designed so that test users have some small number of ratings.
+  We will do this by running separate instances of the standard train and test methodology, each one designed so that every test user has a specific  number of ratings.
   We will then plot metric results against the number of retained items to understand both how the algorithms compare on these metrics, and how the algorithm performance changes with the number of ratings provided.
-  This, or a similar technique, has been used by previous authors \cite{DrennerInitialExperiance, TenIsEnough, AdaptiveBootstrap}, however, these normally are used with only one algorithm or metric, and normally just to make the point that prediction accuracy improves as the number of ratings grows.
-  We will use this technique to develop a much more nuanced understanding of the differences between algorithms, and how they behave for users with few ratings.
-%%%%%%%%%%%%%%%%%%%%%%
+  This, or a similar technique, has been used by previous authors to make plots of metrics over profile size \cite{DrennerInitialExperiance, TenIsEnough, AdaptiveBootstrap}.
+  These authors did this simply to establish that recommender accuracy does improve as the number of ratings grows.
+  Therefore these authors did not carefully compare multiple algorithms on a breadth of metrics.
+  We will use this technique to develop a much more nuanced understanding of the differences between algorithms.
+
+  Our approach is related to a temporal evaluation, in which a recommender is trained on all ratings up to a point in time, and then tested using the next ratings to be made.
+  Temporal evaluations have been used in the past to evaluate how the list of recommendations from different algorithms change as users enter more ratings into the system \cite{LathiaTemporal}.
+  While the output of a temporal evaluation are very similar to our own, the methodology has some key differences.
+  Where a temporal evaluation is designed specifically to consider the order in which users make ratings, our methodology is designed to try to eliminate this ordering effect from our consideration of the algorithm's behavior.
+
+  As mentioned earlier, many recommender systems use a new user survey to collect the first user ratings.
+  These ratings are intentionally different than normal user ratings.
+  As we will discuss shortly, our methodology uses randomization in key places to remove this ordering effect.
+  This allows us to better capture the underlying algorithm performance on ``normal'' user data.
+  While this makes our methodology unsuited for temporal metrics, it allows us to separate an algorithm's behavior from how ratings are collected for new users.
+  Therefore, we feel that our methodology is more suitable to answer our research question.
 

### 2014-05-06 lenskit/src/eval/chart.R

@@ -72,6 +72,7 @@ dev.off()
 
 vars = c("SeenItems@20", "MeanRating@20", "RMSE@20")
 data.melt.sub = subset(data.melt, variable %in% vars)
+data.melt.sub = subset(data.melt.sub, Algorithm != "UserUser" | variable == "SeenItems@20")
 plot = ggplot(data.melt.sub, aes(x=Retain, y=value, color=Algorithm, shape=Algorithm, ymin=lb, ymax=ub))+geom_point()+geom_line()+geom_errorbar()+scale_x_continuous(breaks=4*(0:5))+facet_wrap(~variable, scales="free_y", nrow=1)
 pdf("rmse_20.pdf", width=8.5, height=3)
 print(plot)

### 2014-05-06 paper/main.tex
finished metrics subsection

@@
-%TODO: diversity config matches item-item config.
 %TODO: review results
@@
   Rather than picking advanced algorithms that are tuned for new user performance, we decided to first understand how representatives of three common types of algorithms perform.
-  Therefore we will compare User-User, Item-Item, and Funk's SVD algorithms against two simple baseline algorithms.
+  Therefore we will compare User-User, Item-Item, and Simon Funk's SVD algorithms against two simple baseline algorithms.
   More information about these algorithms can be found in table \ref{tbl:algo}
@@
       ItemBaseline       & Item's average rating, with mild bayesian damping towards the global mean \cite{funk_netflix_2006}. \\\hline
-      UserItem\-Baseline & ItemBasline adjusted by the user's average offset from the ItemBaseline. Mild mean damping is applied\cite{funk_netflix_2006}. \\\hline
+      UserItem\-Baseline & ItemBasline adjusted by the user's average offset from the ItemBaseline. Mild mean damping is applied \cite{funk_netflix_2006}. \\\hline
       User-User          & User based nearest neighbor collaborative filtering \cite{resnick1994grouplens} with a neighborhood size of 30 and ratings normalized by subtracting the UserItemBaseline score.
@@
   Each algorithm was configured and tuned similarly to Ekstrand's 2012 evaluation \cite{ekstrand2012recommenders}.
-  Since this evaluation was on a slightly different dataset (MovieLens 10M), we performed minor tuning from these parameters.
-  Most notable, we found that the previous damping parameter used in the baselines was too large.
-  We found that a value of 5 gave much better results for new user recommendations.
-  During tuning we found that, among the configurations that performed well, our key results are consistent.
+  Since this evaluation was on a different dataset (MovieLens 10M vs. MovieLens 1M), we performed minor tuning from these parameters.
+  Most notable, we found that the previous damping parameter used for the baselines was too large.
+  We found that a value of 5 gave much better results for new user recommendation.
+  During tuning we found that our key results are consistent among those configurations that perform well.
   %TODO: carefully run this wording by joe.
@@
   We will discuss a total of eleven metrics.
-%%%%%%%%%%%%%%%%%%%%%%
-  Some of these metrics will prove to be redundant, and future work should be able to use a much smaller list.
+  Some of these metrics will prove to be redundant, and future work should be able to use a much smaller list of metrics.
   The metrics are listed in table \ref{tbl:metrics}.
@@
 
-  All metrics evaluating recommendations (all but RMSE and nDCG) are based on the top 20 predicted items for the user, excluding items in the training set.
+  All metrics evaluating recommendations (all but RMSE and nDCG) are based on the top 20 highest predicted items for the user, excluding items in the training set.
   The choice of 20 items for the evaluations was arbitrary, testing with other values showed similar results, except on one metric.
@@
 
-  Prediction accuracy, as measured by RMSE or MAE is one of the classic techniques in recommended system evaluation.
-  The idea behind these evaluations is to measure the accuracy of predictions on withheld recommendations.
-  If the algorithm can successfully predict the withheld ratings, then we expect it will accurately predict future ratings as well.
-  Because users may build trust in the system by looking at the predictions of items, we are interested in knowing which algorithms are most accurate, and which show the most notable change in accuracy as ratings are entered.
+  Most of our metrics are standard techniques, so we will only address the latter 6.
+  As we will discuss later, we found Precision, MAP, and fallout to not be useful for this evaluation, therefore to help understand the quality of our recommendations, we include SeenItems@20, MeanRating@20, and RMSE@20.
+  SeemItems@20 tells us how frequently the user is to have previously seen a recommended item.
+  This is useful for two reasons.
+  First this provides important context for the next two metrics, as this is how many items those metrics are often averaged over.
+  Secondly, if too many items are known to the user, the user might feel that the recommendations are not interesting.
+  Alternatively, if too few of the items are known to the user, the user may have trouble evaluating the recommendations, which could lead the user to leave the system.
 
-  More recently, the research community has shifted away from accuracy metrics in favor of metrics that evaluate a recommender system's ability to rank or recommend.
-  These algorithms try to evaluate how useful the recommendations provided by the system are to the user.
-  This is commonly done by looking either at how withheld items are ranked with metrics such as nDCG, or by looking at if good or bad withheld items tend to be recommended positively by the recommender.
+  MeanRatings@20, and RMSE@20 both try to establish how much the user likes the recommendations.
+  MeanRatings@20 tells us directly how much the user tends to like recommended movies.
+  RMSE@20 accompanies this and lets us know if the recommended items are likely to be mispredicted, which can explain why users may or may not like their recommendations.
 
-  Prior work has shown that users are sensitive to properties of a recommender other than simply how well it performs.
-  These properties, such as the diversity of the output recommendations, can effect how the user interacts with the recommender and drastically change the user experience \cite{zieglerDiversity, martijnDiversity}.
-  Unfortunately, there is no way to know all such interesting metrics
-  We will look at three such metrics, the popularity, diversity, and spread of the recommendations.
+  The average popularity metric is a crude metric of how novel the recommendations are to the user \cite{zieglerDiversity}.
+  Broadly speaking we expect users will prefer recommendation lists containing more novel (less popular) items.
+  However, if the recommended items are too unpopular we risk that the user will not have heard of the items, which could make the user doubt the recommendations, hurting the user experiance.
 
-  The popularity of the items recommended to a user can have an impact on the user's satisfaction.
-  If the user is only recommended items that are very popular, then the recommender is likely failing to make novel and interesting recommendations.
-  In this case the user may feel that the recommender is not helpful in finding new and interesting things to watch.
-  However, if a new user is only recommended items that are not well known, it is unlikely that the user will have heard of the items recommended.
-  This could cause the user to not trust the system, as they have no way of comparing the recommendations against their personal knowledge.
-
-  Past work \cite{zieglerDiversity, martijnDiversity} has shown that users tend to prefer recommendation lists that are more diverse.
+  Past work has shown that users like their recommendations to be more diverse \cite{zieglerDiversity, martijnDiversity}.
   The belief is that by providing more diverse lists we are giving the user a larger range of items to choose from.
   This can make choosing an item easier as the user is less likely to have to compare very similar options.
+  We will measure diversity with the AILS@20 metric.
 
-  Finally, we expect that algorithms with a good understand of its users will be able to recommend different items to different users.
-  To measure this we will look at how well the recommender spreads its attention of many items.
-  We will do this by computing the Shannon's entropy of the items recommended to each user.
-  If the recommender system recommends the same items to each user, then the entropy will be low.
-  Conversely, if the recommender recommends different items to each user, then the entropy will high.
-  In practice, a recommender may focus on both a shared core of popular or well reviewed items, and more personal items, which would lead to middle values.
-
-  Based on this list of possible metrics, we will clarify our research questions with three sub questions:
-
-  % Provide computation details for AILS@N
-  Most of our metrics are standard techniques, or easily commutable, therefore we will only comment on the AILS, and spread metrics.
-%%%%%%%%%%%%%%%%%%%%%%%%%
   The AILS@20 metric is based on the ILS metric introduced by ziegler et al \cite{zieglerDiversity}.
@@
 
-  % Provide computation details for spread
-  The spread metric is computed by computing recommendations for each test user.
-%%%%%%%%%%%%%%%%%%%%%%%%%%  
+  The Spread@20 metric is designed to measure how well the recommender spreads its attention across many items.
+  We expect that algorithms with a good understanding of its users will be able to recommend different items to different users.
+  Spread is computed by computing recommendations for each test user.
   Let $C(I)$ be the count of how many times item $I$ showed up in the recommendations.
@@
   Finally we define spread as the Shannon's entropy of $P(I)$, $Spread = -\sum_I P(I) \log(P(I))$.
-%%%%%%%%%%%%%%%%%%%%%%%%%%
+  When this is large we know that the recommender is recommending many items with a relatively even distribution.
+  When this is low, the recommender is focusing on a small set of items which it recommends to every user.
 
@@
   We expect that the item baseline algorithm will have approximately constant performance on all metrics.
-  Since the Item Baseline is such a simple algorithm we expect that any reasonable algorithm should be able to do better on it on the accuracy on recommendation metrics.
+  Since the Item Baseline is such a simple algorithm we expect that any reasonable algorithm should be able to do better on it on the accuracy and recommendation metrics.
   Furthermore, we expect the improvement over the item baseline to be gradual, with larger improvements only happening with more data.
 
-  Many of the algorithms are configured with 
-  We expect that our intelligent algorithms, SVD, Item-Item, and User-User should be able to beat the item user baseline after
+  The user-item baseline should only differ from the item baseline for metrics that use the prediction value.
+  For all other metrics these two should provide the same responses.
+  As the user-item baseline is still quite a simple baseline, we expect that each algorithm will beat this baseline as well.
+  That said, for small rating values its possible that the improvements over this baseline will be very minimal.
+
+  We expect that the baseline algorithms will provide the most popular recommendations, with the least diversity, and the least spread.
+  We expect that any personalized algorithm should be able to make more nuanced recommendations than simply recommending the best rated items, which should translate as gains on these metrics.
 

### 2014-05-06 paper/main.tex
drafted methodology section

@@
     % establish range of evaluation.
-  For this evaluation we will use the MovieLens 1M dataset.
+  For this evaluation we will use the MovieLens 1M dataset \footnote{\url{http://grouplens.org/datasets/movielens/}}.
+  This dataset contains 1 million ratings from 6000 users on 4000 movies.
 
-  % movielens system
-      % silly cold start
-      % randomization
-  The MovieLens system uses a new user rating survey to allow users to enter the system.
+  These ratings were collected from the MovieLens system, which requires 15 ratings in a new user survey before users can enter the system.
+  To the best of our knowledge, at the time this data was collected the new user survey preferred items that were very popular.
+  Therefore, as we will discuss shortly, will randomize the order of the ratings to try to avoid measuring the effect of these initial ratings, and instead focus on the algorithm's performance on normal ratings.
+
+  Each user in the dataset has at least 20 ratings.
+  Therefore we will evaluate new users with up to 19 ratings.
+  This will allow us to ensure that each user in the dataset has at least one test rating.
 
@@
   The natural way to perform this evaluation is with a modified user crossfolding strategy.
+  The normal 5 fold cross validation strategy splits the users into five groups.
+  In each of five folds all ratings from four of the groups will be kept in the training set, and then either some constant number or percent of ratings from each user in the fifth group will be set aside for the test set.
+  The selection of a test set can be random to avoid any possible ordering effects in the data due to a new user rating strategy.
+  To understand how metrics change with a different number of users we can simply generate separate crossfolds for each number of ratings we wish to retain.
+  At each crossfold, rather than only taking a constant number of ratings for the test set, we can instead only keep a constant number of ratings for the training set.
 
@@
   Unfortunately, we found that this strategy can causes biases in our evaluation.
+  In particular, this strategy led to the size of the test sets to change as a function of the number of ratings retained.
+  As we kept more ratings for the training set, we would have fewer ratings in the testing set.
+  This turned out to bais our results, as some metrics, such as nDCG, proved to be sensitive to the number of items in the testing set.
 
@@
   Our solution to this bais is to subsample an existing crossfold.
+  Therefore, to generate training and test sets for different number of ratings from 1 to 19, we first generate a set of test and training splits with 19 training items using the crossfold strategy outlined above.
+  Then, to generate any other number of ratings, we randomly downsample the 19 item training set.
+  This allows us to generate training sets with a given number of ratings retained for test users, such that each number of ratings retained uses the same tests set.
 
@@
   % Prediction accuracy
-\begin{figure*}
+\begin{figure*}[ht!]
   \centering
-
   \includegraphics[width=\linewidth]{../lenskit/output/ekstrandTuned20/accuracy.pdf}
@@
 
-\begin{figure}[ht!]
+\begin{figure}[]
   \centering

### 2014-05-06 paper/main.tex
Results section re-writen, and conclusions rendered.

@@
       nDCG                   & Standard ranking quality evaluation.\\\hline
-      Precision@20           & Count of recommended items in the test set and rated 4.0 or higher.\\\hline
+      Precision@20           & Count of recommended items in the test set and rated 4.0 or higher\footnotemark[1].\\hline
       MAP@20                 & Average of precision at each rank one through twenty.\\\hline
-      Fallout@20             & Count of recommended items in the test set and rated 2.0 or lower.\\\hline
+      Fallout@20             & Count of recommended items in the test set and rated 2.0 or lower\footnotemark[1].\\\hline
       SeenItems@20           & Count of recommended items in the test set.\\\hline
@@
   Alternatively, if too few of the items are known to the user, the user may have trouble evaluating the recommendations, which could lead the user to leave the system.
+
+ % positioned here to get this on the right page.
+  \footnotetext[1]{We also tried 5.0 and 1.0 as cutoffs respectively, and found no meaningful difference.}\addtocounter{footnote}{1}
 
@@
   We expect that any personalized algorithm should be able to make more nuanced recommendations than simply recommending the best rated items, which should translate as gains on these metrics.
+  Furthermore, we expect that as the algorithm learns more about the user it should decrease in popularity, and increase in diversity, and spread, as it becomes able to make more personal and nuanced recommendations to the user.
 
@@
 
-  Using the methodology described in section \ref{sec:methodology} we generate plots for each metric showing how the five algorithms perform new users.
+  Using the methodology described in section \ref{sec:methodology} we generate plots for each metric showing how the five algorithms perform new users, versus how many of the user's ratings were retained in the training set.
   Where appropriate, we will compute metrics independently for each user, and plot the average user value, with a 95\% confidence interval around this mean.
@@
     % explain / interpret plot
+  \paragraph{RMSE}
   Figure \ref{fig:rmse} shows how accurate our algorithms are as measured by RMSE.
-  As would be expected the accuracy of our item baseline doesn't change as the amount of information changes.
-  The User-User and svd algorithms both perform as expected, starting out with the same accuracy as the item baseline, and then monotonically improving.
-  User-User does show a small increase of error with only one rating, but the change is quite small and unlikely to be noticed by users.
-  Interestingly, User-User and svd both perform about as well as our smarter user item baseline. %TODO: explain this better
+  User-User, and SVD both behaved essentially as expected.
+  SVD shows no improvement with only one rating, and User-User, actually appears to get worse with the first rating, but other than that, these algorithms outperform the item baseline.
+  Interestingly, the User-Item baseline outperforms both algorithms until around 8 ratings, at which point both algorithms start to perform better.
+  Given the size of of the differences between UserItemBaseline, UserUser, and SVD we will conclude that all three are at about the same accuracy for new users.
+  That such a simple algorithm can outperform for such small values goes to show that recommending for users with very few ratings is a hard problem.
 
-      % explain strangeness of Item-Item
-  Item-Item performs quite poorly in the cold start setting.
-  Until we have around 20 ratings the RMSE of Item-Item is worse than the item baseline.
-  More interestingly, unlike the other algorithms Item-Item actually appears to performs worse as it gets more ratings for the first few ratings.
-
-  The curve exhibited by Item-Item's RMSE is caused by two trends.
-  First as the system revives more ratings, predictions from Item-Item are more accurate, leading to a downward trend.
-  Secondly, as the system revives more ratings, Item-Item can make predictions for more items, meaning the fewer predictions are serviced by the baseline.
-  Because the baseline is more accurate, this leads to an upward trend.
-  Overall this leads to curved performance.
-  If we were to compute RMSE only over the items where item-item can make a prediction we would see a curve similar to the other algorithms, but shifted up.
-      % possibly coverage plot
-      % true crossing point?
+  Item-Item performs quite poorly for new users.
+  Until we have around 14 ratings, Item-Item is less accurate than the ItemBaseline.
+  In fact, for the first two ratings Item-Item appears to perform \emph{worse} as more ratings are added.
+  This trend is a result of the limited coverage of the Item-Item algorithm.
+  If you consider the accuracy of RMSE on only those items where it makes a personalized prediction it does show a monotonic increase in accuracy similar to the other algorithms.
+  However, for small rating counts ItemItem can only predict for some percent of the item (around 60\% for users with two ratings, and around 75\% for users with four ratings).
+  Because of this, many of the predictions come from the better performing UserItemBaseline.
+  As we get our first few ratings, the predictions become less likely to be from the UserItemBaseline, leading to an artificial upward trend in RMSE.
+  Fortunately, Item-Item seems to improve accuracy faster than the other algorithms, and therefore appears as if it will catch up.
+  Unfortunately, this trend suggests that ItemItem may be unsuitable for new users.
 
@@
     % NDCG
+  \paragraph{nDCG}
   Figure \ref{fig:ndcg} shows the algorithm's performance on the nDCG metric.
-  The algorithm's behavior is essentially equivalent to the RMSE behavior.
-  Again, User-User has a minor dip on very new users, but then tracks SVD in uniformly improving quite well from there.
-  Likewise Item-Item exhibits the same dipped accuracy, although it doesn't catch up the baseline as fast.
-  Unlike previous metrics, the item baseline, and the user-item baseline have the same performance.
-  This is because nDCG is a ranking metric, not an accuracy metric.
-  Because the user item baseline orders items the same way as the item baseline, both will have the same accuracy on any ranking metrics.
+  The algorithm's behavior is very similar to the RMSE behavior.
+  Both UserUser and SVD take an initial hit to accuracy at the first rating.
+  This time UserUser takes much longer to recover from this and only beats the baseline after around 12 ratings.
+  Again ItemItem performs quite poorly.
+  This leaves SVD to perform the best, passing the baseline after about 4 ratings.
+  The fact that the average item rating provides the best ranking for the first four ratings does suggest that ranking is a harder problem for recommenders to solve for new users than prediction accuracy.
 
@@
 \end{figure*}
-
+\paragraph{MAP@20, Precision@20, Fallout@20}
   Figure \ref{fig:map} shows MAP@20, precision@20, and fallout@20.
-  All three metrics show essentially the same trend, User-User gets the lowest score, the baselines get the highest score.
+  All three metrics show essentially the same trend, User-User gets the lowest score, the baseline get the highest score.
   On both MAP@20 and precision@20 svd gets a higher score than Item-Item throughout, performing almost as highly as the baseline.
-  On fallout@20 Item-Item performs higher than svd and the baseline, but decreases faster than bath, ending with a lower score than either. 
-  We also tried recall@20 and mrr@20 and found the same trend.
+  On fallout@20 Item-Item and SVD perform about as well as each other, with SVD showing a very slight lead.
+  We also tested recall and mean reciprocal rank metrics, but found them to exhibit the same trend.
 
   This result is rather strange as MAP and precision are metrics where high scores imply better performance, whereas fallout is a metric where high scores imply poor performance.
-  This leads to an issue, these metrics contradict each other in terms of the quality of the recommendations from our algorithm.
+  Therefore we would not expect to see rank equivalent behavior from these metrics.
+  Unfortunately this means that these metrics contradict each other in terms of which algorithm performs the best.
   One possible reason for this is that these topN metrics are known to be biased toward recommendations that have more popular items \cite{bellogin}.
   Looking at the average popularity (figure \ref{fig:pop}) of recommended items, we see that the average popularity of the recommendations closely replicates the previous three plots.
-  Therefore we seek a different, less biased metric to analyze the quality of the recommendations.
+  Therefore we will discard these metrics, and consider less biased metrics to help us analyze the quality of the recommendations.
 
@@
 \end{figure*}
-
+  \paragraph{SeenItems@20}
   The first graph in figure \ref{fig:topN.rmse} shows the average number of items in the top 20 recommendations for each algorithm that were also in the test set.
   The average number of test set items in recommendations shows a similar trend to the MAP, precision, and recall metrics, user-user getting by far the fewest seen items, and the baseline getting the most.
-  %TODO: talk about trust building here!
+  As User-User gets less than one seen item on average we found that MeanRating@20 and RMSE@20 metrics were not appropriate for UserUser.
+  This was observed in particular when testing our results for sensitivity to recommendation list size.
+  We found that different recommendation lists led UserUser (and only UserUser) to have very different results
+  Because of this issue we will not include User-User in our next evaluation and consider the next two metrics inconclusive for UserUser
+  That said, we consider the small number of seen items itself a bad result for UserUser.
+  Put simply, if we cannot reliably estimate if the recommendations are reasonable, how can we expect a user to do so?
 
+  \paragraph{MeanRating@20}
   The second graph in figure \ref{fig:topN.rmse} shows the average rating for those items in the top 20 recommendations for each algorithm.
   The first thing to note about this graph is that all of these algorithms are doing a reasonable job of recommending items the user might like to watch, with all algorithms averaging above 4 stars (implying that the recommended items are, at least on average, enjoyable).
-  SVD performs as expected, starting at close to the same average rating as the baselines, and then slowing improving as more ratings are entered.
-  Item-Item performs poorly, not only performing worse than the baseline, but showing worse performance as more ratings are added.
-  This is quite possibly a similar trend to the one we observed with RMSE and nDCG, where Item-Item's performance decreased as coverage increased leading to the curved performance at cold start.
+  SVD performs as expected, starting around the baseline for few ratings, and then steadily improving to be the best algorithm.
+  Again, we find that Item-Item performs poorly until quite a few ratings are added, and does worse than baseline throughout.
+  This is quite possibly a similar trend to the one we observed with RMSE and nDCG, where ItemItem's performance decreased as coverage increased leading to the curved performance at cold start.
 
-  During testing with other values of N for this metric and the next, we found that User-User's behavior compared to the other algorithms changed significantly with different values of N.
-  This is probably due to the very low rate of having seen items in the recommendations, which makes its value much less stable, than the other algorithms.
-  Because of this behavior we will draw no conclusion from User-User's performance on this metric, to avoid accidentally attributing behavior to it, that was instead formed by our choice of 20 as N.
-
+  \paragraph{RMSE@20}
   The third graph in figure \ref{fig:topN.rmse} shows the RMSE our algorithms got on items in the top 20 recommendations for each algorithm.
   Not surprisingly, this figure shows the same trends as the last one meaning that the algorithms whose recommendations were (on average) most likable, were those whose recommendations were most accurate.
-  Again, SVD performs quite well, Item-Item shows poor performance, and, due to sensitivity of User-User to our choice of N, we make no conclusions about User-User.
+  Again, SVD performs quite well, easily bearing the UserItemBaseline after only 1 rating.
+
+  Interestingly, we find that all algorithms are quite a bit more accurate on the top 20 items than they are for the general item collection.
+  This is likely an edge effect, where the highest rated (5 star) items are perhaps easier to predict than other items.
 
@@
 
+  \paragraph{AveragePopularity@20}
     % popularity graph
   The first graph in figure \ref{fig:pop} shows the average popularity of recommended items by each algorithm.
-  %TODO: re-write this
-  Across all datapoint sizes the ranking of algorithms is the same, with the baselines recommending the most popular movies, followed by svd, which recommends less popular movies, then Item-Item, which recommends much less popular movies, leaving User-User to recommend the least popular movies.
+  As expected the baseline algorithms make the most popular recommendations, followed by SVD, ItemItem, and then UserUser in a distant last.
   Unfortunately, we don't know of any guideline for understanding how popular users desire their recommendations to be.
-  That in mind, we expect that users will feel the baseline's popularity (around 1000) is too high, and User-User's popularity (around 50) is too low.
-  SVD and Item-Item both seem to perform well, with SVD generating more popular recommendations, than Item-Item, and both algorithms showing a significant trend to make less popular predictions as we know more about the user.
+  That said, we expect that users will feel the baseline's popularity (around 1000) is too high, and User-User's popularity (around 50) is too low.
+  SVD and Item-Item both seem to perform well, with SVD generating more popular recommendations than Item-Item.
+  SVD and ItemItem both show a trend of decreasing popularity as more ratings are retained, suggesting that these algorithms will make more nuanced, less popular recommendations for users they know more about.
+  Interestingly, these algorithms each appear to be converging to their own separate level of popularity, which suggests that this ordering may be independent of number of items retained.
 
     % diversity plot
-  The diversity graph in \ref{fig:pop} shows a similar trend to the popularity trend with the baseline algorithms producing the least diverse list of recommendations and User-User providing by far the most diverse list of recommendations.
-  For the first few ratings SVD is able to provide more diverse recommendations than Item-Item.
-  However, Item-Item seems to grow in diversity faster than SVD in response to more data, so after 4 ratings Item-Item is providing slightly more diverse recommendations than SVD.
-  Overall, Item-Item and SVD are providing recommendations that, at least in an offline context, appear to have quite comparable diversity.
+  \paragraph{AILS@20}
+  The second graph in \ref{fig:pop} shows a similar trend to the popularity trend with the baseline algorithms producing the least diverse list of recommendations and User-User providing by far the most diverse list of recommendations.
+  Unlike popularity, SVD and ItemItem seem to perform equivalently in diversity.
+  In terms of diversity we see that all three collaborative filtering algorithms show a trend of increasing diversity (decreasing  inter-list similarity) as more ratings are retained.
 
@@
   % spread plot
+  \paragraph{Spread@20}
   The graph in \ref{fig:spread} shows the spread of the recommendations from each algorithm.
-  Because this metric can only be computed over a group of people, confidence intervals over the per-user average are not provided, instead the specific spread values computed in each five folds are shown as points, with lines connecting the value from the five metrics.
-  %TODO: make this point earier and better.
-  Like the diversity and popularity metrics we see that the values take a much larger range, with different algorithms seeming to converge to different amounts spread.
-  The baseline algorithms, which recommend every user the same list of recommendations (unless the user has rated one of those items) have the lowest spread.
-  SVD has the second lowest spread, starting with a spread very close to the baselines and then slowly growing more spread as it gets more ratings.
-  Like popularity, we don't know exactly how much spread is best for a recommender system, but we can guess that we want more spread from our recommendations than baseline, and all other things being equal, more spread is probably better.
-  Based on this intuition it seems that SVD might have focus on too few items in its recommendations.
-
-  User-User and Item-Item both have significantly more spread than the other algorithms.
-  They show interestingly opposite trends, with User-User providing more spread at one rating, and then decreasing, and Item-Item giving less spread and then increasing.
-  By 8 ratings, we see that Item-Item has by far the most spread recommendations.
-  It is strange that User-User would get less spread as it learns more about users.
-  This implies that the more it learns about users, the less spread out its recommendations are.
+  Because this metric can only be computed over a group of people, confidence intervals over the per-user average are not provided, instead the specific spread values computed in each five folds are shown as points, with lines connecting the average values.
+  As expected the baseline performs worst on spread, as it recommends the same items to each user, unless the user has already rated one of these items.
+  SVD performs mildly better than the baseline, and shows definite improvement as more ratings are added.
+  UserUser and ItemItem start at around the same value as each other, however, after only two ratings they have crossed, as ItemItem quickly increases to have the highest spread of all the algorithms, and UserUser quickly decreases into second place.
+  Interestingly, only UserUser shows a trend to have less spread ratings as it learns more about the user, ItemItem and SVD both show the expected trend of increasing Spread as they learn more.
   This could be some form of a regression towards the mean, where as it learns more about users its recommendations become less personalized to the users individual quirks.
@@
 \section{Conclusions}
+  \begin{table}[ht!]
+    \centering
+    \begin{tabular}{|p{4.5em}|p{4.5em}|p{7.5em}|p{4.5em}|}
+      \hline
+      Algorithm & Perdiction accuracy & Recommendation quality      & Other properties \\\hline
+      ItemItem  & Poor                & Poor                        & Good             \\\hline
+      SVD       & Good                & Good                        & Good             \\\hline
+      UserUser  & Good                & Inconclusive                & Poor             \\\hline
+    \end{tabular}
+    \caption{Summary of algorithm quality for new users}
+    \label{tbl:results}
+  \end{table}
+
   % summary of results in a big old table.

### 2014-05-06 paper/main.tex
first draft I suppose

@@
-%TODO: review results
-%TODO: outline conclusions
-%TODO: write introduction
-%TODO: write methodology
-%TODO: re-write results
-%TODO: write conclusions
-%TODO: review extensively
 \documentclass[letterpaper]{sig-alternate}
@@
 
-\title{TODO: title}
+\title{Evaluating recommender behavior for new users}
 
@@
 
-TODO: abstract
+  The new user experiance is one of the important problems in recommender systems.
+  Past work on recommending for new users has focused on the process of gathering information from the user.
+  Our work focuses on how different algorithms behave for new users.
+  We describe a methodology that we use to compare representative of three common families of algorithms along eleven different metrics.
+  We find that overall the SVD based algorithm performs best for new users.
+  Our results can inform the design of interfaces and algorithms for new users, and suggest interesting opportunities for future work.
 
@@
     \end{tabular}
-    \caption{TODO:}
+    \caption{Summary of metrics}
     \label{tbl:metrics}
@@
   \centering
-  \includegraphics[width=\linewidth]{../lenskit/output/ekstrandTuned20/accuracy.pdf}
+  \includegraphics[width=0.9\linewidth]{../lenskit/output/ekstrandTuned20/accuracy.pdf}
   \caption{Left: RMSE, Right: nDCG}
@@
   \centering
-  \includegraphics[width=1.1\linewidth]{../lenskit/output/ekstrandTuned20/TopNPrecision.pdf}
-  \caption{TODO}
+  \includegraphics[width=1.05\linewidth]{../lenskit/output/ekstrandTuned20/TopNPrecision.pdf}
+  \caption{Left: Precision@20, Center: MAP@20, Right: Fallout@20}
   \label{fig:map}
@@
   \centering
-  \includegraphics[width=1.1\linewidth]{../lenskit/output/ekstrandTuned20/rmse_20.pdf}
-  \caption{TODO}
+  \includegraphics[width=1.05\linewidth]{../lenskit/output/ekstrandTuned20/rmse_20.pdf}
+  \caption{Left: SeenItems@20, Center: MeanRating@20, Right: RMSE@20}
   \label{fig:topN.rmse}
@@
   \centering
-  \includegraphics[width=\linewidth]{../lenskit/output/ekstrandTuned20/popdiv.pdf}
-  \caption{TODO}
+  \includegraphics[width=1.05\linewidth]{../lenskit/output/ekstrandTuned20/popdiv.pdf}
+  \caption{Left:averagePopularity@20, Center AILS@20, Right Spread@20}
   \label{fig:pop}
@@
 
-\begin{figure}[]
-  \centering
-  \includegraphics[width=1.1\columnwidth]{../lenskit/output/ekstrandTuned20/topN_entropy.pdf}
-  \caption{TODO}
-  \label{fig:spread}
-\end{figure}
-
   % spread plot
   \paragraph{Spread@20}
-  The graph in \ref{fig:spread} shows the spread of the recommendations from each algorithm.
+  The third graph in \ref{fig:spread} shows the spread of the recommendations from each algorithm.
   Because this metric can only be computed over a group of people, confidence intervals over the per-user average are not provided, instead the specific spread values computed in each five folds are shown as points, with lines connecting the average values.
@@
 \section{Conclusions}
+  % summary of results in a big old table.
+    % the accuracy of svd and user-user both seemed good. item-item seemed to have trouble especially with very new users.
+    % svd did quite well on our traditional topN metrics, item-item did OK, and user-user did bad., but there are flaws with these metrics.
+    % on our topN RMSE item-item (again) seemed to have issues, and user-user did quite well, but the low hit rate is troubleing. 
+    % again svd seems the winner here.
+    % on our fluffy metrics item-item and svd seemed to do well. I might argue that item-item seemed to do better (spread, and higher throughout).
+    % user-user seemed to do poor
+
+  % what this means
+    % user-user seems to be able to give accurate predictions, but seems to focus on overly obscure items in recommendation.
+    % item-item seems to give not-bad recommendations, but its predictions have issues
+    % based only on these results svd seems to be the best performer and is our recommendation to new systems.
+
   \begin{table}[ht!]
@@
 
-  % summary of results in a big old table.
-    % the accuracy of svd and user-user both seemed good. item-item seemed to have trouble especially with very new users.
-    % svd did quite well on our traditional topN metrics, item-item did OK, and user-user did bad., but there are flaws with these metrics.
-    % on our topN RMSE item-item (again) seemed to have issues, and user-user did quite well, but the low hit rate is troubleing. 
-    % again svd seems the winner here.
-    % on our fluffy metrics item-item and svd seemed to do well. I might argue that item-item seemed to do better (spread, and higher throughout).
-    % user-user seemed to do poor
-  % what this means
-    % user-user seems to be able to give accurate predictions, but seems to focus on overly obscure items in recommendation.
-    % item-item seems to give not-bad recommendations, but its predictions have issues
-    % based only on these results svd seems to be the best performer and is our recommendation to new systems.
+  The high level results of our evaluation are summarized in table \ref{tbl:results}
+  Overall we see that SVD seems to be the best performing algorithm for cold start use.
+  SVD was the only algorithm that consistently outperformed the baseline algorithms on prediction and recommendation tasks.
+  ItemItem had arguably better results on the average popularity and spread metrics, indicating that it was able to find more novel items, and make a larger range of recommendations to users.
+  Unfortunately, Item-Item also performed quite poorly at both prediction and recommendation, which may make it less suitable for new users.
+
+  UserUser performed quite well are prediction, but tended to favor unpopular items in its recommendations.
+  While some novelty is good, we believe that UserUser goes too far into the obscure movies in its recommendations, and therefore does not make good recommendations.
+  Therefore we believe that UserUser is not well suited to new users if its recommendations are going to be used.
+  Exploring corrections for this issue is left as future work.
+
   % other takeaways
     % regularization
+  This analysis suggests that, at least for new users, system designers use algorithms from the SVD family.
+  We belive that it is the regularization built into these algorithm that give them good performance.
+  UserUser and ItemItem algorithms have no built in method for damping their predictions towards the mean when they don't have much data.
+  The regularization in the cost function of the SVD model makes it so its predictions will gradually shift away from the baseline predictions as more relevant ratings are added.
+  The downside of this, however, could be the higher popularity and lower spread seen by the SVD algorithm.
+  The same regularization that helps avoid making mistakes due to too little data, may also prevents the algorithm from taking occasional risks on more obscure movies.
+
     % if you use a switching recommender, this methodology can help you choose when to switch.
     % non-traditional metrics can tell you more about what is going on, these seem to be more properties of the class of algo than of the data, so its worth looking at these when choosing algorithms, to find one with properties that are good for your task.
+  One interesting result of our analysis is that at very small numbers of ratings the UserItemBaseline seems to be our strongest algorithm, both in terms of accuracy (as measured by RMSE), and in terms of recommendation quality (as measured by MeanRating@20).
+  This suggests a recommendation strategy for live systems.
+  Instead of recommending from one recommender right away, a system could provide baseline predictions and recommendations until the user has enough ratings and then switching to a better recommender.
+  For recommenders like this our methodology provides a way to find the correct number of ratings at which to switch algorithms.
 
@@
       % no quitters in our dataset
-    % metrics limitations
-      % many metrics require user data to help us understand what users want from a recommender
-      % no well known ``goodness'' metric for topN.
+
+  One of the major limitations of our work is that we only evaluated one dataset.
+  Future work should replicate our experiment on other datasets from different recommender domains.
+  By doing this we will find out which of our results are true properties of the algorithms, and which results are instead related to properties of our dataset.
+
+  One important property that needs to be considered is that our dataset only contained users that naturally used the system enough to collect 20 ratings.
+  Unfortunately, this means that we do not know if there are differences between users who do and do not survive in the system.
+  This is a limitation of our methodology, to maintain a constant test set throughout the evaluation we are not able to evaluate users who have less ratings than the maximum amount we wish to study.
+  More work needs to be done to understand users who only ever make one or two ratings to understand why they leave, and how they differ from other users of our systems.
+
+  Finally, we feel that much more work should be done using this methodology to explore different algorithms.
+  There are many algorithms beyond the three e evaluated.
+  Some of these algorithms are designed specifically for cold start.
+  Others are designed to leverage extra, non-rating information to improve quality.
+  Work should be done to understand the differences between these algorithms, and what effect extra information has on the recommendation process.
+  Using our methodology we might be able to come up with an understanding of the value that extra information about the user has on recommendation, and to compare it the value of extra ratings for new users.
     % algorithm limitations

### 2014-05-06 lenskit/src/eval/chart.R

@@ -20,7 +20,7 @@ for (metric in metrics) {
 
 
 dat <- read.csv("eval-results.csv")
-plot = ggplot(dat, aes(x=Retain, y=TopN.Entropy, color=Algorithm, shape=Algorithm))+geom_point()+stat_summary(fun.y=mean, geom="line")+ylab("Spread")+scale_x_continuous(breaks=4*(0:5))
+plot = ggplot(dat, aes(x=Retain, y=TopN.Entropy, color=Algorithm, shape=Algorithm))+geom_point()+stat_summary(fun.y=mean, geom="line")+ylab("Spread")+scale_x_continuous(breaks=4*(0:5))+ theme(legend.position="none")
 pdf("topN_entropy.pdf", width=6, height=4)
 print(plot)
 dev.off()
@@ -69,7 +69,6 @@ pdf("TopNPrecision.pdf", width=8.5, height=3)
 print(plot)
 dev.off()
 
-
 vars = c("SeenItems@20", "MeanRating@20", "RMSE@20")
 data.melt.sub = subset(data.melt, variable %in% vars)
 data.melt.sub = subset(data.melt.sub, Algorithm != "UserUser" | variable == "SeenItems@20")
@@ -78,10 +77,34 @@ pdf("rmse_20.pdf", width=8.5, height=3)
 print(plot)
 dev.off()
 
+spread.dat <- read.csv("eval-results.csv")
+spread.dat = spread.dat[c("Algorithm","Retain", "TopN.Entropy", "Partition")]
+spread.dat$variable="Spread@20"
+spread.mean = aggregate(TopN.Entropy~Algorithm+Retain, data=spread.dat, mean)
+names(spread.mean) <- c("Algorithm", "Retain", "value")
+spread.mean$variable = "Spread@20"
+spread.mean$var = 0
+spread.mean$length = 0
+spread.mean$cisize = 0
+spread.mean$lb = 0
+spread.mean$ub = 0
 
 vars = c("AveragePopularity@20", "AILS@20")
 data.melt.sub = subset(data.melt, variable %in% vars)
-plot = ggplot(data.melt.sub, aes(x=Retain, y=value, color=Algorithm, shape=Algorithm, ymin=lb, ymax=ub))+geom_point()+geom_line()+geom_errorbar()+scale_x_continuous(breaks=4*(0:5))+facet_wrap(~variable, scales="free_y", nrow=1)
-pdf("popdiv.pdf", width=8.5, height=3.5)
+
[…]

### 2014-05-06 paper/main.tex
first stab at descriptors, abstract, etc.
cleaned up plots a little

@@
+%TODO: cleanup plots.
+
 \documentclass[letterpaper]{sig-alternate}
@@
 % A category with the (minimum) three required fields
-\category{H.4}{Information Systems Applications}{Miscellaneous}
-%A category including the fourth, optional field follows...
-\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]
+\category{H.3.3}{nformation Search and Retrieval}{Information filtering}
 
 %TODO: update these things.
-\terms{Theory}
+\terms{Algorithms, Measurement}
 
 %TODO: update these things.
-\keywords{ACM proceedings, \LaTeX, text tagging}
+\keywords{Recommender Systems, Evaluation, Profile Size, New User Experience, New User Problem, User Cold Start}
 
@@
   \centering
-  \includegraphics[width=0.9\linewidth]{../lenskit/output/ekstrandTuned20/accuracy.pdf}
+  \includegraphics[width=0.666\linewidth]{../lenskit/output/ekstrandTuned20/accuracy.pdf}
   \caption{Left: RMSE, Right: nDCG}
@@
   \centering
-  \includegraphics[width=1.05\linewidth]{../lenskit/output/ekstrandTuned20/TopNPrecision.pdf}
+  \includegraphics[width=\linewidth]{../lenskit/output/ekstrandTuned20/TopNPrecision.pdf}
   \caption{Left: Precision@20, Center: MAP@20, Right: Fallout@20}
@@
   \centering
-  \includegraphics[width=1.05\linewidth]{../lenskit/output/ekstrandTuned20/rmse_20.pdf}
+  \includegraphics[width=\linewidth]{../lenskit/output/ekstrandTuned20/rmse_20.pdf}
   \caption{Left: SeenItems@20, Center: MeanRating@20, Right: RMSE@20}
@@
   \centering
-  \includegraphics[width=1.05\linewidth]{../lenskit/output/ekstrandTuned20/popdiv.pdf}
+  \includegraphics[width=\linewidth]{../lenskit/output/ekstrandTuned20/popdiv.pdf}
   \caption{Left:averagePopularity@20, Center AILS@20, Right Spread@20}

### 2014-05-06 lenskit/src/eval/chart.R

@@ -1,45 +1,10 @@
 # Create a chart comparing the algorithms
 
 library("ggplot2")
-
-all.data <- read.csv("eval-user.csv")
-
-nonMetrics = c("Algorithm","Retain", "DataSet", "User", "Partition", "NGood", "TestEvents", "TrainEvents", "NAttempted", "RatingEntropy")
-
-metrics = setdiff(names(all.data),nonMetrics)
-
-for (metric in metrics) {
-  plot = ggplot(all.data, aes_string(x="Retain", y=metric, color="Algorithm", shape="Algorithm"))+stat_summary(fun.y=mean, geom="line")+stat_summary(fun.y=mean, geom="point") + stat_summary(fun.data = mean_cl_normal, geom="errorbar")+scale_x_continuous(breaks=4*(0:5))
-  metric = gsub('\\.', '_', metric)
-  filename = paste(metric,".pdf",sep="")
-  cat("Outputting to",filename,"\n")
-  pdf(filename, width=4, height=3)
-  print(plot)
-  dev.off()
-}
-
-
-dat <- read.csv("eval-results.csv")
-plot = ggplot(dat, aes(x=Retain, y=TopN.Entropy, color=Algorithm, shape=Algorithm))+geom_point()+stat_summary(fun.y=mean, geom="line")+ylab("Spread")+scale_x_continuous(breaks=4*(0:5))+ theme(legend.position="none")
-pdf("topN_entropy.pdf", width=6, height=4)
-print(plot)
-dev.off()
-
 library(reshape2)
 
 all.data.melt = melt(all.data, c("Algorithm", "Retain", "User"), metrics)
-melt.mean = aggregate(value~Algorithm+Retain+variable, data=all.data.melt, mean)
-melt.var = aggregate(value~Algorithm+Retain+variable, data=all.data.melt, var)
-names(melt.var)<- c("Algorithm","Retain","variable","var")
-melt.length = aggregate(value~Algorithm+Retain+variable, data=all.data.melt, length)
-names(melt.length)<- c("Algorithm","Retain","variable","length")
-data.melt = merge(melt.var, merge(melt.length, melt.mean))
-
-z = qnorm(1-0.05/2)
-
-data.melt$cisize = sqrt(data.melt$var/data.melt$length)*z
-data.melt$lb = data.melt$value-data.melt$cisize
[…]

### 2014-05-06 paper/main.tex
intro rewrite in response to joe's comments.

@@
-%TODO: cleanup plots.
+%TODO: update labels. retained ratings? profile size?
 
@@
   We describe a methodology that we use to compare representative of three common families of algorithms along eleven different metrics.
-  We find that overall the SVD based algorithm performs best for new users.
+  We find that making recommendations to users with little to no ratings is hard, with a baseline algorithm performing best for the first few ratings.
+  Once we have a few ratings, however, we find that overall the SVD based algorithm outperforms both user and item nearest neighbor approaches for new users.
   Our results can inform the design of interfaces and algorithms for new users, and suggest interesting opportunities for future work.
@@
   If the system cannot provide a good user experience, then new users are likely to leave and not return.
-  Even if the user does choose to enter the system, the user's first impression of the system can have a lasting impact on how the user interacts with the system.
-  The user's first impression can shape how the user expects the system to behave in future interactions, fundamentally changing how the user perceives the system.
+  Even if the recommender can make accurate recommendations, that is not enough to make a good first impression.
+  If the recommender only recommends items the user has seen, then the user may see no value in the system and leave.
+  Alternatively, if the recommender only recommends obscure items then the user will have no way to know if the evaluations are worth taking, and may leave.
+  This problem is made even worse by how little information we have about new users to inform personalization decisions.
 
@@
   Cold start refers to the problem that no algorithm can make useful recommendations with very limited amounts of information, therefore new users and items must have worse recommendations.
-  We will only discuss the user side of the cold start problem in this work.
+  We will only consider the user side of the cold start problem in this work.
   Most solutions to the cold start problem require some source of extra information about new users.
@@
 
-  While past work has covered the new user rating survey as a solution to new user cold start, much less work has been done comparing other strategies.
-  Most of this work has been focused on strategies for how to get ratings from users.
-  Very little work has carefully considered how best to use the information we receive.
-  Most authors simply assert that more ratings are better, but for new users we may see interesting performance gains by simply choosing algorithms that perform well for new users.
+  While past work has covered the new user rating survey from many different angles, it has not carefully evaluated which algorithm to use to recommend to these new users.
+  The new user survey work sidesteps this issue by assuming that the user will receive no personalization until they survey is complete.
+  For some recommenders deployed in the wild, however, providing no recommendations to users is not acceptable.
+  Even for those systems that do not personalize until the user has made enough ratings, carefully choosing algorithms for new users may be hugely important, and can help choose how many ratings is considered ``enough''.
 
-  We propose to combine two traditions of recommender system research, offline algorithm analysis and new user experience research.
-  We seek to compare algorithms based on their performance for new users.
+  Fortunately, there has been a growing trend in recommender system research to perform multiple metric evaluations of recommender systems.
+  This work comes out of a pushback against early recommendation evaluations which only evaluated recommenders in terms of their predictive accuracy \cite{accurateIsNotEnough}.
+  The response of this pushback has been the generating of many different metrics which allow us to more carefully understand the separate aspects of a recommender's output \cite{LathiaTemporal, CremonesiTopN, zieglerDiversity}.
+  %TODO: I want to cite more papers here, I don't think I have good ones either. Joe would probably know.
+  By evaluating how algorithms differ, not just in accuracy, but on dimensions such as the novelty of the recommendations, or how frequently the algorithm recommends new items, we can understand the tradeoffs between two otherwise comparable algorithms.
+  This will let us know which algorithm may be more suited for different domains, where different properties are more or less important.
+
+  We propose to combine two these two traditions of recommender system research, multiple metric algorithm analysis and new user experience research.
+  We seek to compare algorithms based on how the behave for new users.
   Our goal is to understand the differences between algorithms for new users, especially how these algorithms respond to different amounts of information.
-  While there has been much work comparing algorithms in an offline setting, very little of it has explored what are the best algorithms for users with very few ratings.
+  by using multiple metrics we can understand not just what algorithm is best in terms of absolute performance, but also gain a more nuanced understanding of the qualities of different algorithm's recommendations.
+  While there has been much work comparing algorithms in an offline setting, very little of it has been done in the context of new users.
   %Furthermore, we do not seek to find the algorithm that scores best on some metric of choice, instead, we seek to form a complete understanding of the differences between algorithms for cold start use.
@@
   \end{itemize}
-  Unfortunately, algorithm behavior is a broad and hard to define subject.
+  Unfortunately, algorithm behavior is a broad subject.
   We will look at three distinct sub questions to help capture recommender system behavior.
@@
   \begin{itemize}
-  \item RQ1.a: How do algorithms behave with respect to prediction accuracy?
-  \item RQ1.b: How do algorithms behave with respect to ranking and recommending?
-  \item RQ1.c: How do algorithms behave as measured by other metrics such as popularity and diversity?
+  \item RQ1.a: How accurately can different algorithms predict for new users?
+  \item RQ1.b: How well can different algorithms rank and recommend good items to new users?
+  \item RQ1.c: How do algorithms behave as measured by other metrics such as popularity and diversity for new users?
   \end{itemize}
@@
   These ratings are intentionally different than normal user ratings.
-  As we will discuss shortly, our methodology uses randomization in key places to remove this ordering effect.
+  As we will explain shortly, our methodology uses randomization in key places to remove this ordering effect.
   This allows us to better capture the underlying algorithm performance on ``normal'' user data.
@@
   Rather than picking advanced algorithms that are tuned for new user performance, we decided to first understand how representatives of three common types of algorithms perform.
-  Therefore we will compare User-User, Item-Item, and Simon Funk's SVD algorithms against two simple baseline algorithms.
+  Therefore we will compare User-User\cite{resnick1994grouplens}, Item-Item\cite{sarwar2001item}, and Simon Funk's SVD\cite{funk_netflix_2006} algorithms against two simple baseline algorithms.
   More information about these algorithms can be found in table \ref{tbl:algo}
@@
   \subsection*{Metrics}
-  We will discuss a total of eleven metrics.
-  Some of these metrics will prove to be redundant, and future work should be able to use a much smaller list of metrics.
+  We use a total of eleven metrics.
+  In our evaluation we found that some of these metrics are redundant, and future work should be able to use a much smaller list of metrics.
   The metrics are listed in table \ref{tbl:metrics}.
@@
   Most of our metrics are standard techniques, so we will only address the latter 6.
-  As we will discuss later, we found Precision, MAP, and fallout to not be useful for this evaluation, therefore to help understand the quality of our recommendations, we include SeenItems@20, MeanRating@20, and RMSE@20.
+  As we show later, we found Precision, MAP, and fallout to not be useful for this evaluation, therefore to help understand the quality of our recommendations, we include SeenItems@20, MeanRating@20, and RMSE@20.
   SeemItems@20 tells us how frequently the user is to have previously seen a recommended item.
@@
   To the best of our knowledge, at the time this data was collected the new user survey preferred items that were very popular.
-  Therefore, as we will discuss shortly, will randomize the order of the ratings to try to avoid measuring the effect of these initial ratings, and instead focus on the algorithm's performance on normal ratings.
+  Therefore, as we cover shortly, will randomize the order of the ratings to try to avoid measuring the effect of these initial ratings, and instead focus on the algorithm's performance on normal ratings.
 
@@
   \centering
-  \includegraphics[width=0.666\linewidth]{../lenskit/output/ekstrandTuned20/accuracy.pdf}
+  \includegraphics[width=0.75\linewidth]{../lenskit/output/ekstrandTuned20/accuracy.pdf}
   \caption{Left: RMSE, Right: nDCG}

### 2014-05-06 lenskit/src/eval/chart.R

@@ -1,7 +1,11 @@
 # Create a chart comparing the algorithms
 
+cp = c("#e41a1c","#377eb8","#4daf4a","#984ea3","#ff7f00")
+
 library("ggplot2")
 library(reshape2)
+
+all.data = read.csv("eval-user.csv")
 
 all.data.melt = melt(all.data, c("Algorithm", "Retain", "User"), metrics)
 data.melt = aggregate(value~Algorithm+Retain+variable, data=all.data.melt, mean)
@@ -18,20 +22,22 @@ data.melt$variable[data.melt$variable == "TopN.MeanPopularity"] = "AveragePopula
 data.melt$variable[data.melt$variable == "diversity"] = "AILS@20"
 data.melt$variable = ordered(data.melt$variable, levels = c("Coverage", "RMSE", "nDCG", "Precision@20", "MAP@20", "Fallout@20", "SeenItems@20", "MeanRating@20", "RMSE@20", "AveragePopularity@20", "AILS@20", "TestTime","MAE","MutualInformation","PredicitonEntropy","ItemScorer.NAttempted","ItemScorer.NGood","ItemScorer.Coverage","TopN.ActualLength", "Recall.fallout", "Recall"))
 
+data.melt$Algorithm = ordered(data.melt$Algorithm, c("ItemItem", "UserUser", "svd", "ItemBaseline", "UserItemBaseline"))
+
 # make plots
 vars = c("RMSE", "nDCG")
 data.melt.sub = subset(data.melt, variable %in% vars)
-data.melt.sub = subset(data.melt.sub, Algorithm != "UserUser" | variable == "RMSE")
-plot = ggplot(data.melt.sub, aes(x=Retain, y=value, color=Algorithm, shape=Algorithm))+geom_point()+geom_line()+scale_x_continuous(breaks=4*(0:5))+facet_wrap(~variable, scales="free_y", nrow=1)+theme_minimal()+theme(legend.position="bottom")
-pdf("accuracy.pdf", width=8.5*2/3, height=3.5)
+data.melt.sub = subset(data.melt.sub, Algorithm != "UserItemBaseline" | variable == "RMSE")
[…]

### 2014-05-06 paper/main.tex
rewrote some of the methodology section, and some of the metrics

@@
 
-  \begin{table}
+  \begin{table}[ht!]
     \centering
@@
   The metrics are listed in table \ref{tbl:metrics}.
-  \begin{table}[ht!]
+  \begin{table*}[ht!]
     \centering
-    \begin{tabular}{|p{7em}|p{16em}|}
+    \begin{tabular}{|p{7em}|p{20em}|p{4em}|p{18em}|}
       \hline
-      \multicolumn{2}{|c|}{{\bf Accuracy Metrics}} \\\hline
-      RMSE                   & Standard prediction accuracy metric. \\\hline
+      \multicolumn{4}{|c|}{{\bf Accuracy and Ranking Metrics}} \\\hline
+      RMSE                   & Standard prediction accuracy metric \cite{handbook}. & 0 to 5 & Smaller is better. \\\hline
+      nDCG                   & Standard ranking quality evaluation \cite{handbook}. & 0 to 1 & Larger is better. \\\hline
       \hline
-      \multicolumn{2}{|c|}{{\bf Recommendation Metrics}} \\\hline
-      nDCG                   & Standard ranking quality evaluation.\\\hline
-      Precision@20           & Count of recommended items in the test set and rated 4.0 or higher\footnotemark[1].\\hline
-      MAP@20                 & Average of precision at each rank one through twenty.\\\hline
-      Fallout@20             & Count of recommended items in the test set and rated 2.0 or lower\footnotemark[1].\\\hline
-      SeenItems@20           & Count of recommended items in the test set.\\\hline
-      MeanRating@20          & Average rating of recommended items that are in the test set.\\\hline
-      RMSE@20                & RMSE computed only over those items that were recommended.\\\hline
+      \multicolumn{4}{|c|}{{\bf Monotonic Recommendation Metrics}} \\\hline
+      Precision@20           & Count of recommended items in the test set and rated 4.0 or higher\footnotemark[1] \cite{handbook}. & 0 to 1 & Larger is better. \\\hline
+      MAP@20                 & Average of precision at each rank one through twenty \cite{manning2008introduction}. & 0 to 1 & Larger is better.\\\hline
+      Fallout@20             & Count of recommended items in the test set and rated 2.0 or lower\footnotemark[1] \cite{handbook}. & 0 to 1 & Smaller is better. \\\hline
+      MeanRating@20          & Average rating of recommended items that are in the test set.& 0 to 5 & Larger is better \\\hline
+      RMSE@20                & RMSE computed only over those items that were recommended.& 0 to 5 & Smaller is better \\\hline
       \hline
-      \multicolumn{2}{|c|}{{\bf Other Metrics}} \\\hline
-      Average\-Popularity@20 & The average number of users in the training set that have seen recommended items.\\\hline
-      AILS@20                & The average pairwise similarity between recommended items. \\\hline
-      Spread@20              & The Shannon's entropy of the distribution of recommended items for users in the test set.\\\hline
+      \multicolumn{4}{|c|}{{\bf Nonmonotonic Recommender Metrics}} \\\hline
+      SeenItems@20           & Count of recommended items in the test set. & 0 to 20 & Not well understood. \newline Larger is \emph{probably} better within reason. \\\hline
+      Average\-Popularity@20 & The average number of users in the training set that have seen recommended items \cite{zieglerDiversity}.& 1 to 6000 & Not well understood.
+\newline Smaller is \emph{probably} better within reason.
+\\\hline
+      AILS@20                & The average pairwise similarity between recommended items \cite{zieglerDiversity}.
+& 0 to 1 & Not well understood.
+\newline  Smaller is \emph{probably} better within reason.
+\\\hline
+      Spread@20              & The Shannon's entropy of the distribution of recommended items for users in the test set.
+& 0 to 12 & Not well understood.
+\newline Larger is \emph{probably} better within reason.
+\\\hline
     \end{tabular}
@@
     \label{tbl:metrics}
-  \end{table}
+  \end{table*}
+  \footnotetext[1]{We also tried 5.0 and 1.0 as cutoffs respectively, and found no meaningful difference.}\addtocounter{footnote}{1}
 
-  All metrics evaluating recommendations (all but RMSE and nDCG) are based on the top 20 highest predicted items for the user, excluding items in the training set.
+  We split our metrics into three groups roughly along our three sub research questions.
+
+  \paragraph{Accuracy and Ranking Metrics}
+  The first two metrics evaluate the recommender based on its predictions for each item.
+  These evaluate, the accuracy of the predictions, and the ability of the recommender to rank the list of all items.
+  We find that these two metrics perform quite similarly in our evaluation.
+
+  \paragraph{Monotonic Recommendation Metrics}
+  These metrics all evaluate the recommendations produced by the recommender.
+  For recommendations we will use the 20 highest predicted items for each user, excluding items in the training set.
+  As the UserItem-Baseline and ItemBaseline make the same recommendations for all users, we will only report results for the ItemBaseline here.
   The choice of 20 items for the evaluations was arbitrary, testing with other values showed similar results, except on one metric.
@@
 
-  Most of our metrics are standard techniques, so we will only address the latter 6.
-  As we show later, we found Precision, MAP, and fallout to not be useful for this evaluation, therefore to help understand the quality of our recommendations, we include SeenItems@20, MeanRating@20, and RMSE@20.
-  SeemItems@20 tells us how frequently the user is to have previously seen a recommended item.
-  This is useful for two reasons.
-  First this provides important context for the next two metrics, as this is how many items those metrics are often averaged over.
-  Secondly, if too many items are known to the user, the user might feel that the recommendations are not interesting.
-  Alternatively, if too few of the items are known to the user, the user may have trouble evaluating the recommendations, which could lead the user to leave the system.
+  There metrics are useful because these metrics have a well defined monotonic relation to the satisfaction of the user.
+  As the metric grows larger, or smaller (depending on metric), we expect the output recommendations to satisfy users more.
+  This makes these metrics easier to understand than our later metrics.
 
- % positioned here to get this on the right page.
-  \footnotetext[1]{We also tried 5.0 and 1.0 as cutoffs respectively, and found no meaningful difference.}\addtocounter{footnote}{1}
+  Our first three metrics (Precision, MAP, and Fallout) are standard information retrieval metrics.
+  As we show later, we found Precision, MAP, and fallout to not be useful for this evaluation.
+  Therefore, to help understand the quality of our recommendations, we also include MeanRating@20, and RMSE@20.
 
@@
   RMSE@20 accompanies this and lets us know if the recommended items are likely to be mispredicted, which can explain why users may or may not like their recommendations.
+  As it helps us interpret these two metrics we will also discuss SeenItems@20 with these metrics, although SeenItems@20 does not have as simple a relation to user satisfaction.
+
+  \paragraph{Nonmonotonic Recommendation Metrics}
+  Our final four metrics are all nonmonotonic measures of the user's recommendations.
+  These metrics are harder to understand than the monotonic measures of recommendation quality.
+  Unlike the previous metrics, we expect that both of the extremes of this metric would be perceived negatively by the user.
+
+  We use these metrics as they can help us understand important properties of how users may perceive the recommendations.
+  For example, as mentioned in the introduction how many of the recommended items the user has seen (estimated by SeenItems@20) can have a major impact on how much users trust the recommendations.
+  If an algorithm recommends too few items that the user has seen then we expect the user will have trouble evaluating if the recommendations are worth taking.
+  That said, however, if too many items in the recommendations have already been seen, the user may feel that the recommendations are not interesting or useful.
+  Unfortunately, without more research on these metrics to establish a mapping between user satisfaction and these values we will have to make educated guesses at what values for these metrics are too extreme.
 
@@
   Broadly speaking we expect users will prefer recommendation lists containing more novel (less popular) items.
-  However, if the recommended items are too unpopular we risk that the user will not have heard of the items, which could make the user doubt the recommendations, hurting the user experiance.
+  However, if the recommended items are too unpopular we risk that the user will not have heard of the items, which could make the user doubt the recommendations, hurting the user experience.
 
@@
   This can make choosing an item easier as the user is less likely to have to compare very similar options.
+  It has also been shown, however, that too much diversity can lead to less user satisfaction \cite{zieglerDiversity}.
+  This is probably due to a tradeoff between how diverse a list is and how well it captures a user's tastes.
+  At some point, a list cannot become more diverse without include items that the user is not interested in.
   We will measure diversity with the AILS@20 metric.
@@
   We expect that algorithms with a good understanding of its users will be able to recommend different items to different users.
+  Like diversity, however, we expect that it is not possible to reach complete spread (recommending each item an equal number of times) without making avoidably bad recommendations.
+  When Spread is large we know that the recommender is recommending many items with a relatively even distribution, this might be expected from a random recommender.
+  When Spread is small, the recommender is focusing on a small set of items which it recommends to every user, this might be expected from a baseline recommender.
+
   Spread is computed by computing recommendations for each test user.
@@
   Finally we define spread as the Shannon's entropy of $P(I)$, $Spread = -\sum_I P(I) \log(P(I))$.
-  When this is large we know that the recommender is recommending many items with a relatively even distribution.
-  When this is low, the recommender is focusing on a small set of items which it recommends to every user.
 
+  \paragraph{Expectations about algorithm performance}
   % provide expectations.
   We expect that the item baseline algorithm will have approximately constant performance on all metrics.
-  Since the Item Baseline is such a simple algorithm we expect that any reasonable algorithm should be able to do better on it on the accuracy and recommendation metrics.
+  Since the Item Baseline is such a simple algorithm we expect that any reasonable algorithm should be able to do better on it the monotonic metrics.
   Furthermore, we expect the improvement over the item baseline to be gradual, with larger improvements only happening with more data.
@@
   For all other metrics these two should provide the same responses.
-  As the user-item baseline is still quite a simple baseline, we expect that each algorithm will beat this baseline as well.
-  That said, for small rating values its possible that the improvements over this baseline will be very minimal.
+  As the user-item baseline is still quite a simple baseline, we expect that the non baseline algorithm will beat this baseline as well.
+  That said, for small rating values its possible that the improvements over this baseline will be very small.
 
@@
   Furthermore, we expect that as the algorithm learns more about the user it should decrease in popularity, and increase in diversity, and spread, as it becomes able to make more personal and nuanced recommendations to the user.
-
   \subsection*{Dataset}
@@
   Using the methodology described in section \ref{sec:methodology} we generate plots for each metric showing how the five algorithms perform new users, versus how many of the user's ratings were retained in the training set.
-  Where appropriate, we will compute metrics independently for each user, and plot the average user value, with a 95\% confidence interval around this mean.
-  Where this cannot be done we will compute the metric once for each of the five folds, and plot the values for each fold, with a lines through the average.
 
-\subsection{Prediction accuracy}
+\subsection{Accuracy and Ranking Metrics}
   % Prediction accuracy
@@
   \includegraphics[width=0.75\linewidth]{../lenskit/output/ekstrandTuned20/accuracy.pdf}
-  \caption{Left: RMSE, Right: nDCG}
+  \caption{Accuracy and Ranking Algorithms}
   \label{fig:rmse}
@@
   SVD shows no improvement with only one rating, and User-User, actually appears to get worse with the first rating, but other than that, these algorithms outperform the item baseline.
-  Interestingly, the User-Item baseline outperforms both algorithms until around 8 ratings, at which point both algorithms start to perform better.
-  Given the size of of the differences between UserItemBaseline, UserUser, and SVD we will conclude that all three are at about the same accuracy for new users.
-  That such a simple algorithm can outperform for such small values goes to show that recommending for users with very few ratings is a hard problem.
+  Interestingly, the UserItem-Baseline outperforms both algorithms until around 8 ratings, at which point both algorithms start to perform better.
+  Given the size of the differences between UserItem-Baseline, UserUser, and SVD we conclude that all three provide about the same accuracy for new users.
+  That such a simple algorithm can outperform more complicated for such small values goes to show that recommending for users with very few ratings is a hard problem.
 
   Item-Item performs quite poorly for new users.
-  Until we have around 14 ratings, Item-Item is less accurate than the ItemBaseline.
+  Until we have 13 ratings, Item-Item is less accurate than the ItemBaseline.
   In fact, for the first two ratings Item-Item appears to perform \emph{worse} as more ratings are added.
@@
   However, for small rating counts ItemItem can only predict for some percent of the item (around 60\% for users with two ratings, and around 75\% for users with four ratings).
-  Because of this, many of the predictions come from the better performing UserItemBaseline.
-  As we get our first few ratings, the predictions become less likely to be from the UserItemBaseline, leading to an artificial upward trend in RMSE.
+  Because of this, many of the predictions come from the better performing UserItem-Baseline.
+  As we get our first few ratings, the predictions become less likely to be from the UserItem-Baseline, leading to an artificial upward trend in RMSE.
   Fortunately, Item-Item seems to improve accuracy faster than the other algorithms, and therefore appears as if it will catch up.
-  Unfortunately, this trend suggests that ItemItem may be unsuitable for new users.
+  Unfortunately, this result suggests that ItemItem may be unsuitable for new users.
 
-\subsection {Recommendatin quality metrics}
   % Recommendation quality metrics
@@
   Figure \ref{fig:ndcg} shows the algorithm's performance on the nDCG metric.
-  The algorithm's behavior is very similar to the RMSE behavior.
+  The results for nDCG are very similar to the RMSE results.
   Both UserUser and SVD take an initial hit to accuracy at the first rating.
@@
   Again ItemItem performs quite poorly.
-  This leaves SVD to perform the best, passing the baseline after about 4 ratings.
-  The fact that the average item rating provides the best ranking for the first four ratings does suggest that ranking is a harder problem for recommenders to solve for new users than prediction accuracy.
+  This leaves SVD to perform the best, passing the baseline after about 3 ratings.
+  Again we find that our best algorithm for the first few ratings is our baseline.
 
@@
 
+\subsection {Monotonic Recommendation Quality Metrics}
+
 \begin{figure*}[ht!]
@@
   \includegraphics[width=\linewidth]{../lenskit/output/ekstrandTuned20/TopNPrecision.pdf}
-  \caption{Left: Precision@20, Center: MAP@20, Right: Fallout@20}
+  \caption{Information Retrieval Recommendation Quality Metrics}
   \label{fig:map}
@@
   On fallout@20 Item-Item and SVD perform about as well as each other, with SVD showing a very slight lead.
-  We also tested recall and mean reciprocal rank metrics, but found them to exhibit the same trend.
+  We also tested recall and mean reciprocal rank metrics, but found them to exhibit the same trend, so they were not explored further.
 
-  This result is rather strange as MAP and precision are metrics where high scores imply better performance, whereas fallout is a metric where high scores imply poor performance.
+  This result is rather strange as MAP and Precision are metrics where larger scores imply better performance, whereas fallout is a metric where larger scores imply poor performance.
   Therefore we would not expect to see rank equivalent behavior from these metrics.
@@
   One possible reason for this is that these topN metrics are known to be biased toward recommendations that have more popular items \cite{bellogin}.
-  Looking at the average popularity (figure \ref{fig:pop}) of recommended items, we see that the average popularity of the recommendations closely replicates the previous three plots.
+  Looking forward to the average popularity (figure \ref{fig:pop}) of recommended items, we see that the average popularity of the recommendations closely replicates the previous three plots.
   Therefore we will discard these metrics, and consider less biased metrics to help us analyze the quality of the recommendations.
@@
   \includegraphics[width=\linewidth]{../lenskit/output/ekstrandTuned20/rmse_20.pdf}
-  \caption{Left: SeenItems@20, Center: MeanRating@20, Right: RMSE@20}
+  \caption{Recommendation Quality Metrics}
   \label{fig:topN.rmse}
 \end{figure*}
+
   \paragraph{SeenItems@20}
   The first graph in figure \ref{fig:topN.rmse} shows the average number of items in the top 20 recommendations for each algorithm that were also in the test set.
-  The average number of test set items in recommendations shows a similar trend to the MAP, precision, and recall metrics, user-user getting by far the fewest seen items, and the baseline getting the most.
+  We address this metric out of turn here as it provides important context for the next two metrics.
+  The average number of test set items in recommendations shows a similar trend to the MAP, precision, and recall metrics, with user-user getting by far the fewest seen items, and the baseline getting the most.
   As User-User gets less than one seen item on average we found that MeanRating@20 and RMSE@20 metrics were not appropriate for UserUser.
   This was observed in particular when testing our results for sensitivity to recommendation list size.
-  We found that different recommendation lists led UserUser (and only UserUser) to have very different results
-  Because of this issue we will not include User-User in our next evaluation and consider the next two metrics inconclusive for UserUser
+  We found that different recommendation lists lead UserUser (and only UserUser) to have very different results.
+  Because of this issue we will not include User-User in our next evaluations and consider the next two metrics inconclusive for UserUser.
   That said, we consider the small number of seen items itself a bad result for UserUser.
@@
   \includegraphics[width=\linewidth]{../lenskit/output/ekstrandTuned20/popdiv.pdf}
-  \caption{Left:averagePopularity@20, Center AILS@20, Right Spread@20}
+  \caption{Nonmonotonic Recommendation Quality Metrics}
   \label{fig:pop}

### 2014-05-06 paper/main.tex
finished second pass, now I need to edit and trim

@@
 
-%\usepackage[factor=250,spacing=true]{microtype}
+\usepackage[factor=250,spacing=true]{microtype}
 
@@
   The first thing to note about this graph is that all of these algorithms are doing a reasonable job of recommending items the user might like to watch, with all algorithms averaging above 4 stars (implying that the recommended items are, at least on average, enjoyable).
-  SVD performs as expected, starting around the baseline for few ratings, and then steadily improving to be the best algorithm.
+  SVD performs as expected, starting around the baseline for few ratings, and then steadily improving to be the best algorithm after 3 ratings.
   Again, we find that Item-Item performs poorly until quite a few ratings are added, and does worse than baseline throughout.
   This is quite possibly a similar trend to the one we observed with RMSE and nDCG, where ItemItem's performance decreased as coverage increased leading to the curved performance at cold start.
+  Regardless, we find that ItemItem's recommendations do not seem to be as well rated as either the baseline or SVD.
 
@@
   The third graph in figure \ref{fig:topN.rmse} shows the RMSE our algorithms got on items in the top 20 recommendations for each algorithm.
-  Not surprisingly, this figure shows the same trends as the last one meaning that the algorithms whose recommendations were (on average) most likable, were those whose recommendations were most accurate.
-  Again, SVD performs quite well, easily bearing the UserItemBaseline after only 1 rating.
+  Not surprisingly, this figure shows the same trends as the last one, meaning that the algorithms whose recommendations were (on average) most likable, were those whose recommendations were most accurate.
+  Again, SVD performs quite well, easily beating the UserItemBaseline after only 1 rating.
 
@@
 
-\subsection{Other recommender metrics}
+\subsection{Nonmonotonic Recommender Quality Metrics}
   % Other recommender metrics
@@
   The first graph in figure \ref{fig:pop} shows the average popularity of recommended items by each algorithm.
-  As expected the baseline algorithms make the most popular recommendations, followed by SVD, ItemItem, and then UserUser in a distant last.
-  Unfortunately, we don't know of any guideline for understanding how popular users desire their recommendations to be.
-  That said, we expect that users will feel the baseline's popularity (around 1000) is too high, and User-User's popularity (around 50) is too low.
+  As expected the baseline algorithms makes the most popular recommendations, followed by SVD, ItemItem, and then UserUser in a distant last.
+  As stated earlier, We expect that users will feel the baseline's popularity (around 1000) is too high.
+  We also expect that users will perceive User-User's popularity (around 50) as too low.
   SVD and Item-Item both seem to perform well, with SVD generating more popular recommendations than Item-Item.
   SVD and ItemItem both show a trend of decreasing popularity as more ratings are retained, suggesting that these algorithms will make more nuanced, less popular recommendations for users they know more about.
-  Interestingly, these algorithms each appear to be converging to their own separate level of popularity, which suggests that this ordering may be independent of number of items retained.
+  Interestingly, these algorithms each appear to be converging to their own separate level of popularity, which suggests that this ordering may be independent of the number of ratings.
 
@@
   \paragraph{AILS@20}
-  The second graph in \ref{fig:pop} shows a similar trend to the popularity trend with the baseline algorithms producing the least diverse list of recommendations and User-User providing by far the most diverse list of recommendations.
-  Unlike popularity, SVD and ItemItem seem to perform equivalently in diversity.
-  In terms of diversity we see that all three collaborative filtering algorithms show a trend of increasing diversity (decreasing  inter-list similarity) as more ratings are retained.
+  The second graph in \ref{fig:pop} shows a similar trend to the popularity trend.
+  Again, we expect that the baseline, which provides the least diverse recommendations, would be precised as not very diverse by users.
+  Quite possibly due to the preference for obscure movies, UserUser seems to produce the most diverse lists of any algorithm.
+  We expect that the UserUser recommendation lists might be perceived as too diverse.
+  Unlike popularity, SVD and ItemItem seem to perform equivalently in terms of diversity.
+  We also see that all three collaborative filtering algorithms show a trend of increasing diversity (decreasing  inter-list similarity) as more ratings are retained.
 
@@
   The third graph in \ref{fig:spread} shows the spread of the recommendations from each algorithm.
-  Because this metric can only be computed over a group of people, confidence intervals over the per-user average are not provided, instead the specific spread values computed in each five folds are shown as points, with lines connecting the average values.
-  As expected the baseline performs worst on spread, as it recommends the same items to each user, unless the user has already rated one of these items.
+  As expected the baseline performs worst on spread, as it recommends the same items to each user (unless the user has already rated one of these items).
   SVD performs mildly better than the baseline, and shows definite improvement as more ratings are added.
@@
 
-\section{Conclusions}
+\section{Discussion}
   % summary of results in a big old table.
@@
       \hline
-      Algorithm & Perdiction accuracy & Recommendation quality      & Other properties \\\hline
+      Algorithm & Prediction accuracy & Recommendation quality      & Other properties \\\hline
       ItemItem  & Poor                & Poor                        & Good             \\\hline
+      UserUser  & Good                & Inconclusive                & Poor             \\\hline
       SVD       & Good                & Good                        & Good             \\\hline
-      UserUser  & Good                & Inconclusive                & Poor             \\\hline
     \end{tabular}
@@
   While some novelty is good, we believe that UserUser goes too far into the obscure movies in its recommendations, and therefore does not make good recommendations.
-  Therefore we believe that UserUser is not well suited to new users if its recommendations are going to be used.
+  Therefore we believe that UserUser is not well suited for new users unless only its predictions are used.
   Exploring corrections for this issue is left as future work.
+
+    % if you use a switching recommender, this methodology can help you choose when to switch.
+    % non-traditional metrics can tell you more about what is going on, these seem to be more properties of the class of algo than of the data, so its worth looking at these when choosing algorithms, to find one with properties that are good for your task.
+  Interestingly, however, we see that for very small numbers of ratings the UserItemBaseline seems to be our strongest algorithm, both in terms of accuracy (as measured by RMSE), and recommendation quality (as measured by MeanRating@20).
+  This suggests a switching recommendation strategy for live systems.
+  Instead of recommending from one recommender right away, a system could provide baseline predictions and recommendations until the user has enough ratings and then switching to a better recommender.
+  This has the added avantage that it might help users notice the changes in popularity and diversity, helping them appreciate these properties more.
+  For switching recommendation strategies like this, our methodology provides a way to find the correct number of ratings at which to switch algorithms.
 
@@
     % regularization
-  This analysis suggests that, at least for new users, system designers use algorithms from the SVD family.
-  We belive that it is the regularization built into these algorithm that give them good performance.
+  One possible reason that the SVD algorithm performs as well as it does is the regularization built into the algorithm.
   UserUser and ItemItem algorithms have no built in method for damping their predictions towards the mean when they don't have much data.
@@
   The same regularization that helps avoid making mistakes due to too little data, may also prevents the algorithm from taking occasional risks on more obscure movies.
-
-    % if you use a switching recommender, this methodology can help you choose when to switch.
-    % non-traditional metrics can tell you more about what is going on, these seem to be more properties of the class of algo than of the data, so its worth looking at these when choosing algorithms, to find one with properties that are good for your task.
-  One interesting result of our analysis is that at very small numbers of ratings the UserItemBaseline seems to be our strongest algorithm, both in terms of accuracy (as measured by RMSE), and in terms of recommendation quality (as measured by MeanRating@20).
-  This suggests a recommendation strategy for live systems.
-  Instead of recommending from one recommender right away, a system could provide baseline predictions and recommendations until the user has enough ratings and then switching to a better recommender.
-  For recommenders like this our methodology provides a way to find the correct number of ratings at which to switch algorithms.
 
@@
 
+\section{Conclusion}
+
+  Our work shows that a multiple metric approach to understanding recommender behavior can provide useful insights into the differences between algorithms.
+  Our methodology presents a list of metrics that can be reused in the future for new multiple metric evaluations, and our methodology shows how this can be applied to understand new users.
+  We find that recommending for new users is hard and none of our algorithms can reliably beat a simple baseline for users with very few ratings.
+  Once the user has a small number of ratings we find that the SVD algorithm performs best.
+  This result can be useful both in designing new recommender systems, and in designing new recommendation algorithms.
+
 \section{Acknowledgements}
- TODO:
+  The authors would like to express their gratitude towards our colleagues at GroupLens research for their feedback and support while developing this work.
+  We would like to specially that Michael Ekstrand for his  important work on the Lenskit recommendation framework.
+
+  This work is supported by NSF grant TODO:
 

### 2014-05-07 paper/main.tex
introduction revisions and cuts. changed graph labels. some revision in section 2

@@
-%TODO: update labels. retained ratings? profile size?
+%TODO: update language retain to simulated profile size. do a search ``reta''
 
@@
 
-\title{Evaluating recommender behavior for new users}
+\title{Evaluating Recommender Behavior For New Users}
 
@@
 
-  The new user experiance is one of the important problems in recommender systems.
+  The new user experience is one of the important problems in recommender systems.
   Past work on recommending for new users has focused on the process of gathering information from the user.
@@
   We describe a methodology that we use to compare representative of three common families of algorithms along eleven different metrics.
-  We find that making recommendations to users with little to no ratings is hard, with a baseline algorithm performing best for the first few ratings.
-  Once we have a few ratings, however, we find that overall the SVD based algorithm outperforms both user and item nearest neighbor approaches for new users.
-  Our results can inform the design of interfaces and algorithms for new users, and suggest interesting opportunities for future work.
+  We find for first few ratings a baseline algorithm performs better than three common collaborative filtering algorithms.
+  Once we have a few ratings, we find that Funk's SVD algorithm has the best overall performance.
+  Our results can inform the design of interfaces and algorithms for new users.
 
@@
   If the system cannot provide a good user experience, then new users are likely to leave and not return.
-  Even if the recommender can make accurate recommendations, that is not enough to make a good first impression.
-  If the recommender only recommends items the user has seen, then the user may see no value in the system and leave.
-  Alternatively, if the recommender only recommends obscure items then the user will have no way to know if the evaluations are worth taking, and may leave.
-  This problem is made even worse by how little information we have about new users to inform personalization decisions.
+  Even if the system can make accurate recommendations accuracy alone is not enough to make a good first impression.
+  If the system only recommends items the user has seen, then the user may see no value in the system and leave.
+  Alternatively, if the system only recommends obscure items, then the user will have no way to know if the recommendations are worth taking.
+  This problem is made more complex by how little information we have about new users.
 
@@
   Past work on new user experience in recommender systems has tended to focus on the `cold start problem'.
-  Cold start refers to the problem that no algorithm can make useful recommendations with very limited amounts of information, therefore new users and items must have worse recommendations.
-  We will only consider the user side of the cold start problem in this work.
-  Most solutions to the cold start problem require some source of extra information about new users.
+  Cold start refers to the problem that no algorithm can make highly personal recommendations with very limited amounts of information.
+  Unfortunately this means that regardless of the algorithm, no recommender system can make high quality recommendations for users who are still joining the system.
 
@@
   % (framework from decision tree work)
-  The best studied strategy for new users is to have users go through a new user rating survey.
+
+  The natural way to improve the system for new users is to have access to some extra information.
+  The best studied strategy is for new users to go through a new user rating survey.
   Systems that take this practice do not allow their users to receive recommendations until some fixed number of ratings have been entered.
@@
   For a good discussion of prior work in this direction see \cite{adaptiveBootstrapping}.
-  Generally, this work strives to strike a balance between the effort user needed to get enough ratings, and the usefulness of each rating to the system.
-  This balance is complicated by how effort effects the user's perception of value \cite{TenIsEnough}.
 
@@
   % promising, we don't know much about follow up.
-  One promising approach to the new user rating survey is to have users control which movies they rate.
-  This was pioneered by McNee et al., who looked at comparisons between user driven and system driven new user surveys \cite{mcneeInterfaces}.
-  They found that users who were in control of the items they rated made fewer ratings, but that the profiles they created were more useful in predicting the user's future ratings.
-  Interestingly, while these users took more time to create their profiles, they were just as likely as other users to report that the sign up was short.
-  While these results are quite promising, to our knowledge no follow up work has been done to explore these issues further.
-  This goes to show, however, that the new user experience is a complicated issue and novel strategies may yet prove to be the best solution.
+  %One promising approach to the new user rating survey is to have users control which movies they rate.
+  %This was pioneered by McNee et al., who looked at comparisons between user driven and system driven new user surveys \cite{mcneeInterfaces}.
+  %They found that users who were in control of the items they rated made fewer ratings, but that the profiles they created were more useful in predicting the user's future ratings.
+  %Interestingly, while these users took more time to create their profiles, they were just as likely as other users to report that the sign up was short.
+  %While these results are quite promising, to our knowledge no follow up work has been done to explore these issues further.
+  %This goes to show, however, that the new user experience is a complicated issue and novel strategies may yet prove to be the best solution.
 
@@
 
-  While past work has covered the new user rating survey from many different angles, it has not carefully evaluated which algorithm to use to recommend to these new users.
-  The new user survey work sidesteps this issue by assuming that the user will receive no personalization until they survey is complete.
-  For some recommenders deployed in the wild, however, providing no recommendations to users is not acceptable.
-  Even for those systems that do not personalize until the user has made enough ratings, carefully choosing algorithms for new users may be hugely important, and can help choose how many ratings is considered ``enough''.
+  While past work has covered the new user rating survey from many different angles, it has not carefully evaluated which algorithm is best at recommending to new users.
+  The new user survey work sidesteps this issue by assuming that the user will receive no personalization until the survey is complete.
+  Even for systems with a new user survey carefully choosing algorithms for new users may be very important.
+  Users fresh out of the new user survey will still have far fewer ratings than the average user of the recommender system.
+  Different algorithms may provide very different experiences for these users, even if they give very similar results for the average user.
 
@@
   %TODO: I want to cite more papers here, I don't think I have good ones either. Joe would probably know.
-  By evaluating how algorithms differ, not just in accuracy, but on dimensions such as the novelty of the recommendations, or how frequently the algorithm recommends new items, we can understand the tradeoffs between two otherwise comparable algorithms.
-  This will let us know which algorithm may be more suited for different domains, where different properties are more or less important.
+  This strategy allows us to dig much deeper into the differences between algorithms, looking past simple measures of accuracy to more meaningful dimensions such as how novel the recommendations are, or how frequently the recommendations change for any user.
+  This allows us to capture the qualitative tradeoffs between algorithms which are otherwise quite similar, which helps us choose algorithms that are better suited for a particular domain.
 
-  We propose to combine two these two traditions of recommender system research, multiple metric algorithm analysis and new user experience research.
-  We seek to compare algorithms based on how the behave for new users.
+  We combine these two types of research, new user experience research, and multiple metric algorithm evaluation.
   Our goal is to understand the differences between algorithms for new users, especially how these algorithms respond to different amounts of information.
   by using multiple metrics we can understand not just what algorithm is best in terms of absolute performance, but also gain a more nuanced understanding of the qualities of different algorithm's recommendations.
-  While there has been much work comparing algorithms in an offline setting, very little of it has been done in the context of new users.
   %Furthermore, we do not seek to find the algorithm that scores best on some metric of choice, instead, we seek to form a complete understanding of the differences between algorithms for cold start use.
@@
   \end{itemize}
-  Unfortunately, algorithm behavior is a broad subject.
-  We will look at three distinct sub questions to help capture recommender system behavior.
-  Each sub question may represent several actual metrics in our evaluation.
+  Algorithm behavior is a very broad subject.
+  We will look at three distinct sub questions to help capture algorithm behavior.
   %We will look at three distinct ways to quantify the behavior of a recommender system. each of which will lead to multiple metrics.
@@
   \begin{itemize}
-  \item RQ1.a: How accurately can different algorithms predict for new users?
+  \item RQ1.a: How well can different algorithms predict the future ratings of new users?
   \item RQ1.b: How well can different algorithms rank and recommend good items to new users?
@@
   To answer our research questions we will evaluate algorithms in an offline simulation designed to simulate users as they join the system.
-  We will do this by running separate instances of the standard train and test methodology, each one designed so that every test user has a specific  number of ratings.
-  We will then plot metric results against the number of retained items to understand both how the algorithms compare on these metrics, and how the algorithm performance changes with the number of ratings provided.
-  This, or a similar technique, has been used by previous authors to make plots of metrics over profile size \cite{DrennerInitialExperiance, TenIsEnough, AdaptiveBootstrap}.
-  These authors did this simply to establish that recommender accuracy does improve as the number of ratings grows.
-  Therefore these authors did not carefully compare multiple algorithms on a breadth of metrics.
-  We will use this technique to develop a much more nuanced understanding of the differences between algorithms.
+  We do this by adopting the standard train and test methodology to simulate users with a given number of ratings.
+  We run this methodology for various profile sizes, and then plot the results.
+  This lets us understand both how the algorithms compare on these metrics, and how the algorithm performance changes with the number of ratings provided.
 
-  Our approach is related to a temporal evaluation, in which a recommender is trained on all ratings up to a point in time, and then tested using the next ratings to be made.
-  Temporal evaluations have been used in the past to evaluate how the list of recommendations from different algorithms change as users enter more ratings into the system \cite{LathiaTemporal}.
-  While the output of a temporal evaluation are very similar to our own, the methodology has some key differences.
+  This, or a similar technique, has been used by previous authors to make plots of metric values over profile size \cite{DrennerInitialExperiance, TenIsEnough, AdaptiveBootstrap, LathiaTemporal}.
+  Past work used these plots to establish that more ratings lead to better results.
+  We take this technique farther and develop a much more nuanced understanding of the different algorithms, and how they compare for different profile sizes.
+
+  Our approach is related to the temporal evaluation methodology, in which a recommender is repeatedly trained on all ratings up to a point in time, and then tested using the next ratings to be made for each user.
+  Temporal evaluations have been used in the past to evaluate how the list of recommended items change as users enter more ratings into the system \cite{LathiaTemporal}.
+  While the output of a temporal evaluation can be very similar to our own, the methodologies have some key differences.
   Where a temporal evaluation is designed specifically to consider the order in which users make ratings, our methodology is designed to try to eliminate this ordering effect from our consideration of the algorithm's behavior.
@@
   As mentioned earlier, many recommender systems use a new user survey to collect the first user ratings.
-  These ratings are intentionally different than normal user ratings.
-  As we will explain shortly, our methodology uses randomization in key places to remove this ordering effect.
+  These ratings are different than normal user ratings.
+  As we explain shortly, our methodology uses randomization in key places to remove this ordering effect.
   This allows us to better capture the underlying algorithm performance on ``normal'' user data.
-  While this makes our methodology unsuited for temporal metrics, it allows us to separate an algorithm's behavior from how ratings are collected for new users.
+  While this makes our methodology unsuited for applying temporal metrics, it allows us to separate an algorithm's behavior from how ratings are collected for new users.
   Therefore, we feel that our methodology is more suitable to answer our research question.
@@
   \subsection*{Metrics}
+
   We use a total of eleven metrics.
@@
   Furthermore, we expect that as the algorithm learns more about the user it should decrease in popularity, and increase in diversity, and spread, as it becomes able to make more personal and nuanced recommendations to the user.
+
   \subsection*{Dataset}
@@
     % establish range of evaluation.
+
   For this evaluation we will use the MovieLens 1M dataset \footnote{\url{http://grouplens.org/datasets/movielens/}}.
@@
   \subsection*{Evaluation Procedure}
+
   % subsampling strategy
@@
 \subsection{Accuracy and Ranking Metrics}
+
   % Prediction accuracy
@@
 \section{Acknowledgements}
+
   The authors would like to express their gratitude towards our colleagues at GroupLens research for their feedback and support while developing this work.

### 2014-05-07 lenskit/src/eval/chart.R

@@ -28,7 +28,7 @@ data.melt$Algorithm = ordered(data.melt$Algorithm, c("ItemItem", "UserUser", "sv
 vars = c("RMSE", "nDCG")
 data.melt.sub = subset(data.melt, variable %in% vars)
 data.melt.sub = subset(data.melt.sub, Algorithm != "UserItemBaseline" | variable == "RMSE")
-plot = ggplot(data.melt.sub, aes(x=Retain, y=value, color=Algorithm, shape=Algorithm))+geom_point()+geom_line()+scale_x_continuous(breaks=4*(0:5))+facet_wrap(~variable, scales="free_y", nrow=1)+theme(legend.position="bottom")+scale_color_manual(values=cp)+theme_minimal()
+plot = ggplot(data.melt.sub, aes(x=Retain, y=value, color=Algorithm, shape=Algorithm))+geom_point()+geom_line()+scale_x_continuous(breaks=4*(0:5))+facet_wrap(~variable, scales="free_y", nrow=1)+scale_color_manual(values=cp)+theme_minimal()+xlab("Simulated Profile Size")+theme(legend.position="bottom")
 pdf("accuracy.pdf", width=8.5*2/2.5, height=3.5)
 print(plot)
 dev.off()
@@ -37,7 +37,7 @@ dev.off()
 vars = c("Precision@20", "MAP@20", "Fallout@20")
 data.melt.sub = subset(data.melt, variable %in% vars)
 data.melt.sub = subset(data.melt.sub, Algorithm != "UserItemBaseline")
-plot = ggplot(data.melt.sub, aes(x=Retain, y=value, color=Algorithm, shape=Algorithm))+geom_point()+geom_line()+scale_x_continuous(breaks=4*(0:5))+facet_wrap(~variable, scales="free_y", nrow=1)+theme(legend.position="bottom")+scale_color_manual(values=cp)+theme_minimal()
+plot = ggplot(data.melt.sub, aes(x=Retain, y=value, color=Algorithm, shape=Algorithm))+geom_point()+geom_line()+scale_x_continuous(breaks=4*(0:5))+facet_wrap(~variable, scales="free_y", nrow=1)+scale_color_manual(values=cp)+theme_minimal()+xlab("Simulated Profile Size")+theme(legend.position="bottom")
 pdf("TopNPrecision.pdf", width=8.5, height=3.5)
 print(plot)
 dev.off()
[…]

### 2014-05-07 paper/main.tex
edited the methodology section

@@
 
-  % new user experiance is only chance to make all important first impression
+  % new user experience is only chance to make all important first impression
   One of the important problems in recommender systems is the new user experience.
@@
     % trust is complex
-    % matching user's experiance is probably important.
+    % matching user's experience is probably important.
   %One important part of how the user interacts with a recommender system is how much the user trusts the recommender system.
@@
   % past work tends to focus on the algorithm side, the algorithm side of this problem is cold start.
-    % SMAL DATA TENDS TO IMPLY BAD RESULTS.
+    % Small DATA TENDS TO IMPLY BAD RESULTS.
     % cold start algorithms is a larger problem
-    % cold start algorithms tend to use extra information to help imrpove eary performance.
+    % cold start algorithms tend to use extra information to help improve Early performance.
     % we only look at user cold start as it is ubiquitous.
@@
   Therefore we will compare User-User\cite{resnick1994grouplens}, Item-Item\cite{sarwar2001item}, and Simon Funk's SVD\cite{funk_netflix_2006} algorithms against two simple baseline algorithms.
-  More information about these algorithms can be found in table \ref{tbl:algo}
+  More information about these algorithms can be found in table \ref{tbl:algo}.
   Code for our evaluation, using the Lenskit evaluation framework \cite{lenskit} is available at \url{https://bitbucket.org/kluver/coldstartrecommendation}.
@@
       \hline
-      Algorithm          & Algorithm description \\\hline
+%      Algorithm          & Algorithm description \\\hline
       ItemBaseline       & Item's average rating, with mild bayesian damping towards the global mean \cite{funk_netflix_2006}. \\\hline
-      UserItem\-Baseline & ItemBasline adjusted by the user's average offset from the ItemBaseline. Mild mean damping is applied \cite{funk_netflix_2006}. \\\hline
+      UserItem\-Baseline & ItemBaseline adjusted by the user's average offset from the ItemBaseline. Mild mean damping is applied \cite{funk_netflix_2006}. \\\hline
       User-User          & User based nearest neighbor collaborative filtering \cite{resnick1994grouplens} with a neighborhood size of 30 and ratings normalized by subtracting the UserItemBaseline score.
@@
   Since this evaluation was on a different dataset (MovieLens 10M vs. MovieLens 1M), we performed minor tuning from these parameters.
-  Most notable, we found that the previous damping parameter used for the baselines was too large.
+  Most notably, we found that the previous damping parameter used for the baselines was too large.
   We found that a value of 5 gave much better results for new user recommendation.
-  During tuning we found that our key results are consistent among those configurations that perform well.
+  Our key results are not sensitive to minor changes of algorithm parameters.
   %TODO: carefully run this wording by joe.
@@
       RMSE                   & Standard prediction accuracy metric \cite{handbook}. & 0 to 5 & Smaller is better. \\\hline
-      nDCG                   & Standard ranking quality evaluation \cite{handbook}. & 0 to 1 & Larger is better. \\\hline
+      nDCG                   & Standard ranking quality metric \cite{handbook}. & 0 to 1 & Larger is better. \\\hline
       \hline
@@
       MAP@20                 & Average of precision at each rank one through twenty \cite{manning2008introduction}. & 0 to 1 & Larger is better.\\\hline
-      Fallout@20             & Count of recommended items in the test set and rated 2.0 or lower\footnotemark[1] \cite{handbook}. & 0 to 1 & Smaller is better. \\\hline
-      MeanRating@20          & Average rating of recommended items that are in the test set.& 0 to 5 & Larger is better \\\hline
-      RMSE@20                & RMSE computed only over those items that were recommended.& 0 to 5 & Smaller is better \\\hline
+      Fallout@20             & Count of recommended items in the test set and rated 2.0 or lower\footnotemark[1] \cite{handbook}. & -1 to 1 & Smaller is better. \\\hline
+      MeanRating@20          & Average rating of recommended items that are in the test set.& 0 to 5 & Larger is better. \\\hline
+      RMSE@20                & RMSE computed only over those items that were recommended.& 0 to 5 & Smaller is better. \\\hline
       \hline
@@
   \paragraph{Monotonic Recommendation Metrics}
-  These metrics all evaluate the recommendations produced by the recommender.
-  For recommendations we will use the 20 highest predicted items for each user, excluding items in the training set.
-  As the UserItem-Baseline and ItemBaseline make the same recommendations for all users, we will only report results for the ItemBaseline here.
+  These metrics evaluate the top 20 recommendations produced by the recommender.
+  As the UserItemBaseline and ItemBaseline make the same recommendations for all users, we will only report results for the ItemBaseline here.
   The choice of 20 items for the evaluations was arbitrary, testing with other values showed similar results, except on one metric.
@@
 
-  There metrics are useful because these metrics have a well defined monotonic relation to the satisfaction of the user.
-  As the metric grows larger, or smaller (depending on metric), we expect the output recommendations to satisfy users more.
-  This makes these metrics easier to understand than our later metrics.
+  These metrics are useful because they have a well defined relation to the satisfaction of the user.
+  As the metric grows larger, or smaller (depending on metric), we expect the recommendations to satisfy users more.
+  This makes these metrics easy to apply.
 
   Our first three metrics (Precision, MAP, and Fallout) are standard information retrieval metrics.
-  As we show later, we found Precision, MAP, and fallout to not be useful for this evaluation.
+  As we show later, we found these metrics to not be useful for this evaluation.
   Therefore, to help understand the quality of our recommendations, we also include MeanRating@20, and RMSE@20.
@@
   MeanRatings@20, and RMSE@20 both try to establish how much the user likes the recommendations.
-  MeanRatings@20 tells us directly how much the user tends to like recommended movies.
-  RMSE@20 accompanies this and lets us know if the recommended items are likely to be mispredicted, which can explain why users may or may not like their recommendations.
-  As it helps us interpret these two metrics we will also discuss SeenItems@20 with these metrics, although SeenItems@20 does not have as simple a relation to user satisfaction.
+  MeanRatings@20 tells us directly how much users tend to like recommended movies.
+  RMSE@20 accompanies this and lets us know if the recommended items are likely to be incorrectly predicted, which can explain why users may or may not like their recommendations.
+  As it helps us interpret these two metrics we will also discuss SeenItems@20 with these metrics, despite SeenItems@20 belonging to the next group of metrics.
 
@@
   Our final four metrics are all nonmonotonic measures of the user's recommendations.
-  These metrics are harder to understand than the monotonic measures of recommendation quality.
-  Unlike the previous metrics, we expect that both of the extremes of this metric would be perceived negatively by the user.
+  These metrics are harder to apply than the monotonic measures of recommendation quality.
+  Unlike the previous metrics, we expect that both of the extremes of these metrics relate to a negative user experience.
 
-  We use these metrics as they can help us understand important properties of how users may perceive the recommendations.
+  We use these metrics as they can help us understand important properties of how users perceive their recommendations.
   For example, as mentioned in the introduction how many of the recommended items the user has seen (estimated by SeenItems@20) can have a major impact on how much users trust the recommendations.
-  If an algorithm recommends too few items that the user has seen then we expect the user will have trouble evaluating if the recommendations are worth taking.
-  That said, however, if too many items in the recommendations have already been seen, the user may feel that the recommendations are not interesting or useful.
-  Unfortunately, without more research on these metrics to establish a mapping between user satisfaction and these values we will have to make educated guesses at what values for these metrics are too extreme.
+  If an algorithm recommends too few items that the user has seen then we expect the user will have trouble evaluating if the recommendations are trustworthy.
+  That said, if too many items in the recommendations have already been seen, the user may feel that the recommendations are not interesting or useful.
+  Unfortunately, without more research on these metrics to establish a mapping between user satisfaction and metric values we will have to rely on educated guesses to establish which values for these metrics are too extreme.
 
   The average popularity metric is a crude metric of how novel the recommendations are to the user \cite{zieglerDiversity}.
-  Broadly speaking we expect users will prefer recommendation lists containing more novel (less popular) items.
+  Broadly speaking, we expect users will prefer recommendation lists containing more novel (less popular) items.
   However, if the recommended items are too unpopular we risk that the user will not have heard of the items, which could make the user doubt the recommendations, hurting the user experience.
@@
   This is probably due to a tradeoff between how diverse a list is and how well it captures a user's tastes.
-  At some point, a list cannot become more diverse without include items that the user is not interested in.
+  At some point, a list cannot become more diverse without including items that do not interest the user.
   We will measure diversity with the AILS@20 metric.
@@
   Like diversity, however, we expect that it is not possible to reach complete spread (recommending each item an equal number of times) without making avoidably bad recommendations.
-  When Spread is large we know that the recommender is recommending many items with a relatively even distribution, this might be expected from a random recommender.
-  When Spread is small, the recommender is focusing on a small set of items which it recommends to every user, this might be expected from a baseline recommender.
+  When Spread is large we know that the recommender is recommending many items with a relatively even distribution.
+  When Spread is small, the recommender is focusing on a small set of items which it recommends to every user.
 
@@
   Let $C(I)$ be the count of how many times item $I$ showed up in the recommendations.
-  Let $P(I) = C(I) / \sum_I C(I)$ be the probability that a random recommendation is item $I$.
+  Let $P(I) = C(I) / \sum_I C(I)$ be the probability that a random recommendation is for item $I$.
   Finally we define spread as the Shannon's entropy of $P(I)$, $Spread = -\sum_I P(I) \log(P(I))$.
@@
   % provide expectations.
-  We expect that the item baseline algorithm will have approximately constant performance on all metrics.
-  Since the Item Baseline is such a simple algorithm we expect that any reasonable algorithm should be able to do better on it the monotonic metrics.
-  Furthermore, we expect the improvement over the item baseline to be gradual, with larger improvements only happening with more data.
+  We expect that the ItemBaseline algorithm will have approximately constant performance on all metrics.
+  Since the ItemBaseline is such a simple algorithm we expect that any reasonable algorithm should be able to do better than it.
+  Furthermore, we expect the improvement over the ItemBaseline to be gradual, with larger improvements only happening with more data.
 
-  The user-item baseline should only differ from the item baseline for metrics that use the prediction value.
+  The UserItemBaseline should only differ from the ItemBaseline for metrics that use the prediction value.
   For all other metrics these two should provide the same responses.
-  As the user-item baseline is still quite a simple baseline, we expect that the non baseline algorithm will beat this baseline as well.
+  As the UserItemBaseline is still quite a simple baseline, we expect that the non-baseline algorithm will beat this baseline as well.
   That said, for small rating values its possible that the improvements over this baseline will be very small.
@@
   We expect that any personalized algorithm should be able to make more nuanced recommendations than simply recommending the best rated items, which should translate as gains on these metrics.
-  Furthermore, we expect that as the algorithm learns more about the user it should decrease in popularity, and increase in diversity, and spread, as it becomes able to make more personal and nuanced recommendations to the user.
+  Furthermore, we expect that as the algorithm learns more about the user it should decrease in popularity, and increase in diversity, and spread, as it becomes able to make more personal and nuanced recommendations.
 
@@
 
-  For this evaluation we will use the MovieLens 1M dataset \footnote{\url{http://grouplens.org/datasets/movielens/}}.
+  For this evaluation we will use the MovieLens 1M dataset\footnote{\url{http://grouplens.org/datasets/movielens/}}.
   This dataset contains 1 million ratings from 6000 users on 4000 movies.
@@
   To the best of our knowledge, at the time this data was collected the new user survey preferred items that were very popular.
-  Therefore, as we cover shortly, will randomize the order of the ratings to try to avoid measuring the effect of these initial ratings, and instead focus on the algorithm's performance on normal ratings.
+  Therefore will randomize the order of the ratings to try to avoid measuring the effect of these initial ratings, and instead focus on the algorithm's performance on normal ratings.
 
@@
   The selection of a test set can be random to avoid any possible ordering effects in the data due to a new user rating strategy.
-  To understand how metrics change with a different number of users we can simply generate separate crossfolds for each number of ratings we wish to retain.
+  To understand how metrics change with a different number of users we can simply generate separate crossfolds for each simulated profile size.
   At each crossfold, rather than only taking a constant number of ratings for the test set, we can instead only keep a constant number of ratings for the training set.
@@
   Unfortunately, we found that this strategy can causes biases in our evaluation.
-  In particular, this strategy led to the size of the test sets to change as a function of the number of ratings retained.
-  As we kept more ratings for the training set, we would have fewer ratings in the testing set.
-  This turned out to bais our results, as some metrics, such as nDCG, proved to be sensitive to the number of items in the testing set.
+  In particular, this strategy led to the size of the test sets to change as a function of the simulated profile size.
+  As we kept more ratings for the training set, we would have fewer ratings in the test set.
+  This turned out to bias our results, as some metrics, such as nDCG, proved to be sensitive to the number of items in the test set.
 
   % describe our crossfold then subsample solution.
-  Our solution to this bais is to subsample an existing crossfold.
-  Therefore, to generate training and test sets for different number of ratings from 1 to 19, we first generate a set of test and training splits with 19 training items using the crossfold strategy outlined above.
+  Our solution to this bias is to subsample an existing crossfold.
+  Therefore, to generate training and test sets for different number of ratings from 1 to 19, we first generate a set of test and training splits so that test users have 19 training items using the crossfold strategy outlined above.
   Then, to generate any other number of ratings, we randomly downsample the 19 item training set.
-  This allows us to generate training sets with a given number of ratings retained for test users, such that each number of ratings retained uses the same tests set.
+  This allows us to generate training sets with a given number of ratings for test users, such that profile size is evaluated with the same test set.
 
@@
   User-User, and SVD both behaved essentially as expected.
-  SVD shows no improvement with only one rating, and User-User, actually appears to get worse with the first rating, but other than that, these algorithms outperform the item baseline.
-  Interestingly, the UserItem-Baseline outperforms both algorithms until around 8 ratings, at which point both algorithms start to perform better.
-  Given the size of the differences between UserItem-Baseline, UserUser, and SVD we conclude that all three provide about the same accuracy for new users.
+  SVD shows no improvement with only one rating, and User-User, actually appears to get worse with the first rating, but other than that, these algorithms outperform the ItemBaseline.
+  Interestingly, the UserItemBaseline outperforms both algorithms until around 8 ratings, at which point both algorithms start to perform better.
+  Given the size of the differences between UserItemBaseline, User-User, and SVD we conclude that all three provide about the same accuracy for new users.
   That such a simple algorithm can outperform more complicated for such small values goes to show that recommending for users with very few ratings is a hard problem.
@@
   If you consider the accuracy of RMSE on only those items where it makes a personalized prediction it does show a monotonic increase in accuracy similar to the other algorithms.
-  However, for small rating counts ItemItem can only predict for some percent of the item (around 60\% for users with two ratings, and around 75\% for users with four ratings).
-  Because of this, many of the predictions come from the better performing UserItem-Baseline.
-  As we get our first few ratings, the predictions become less likely to be from the UserItem-Baseline, leading to an artificial upward trend in RMSE.
+  However, for small rating counts Item-Item can only predict for some percent of the item (around 60\% for users with two ratings, and around 75\% for users with four ratings).
+  Because of this, many of the predictions come from the better performing UserItemBaseline.
+  As we get our first few ratings, the predictions become less likely to be from the UserItemBaseline, leading to an artificial upward trend in RMSE.
   Fortunately, Item-Item seems to improve accuracy faster than the other algorithms, and therefore appears as if it will catch up.
-  Unfortunately, this result suggests that ItemItem may be unsuitable for new users.
+  Unfortunately, this result suggests that Item-Item may be unsuitable for new users.
 
@@
   The results for nDCG are very similar to the RMSE results.
-  Both UserUser and SVD take an initial hit to accuracy at the first rating.
-  This time UserUser takes much longer to recover from this and only beats the baseline after around 12 ratings.
-  Again ItemItem performs quite poorly.
+  Both User-User and SVD take an initial hit to accuracy at the first rating.
+  This time User-User takes much longer to recover from this and only beats the baseline after around 12 ratings.
+  Again Item-Item performs quite poorly.
   This leaves SVD to perform the best, passing the baseline after about 3 ratings.
@@
   All three metrics show essentially the same trend, User-User gets the lowest score, the baseline get the highest score.
-  On both MAP@20 and precision@20 svd gets a higher score than Item-Item throughout, performing almost as highly as the baseline.
+  On both MAP@20 and precision@20 SVD gets a higher score than Item-Item throughout, performing almost as highly as the baseline.
   On fallout@20 Item-Item and SVD perform about as well as each other, with SVD showing a very slight lead.
@@
   The average number of test set items in recommendations shows a similar trend to the MAP, precision, and recall metrics, with user-user getting by far the fewest seen items, and the baseline getting the most.
-  As User-User gets less than one seen item on average we found that MeanRating@20 and RMSE@20 metrics were not appropriate for UserUser.
+  As User-User gets less than one seen item on average we found that MeanRating@20 and RMSE@20 metrics were not appropriate for User-User.
   This was observed in particular when testing our results for sensitivity to recommendation list size.
-  We found that different recommendation lists lead UserUser (and only UserUser) to have very different results.
-  Because of this issue we will not include User-User in our next evaluations and consider the next two metrics inconclusive for UserUser.
-  That said, we consider the small number of seen items itself a bad result for UserUser.
+  We found that different recommendation lists lead User-User (and only User-User) to have very different results.
+  Because of this issue we will not include User-User in our next evaluations and consider the next two metrics inconclusive for User-User.
+  That said, we consider the small number of seen items itself a bad result for User-User.
   Put simply, if we cannot reliably estimate if the recommendations are reasonable, how can we expect a user to do so?
@@
   Again, we find that Item-Item performs poorly until quite a few ratings are added, and does worse than baseline throughout.
-  This is quite possibly a similar trend to the one we observed with RMSE and nDCG, where ItemItem's performance decreased as coverage increased leading to the curved performance at cold start.
-  Regardless, we find that ItemItem's recommendations do not seem to be as well rated as either the baseline or SVD.
+  This is quite possibly a similar trend to the one we observed with RMSE and nDCG, where Item-Item's performance decreased as coverage increased leading to the curved performance at cold start.
+  Regardless, we find that Item-Item's recommendations do not seem to be as well rated as either the baseline or SVD.
 
@@
   The first graph in figure \ref{fig:pop} shows the average popularity of recommended items by each algorithm.
-  As expected the baseline algorithms makes the most popular recommendations, followed by SVD, ItemItem, and then UserUser in a distant last.
+  As expected the baseline algorithms makes the most popular recommendations, followed by SVD, Item-Item, and then User-User in a distant last.
   As stated earlier, We expect that users will feel the baseline's popularity (around 1000) is too high.
@@
   SVD and Item-Item both seem to perform well, with SVD generating more popular recommendations than Item-Item.
-  SVD and ItemItem both show a trend of decreasing popularity as more ratings are retained, suggesting that these algorithms will make more nuanced, less popular recommendations for users they know more about.
+  SVD and Item-Item both show a trend of decreasing popularity as more ratings are retained, suggesting that these algorithms will make more nuanced, less popular recommendations for users they know more about.
   Interestingly, these algorithms each appear to be converging to their own separate level of popularity, which suggests that this ordering may be independent of the number of ratings.
@@
   Again, we expect that the baseline, which provides the least diverse recommendations, would be precised as not very diverse by users.
-  Quite possibly due to the preference for obscure movies, UserUser seems to produce the most diverse lists of any algorithm.
-  We expect that the UserUser recommendation lists might be perceived as too diverse.
-  Unlike popularity, SVD and ItemItem seem to perform equivalently in terms of diversity.
+  Quite possibly due to the preference for obscure movies, User-User seems to produce the most diverse lists of any algorithm.
+  We expect that the User-User recommendation lists might be perceived as too diverse.
+  Unlike popularity, SVD and Item-Item seem to perform equivalently in terms of diversity.
   We also see that all three collaborative filtering algorithms show a trend of increasing diversity (decreasing  inter-list similarity) as more ratings are retained.
@@
   \paragraph{Spread@20}
-  The third graph in \ref{fig:spread} shows the spread of the recommendations from each algorithm.
+  The third graph in \ref{fig:pop} shows the spread of the recommendations from each algorithm.
   As expected the baseline performs worst on spread, as it recommends the same items to each user (unless the user has already rated one of these items).
   SVD performs mildly better than the baseline, and shows definite improvement as more ratings are added.
-  UserUser and ItemItem start at around the same value as each other, however, after only two ratings they have crossed, as ItemItem quickly increases to have the highest spread of all the algorithms, and UserUser quickly decreases into second place.
-  Interestingly, only UserUser shows a trend to have less spread ratings as it learns more about the user, ItemItem and SVD both show the expected trend of increasing Spread as they learn more.
+  User-User and Item-Item start at around the same value as each other, however, after only two ratings they have crossed, as Item-Item quickly increases to have the highest spread of all the algorithms, and User-User quickly decreases into second place.
+  Interestingly, only User-User shows a trend to have less spread ratings as it learns more about the user, Item-Item and SVD both show the expected trend of increasing Spread as they learn more.
   This could be some form of a regression towards the mean, where as it learns more about users its recommendations become less personalized to the users individual quirks.
@@
       Algorithm & Prediction accuracy & Recommendation quality      & Other properties \\\hline
-      ItemItem  & Poor                & Poor                        & Good             \\\hline
-      UserUser  & Good                & Inconclusive                & Poor             \\\hline
+      Item-Item  & Poor                & Poor                        & Good             \\\hline
+      User-User  & Good                & Inconclusive                & Poor             \\\hline
       SVD       & Good                & Good                        & Good             \\\hline
@@
   SVD was the only algorithm that consistently outperformed the baseline algorithms on prediction and recommendation tasks.
-  ItemItem had arguably better results on the average popularity and spread metrics, indicating that it was able to find more novel items, and make a larger range of recommendations to users.
+  Item-Item had arguably better results on the average popularity and spread metrics, indicating that it was able to find more novel items, and make a larger range of recommendations to users.
   Unfortunately, Item-Item also performed quite poorly at both prediction and recommendation, which may make it less suitable for new users.
 
-  UserUser performed quite well are prediction, but tended to favor unpopular items in its recommendations.
-  While some novelty is good, we believe that UserUser goes too far into the obscure movies in its recommendations, and therefore does not make good recommendations.
-  Therefore we believe that UserUser is not well suited for new users unless only its predictions are used.
+  User-User performed quite well are prediction, but tended to favor unpopular items in its recommendations.
+  While some novelty is good, we believe that User-User goes too far into the obscure movies in its recommendations, and therefore does not make good recommendations.
+  Therefore we believe that User-User is not well suited for new users unless only its predictions are used.
   Exploring corrections for this issue is left as future work.
@@
   One possible reason that the SVD algorithm performs as well as it does is the regularization built into the algorithm.
-  UserUser and ItemItem algorithms have no built in method for damping their predictions towards the mean when they don't have much data.
+  User-User and Item-Item algorithms have no built in method for damping their predictions towards the mean when they don't have much data.
   The regularization in the cost function of the SVD model makes it so its predictions will gradually shift away from the baseline predictions as more relevant ratings are added.

### 2014-05-07 paper/main.tex
edited the methodology section

@@
-%TODO: update language retain to simulated profile size. do a search ``reta''
-
 \documentclass[letterpaper]{sig-alternate}
@@
 
-  Using the methodology described in section \ref{sec:methodology} we generate plots for each metric showing how the five algorithms perform new users, versus how many of the user's ratings were retained in the training set.
+  Using the methodology described in section \ref{sec:methodology} we generate plots for each metric showing how the five algorithms perform new users, versus simulated profile size.
 
@@
   Given the size of the differences between UserItemBaseline, User-User, and SVD we conclude that all three provide about the same accuracy for new users.
-  That such a simple algorithm can outperform more complicated for such small values goes to show that recommending for users with very few ratings is a hard problem.
+  That such a simple algorithm can outperform more complicated algorithms for small profile sizes suggests that recommending for users with very few ratings is a hard problem.
 
@@
   SVD and Item-Item both seem to perform well, with SVD generating more popular recommendations than Item-Item.
-  SVD and Item-Item both show a trend of decreasing popularity as more ratings are retained, suggesting that these algorithms will make more nuanced, less popular recommendations for users they know more about.
+  SVD and Item-Item both show a trend of decreasing popularity as more ratings are entered, suggesting that these algorithms will make more nuanced, less popular recommendations for users they know more about.
   Interestingly, these algorithms each appear to be converging to their own separate level of popularity, which suggests that this ordering may be independent of the number of ratings.
@@
   Unlike popularity, SVD and Item-Item seem to perform equivalently in terms of diversity.
-  We also see that all three collaborative filtering algorithms show a trend of increasing diversity (decreasing  inter-list similarity) as more ratings are retained.
+  We also see that all three collaborative filtering algorithms show a trend of increasing diversity (decreasing  inter-list similarity) as more ratings are entered.
 

### 2014-05-07 paper/main.tex
revised results

@@
   Rather than picking advanced algorithms that are tuned for new user performance, we decided to first understand how representatives of three common types of algorithms perform.
-  Therefore we will compare User-User\cite{resnick1994grouplens}, Item-Item\cite{sarwar2001item}, and Simon Funk's SVD\cite{funk_netflix_2006} algorithms against two simple baseline algorithms.
+  Therefore we will compare UserUser\cite{resnick1994grouplens}, ItemItem\cite{sarwar2001item}, and Simon Funk's SVD\cite{funk_netflix_2006} algorithms against two simple baseline algorithms.
   More information about these algorithms can be found in table \ref{tbl:algo}.
@@
       UserItem\-Baseline & ItemBaseline adjusted by the user's average offset from the ItemBaseline. Mild mean damping is applied \cite{funk_netflix_2006}. \\\hline
-      User-User          & User based nearest neighbor collaborative filtering \cite{resnick1994grouplens} with a neighborhood size of 30 and ratings normalized by subtracting the UserItemBaseline score.
+      UserUser           & User based nearest neighbor collaborative filtering \cite{resnick1994grouplens} with a neighborhood size of 30 and ratings normalized by subtracting the UserItemBaseline score.
 \\\hline
-      Item-Item          & Item based nearest neighbor collaborative filtering \cite{sarwar2001item} with a neighborhood size of 30 and ratings normalized by subtracting the ItemBaseline score.
+      ItemItem           & Item based nearest neighbor collaborative filtering \cite{sarwar2001item} with a neighborhood size of 30 and ratings normalized by subtracting the ItemBaseline score.
 \\\hline
@@
   AILS@N is simply a rescaling of ILS to be scale free with regards to N, this was done to make the metric easier to interpret, by putting it on the same scale as similarity values.
-  We used the same item similarity metric for AILS@20 as we used in the Item-Item recommender.
+  We used the same item similarity metric for AILS@20 as we used in the ItemItem recommender.
 
@@
 
-\subsection{Accuracy and Ranking Metrics}
+\subsection*{Accuracy and Ranking Metrics}
 
@@
   Figure \ref{fig:rmse} shows how accurate our algorithms are as measured by RMSE.
-  User-User, and SVD both behaved essentially as expected.
-  SVD shows no improvement with only one rating, and User-User, actually appears to get worse with the first rating, but other than that, these algorithms outperform the ItemBaseline.
-  Interestingly, the UserItemBaseline outperforms both algorithms until around 8 ratings, at which point both algorithms start to perform better.
-  Given the size of the differences between UserItemBaseline, User-User, and SVD we conclude that all three provide about the same accuracy for new users.
+  UserUser, and SVD both behaved essentially as expected.
+  SVD shows no improvement with only one rating, and UserUser actually gets worse with the first rating, but other than that, these algorithms outperform ItemBaseline.
+  Interestingly, UserItemBaseline outperforms both algorithms until around 8 ratings, at which point both algorithms start to perform better.
+  Given the size of the differences between UserItemBaseline, UserUser, and SVD we conclude that all three are equally accurate for new users.
   That such a simple algorithm can outperform more complicated algorithms for small profile sizes suggests that recommending for users with very few ratings is a hard problem.
 
-  Item-Item performs quite poorly for new users.
-  Until we have 13 ratings, Item-Item is less accurate than the ItemBaseline.
-  In fact, for the first two ratings Item-Item appears to perform \emph{worse} as more ratings are added.
-  This trend is a result of the limited coverage of the Item-Item algorithm.
+  ItemItem performs quite poorly for new users.
+  Until we have 13 ratings, ItemItem is less accurate than the ItemBaseline.
+  In fact, for the first two ratings ItemItem appears to perform \emph{worse} as more ratings are added.
+  This trend is a result of the limited coverage of the ItemItem algorithm.
   If you consider the accuracy of RMSE on only those items where it makes a personalized prediction it does show a monotonic increase in accuracy similar to the other algorithms.
-  However, for small rating counts Item-Item can only predict for some percent of the item (around 60\% for users with two ratings, and around 75\% for users with four ratings).
+  However, for small rating counts ItemItem can only predict for some percent of the item (around 60\% for users with two ratings, and around 75\% for users with four ratings).
   Because of this, many of the predictions come from the better performing UserItemBaseline.
   As we get our first few ratings, the predictions become less likely to be from the UserItemBaseline, leading to an artificial upward trend in RMSE.
-  Fortunately, Item-Item seems to improve accuracy faster than the other algorithms, and therefore appears as if it will catch up.
-  Unfortunately, this result suggests that Item-Item may be unsuitable for new users.
 
@@
   \paragraph{nDCG}
-  Figure \ref{fig:ndcg} shows the algorithm's performance on the nDCG metric.
+  Figure \ref{fig:ndcg} shows the algorithms performance on the nDCG metric.
   The results for nDCG are very similar to the RMSE results.
-  Both User-User and SVD take an initial hit to accuracy at the first rating.
-  This time User-User takes much longer to recover from this and only beats the baseline after around 12 ratings.
-  Again Item-Item performs quite poorly.
-  This leaves SVD to perform the best, passing the baseline after about 3 ratings.
+  Both UserUser and SVD take an initial hit to accuracy at the first rating.
+  This time UserUser takes much longer to recover and only beats the baseline after around 12 ratings.
+  Again ItemItem performs quite poorly.
+  This leaves SVD performing the best, passing the baseline after about 3 ratings.
   Again we find that our best algorithm for the first few ratings is our baseline.
@@
 
-\subsection {Monotonic Recommendation Quality Metrics}
+\subsection*{Monotonic Recommendation Quality Metrics}
 
@@
   Figure \ref{fig:map} shows MAP@20, precision@20, and fallout@20.
-  All three metrics show essentially the same trend, User-User gets the lowest score, the baseline get the highest score.
-  On both MAP@20 and precision@20 SVD gets a higher score than Item-Item throughout, performing almost as highly as the baseline.
-  On fallout@20 Item-Item and SVD perform about as well as each other, with SVD showing a very slight lead.
+  All three metrics show essentially the same trend, UserUser gets the lowest score, the baseline get the highest score.
+  On both MAP@20 and precision@20 SVD gets a higher score than ItemItem throughout, performing almost as highly as the baseline.
+  On fallout@20 ItemItem and SVD perform about as well as each other, with SVD showing a very slight lead.
   We also tested recall and mean reciprocal rank metrics, but found them to exhibit the same trend, so they were not explored further.
@@
   Therefore we would not expect to see rank equivalent behavior from these metrics.
-  Unfortunately this means that these metrics contradict each other in terms of which algorithm performs the best.
-  One possible reason for this is that these topN metrics are known to be biased toward recommendations that have more popular items \cite{bellogin}.
+  Unfortunately this means that these metrics contradict each other.
+  One possible reason for this is that these topN metrics are known to be biased toward algorithms that recommend more popular items \cite{bellogin}.
   Looking forward to the average popularity (figure \ref{fig:pop}) of recommended items, we see that the average popularity of the recommendations closely replicates the previous three plots.
@@
   We address this metric out of turn here as it provides important context for the next two metrics.
-  The average number of test set items in recommendations shows a similar trend to the MAP, precision, and recall metrics, with user-user getting by far the fewest seen items, and the baseline getting the most.
-  As User-User gets less than one seen item on average we found that MeanRating@20 and RMSE@20 metrics were not appropriate for User-User.
-  This was observed in particular when testing our results for sensitivity to recommendation list size.
-  We found that different recommendation lists lead User-User (and only User-User) to have very different results.
-  Because of this issue we will not include User-User in our next evaluations and consider the next two metrics inconclusive for User-User.
-  That said, we consider the small number of seen items itself a bad result for User-User.
+  The average number of test set items in recommendations shows a similar trend to the MAP and precision metrics, with UserUser getting by far the fewest seen items, and the baseline getting the most.
+  As UserUser gets less than one seen item on average we found that UserUser was not appropriate for the MeanRating@20 and RMSE@20 metrics.
+  In particular, we found that the UserUser results for these metrics was highly sensitive to recommendation list size.
+  Because of this issue we will not include UserUser in our next evaluations and consider the next two metrics inconclusive for UserUser.
+  That said, we consider the small number of seen items itself a bad result for UserUser.
   Put simply, if we cannot reliably estimate if the recommendations are reasonable, how can we expect a user to do so?
@@
   SVD performs as expected, starting around the baseline for few ratings, and then steadily improving to be the best algorithm after 3 ratings.
-  Again, we find that Item-Item performs poorly until quite a few ratings are added, and does worse than baseline throughout.
-  This is quite possibly a similar trend to the one we observed with RMSE and nDCG, where Item-Item's performance decreased as coverage increased leading to the curved performance at cold start.
-  Regardless, we find that Item-Item's recommendations do not seem to be as well rated as either the baseline or SVD.
+  Again, we find that ItemItem performs poorly until quite a few ratings are added, and does worse than baseline throughout.
 
   \paragraph{RMSE@20}
-  The third graph in figure \ref{fig:topN.rmse} shows the RMSE our algorithms got on items in the top 20 recommendations for each algorithm.
+  The third graph in figure \ref{fig:topN.rmse} shows the accuracy of our algorithms on items in the top 20 recommendations.
   Not surprisingly, this figure shows the same trends as the last one, meaning that the algorithms whose recommendations were (on average) most likable, were those whose recommendations were most accurate.
@@
 
-  Interestingly, we find that all algorithms are quite a bit more accurate on the top 20 items than they are for the general item collection.
-  This is likely an edge effect, where the highest rated (5 star) items are perhaps easier to predict than other items.
+  %Interestingly, we find that all algorithms are quite a bit more accurate on the top 20 items than they are for the general item collection.
+  %This is likely an edge effect, where the highest rated (5 star) items are perhaps easier to predict than other items.
 
-\subsection{Nonmonotonic Recommender Quality Metrics}
+\subsection*{Nonmonotonic Recommender Quality Metrics}
   % Other recommender metrics
@@
     % popularity graph
-  The first graph in figure \ref{fig:pop} shows the average popularity of recommended items by each algorithm.
-  As expected the baseline algorithms makes the most popular recommendations, followed by SVD, Item-Item, and then User-User in a distant last.
+  The first graph in figure \ref{fig:pop} shows the average popularity of items recommended by each algorithm.
+  As expected the baseline algorithms makes the most popular recommendations, followed by SVD, ItemItem, and then UserUser in a distant last.
   As stated earlier, We expect that users will feel the baseline's popularity (around 1000) is too high.
-  We also expect that users will perceive User-User's popularity (around 50) as too low.
-  SVD and Item-Item both seem to perform well, with SVD generating more popular recommendations than Item-Item.
-  SVD and Item-Item both show a trend of decreasing popularity as more ratings are entered, suggesting that these algorithms will make more nuanced, less popular recommendations for users they know more about.
-  Interestingly, these algorithms each appear to be converging to their own separate level of popularity, which suggests that this ordering may be independent of the number of ratings.
+  We also expect that users will perceive UserUser's popularity (around 50) as too low.
+  SVD and ItemItem both seem to perform well, with SVD generating more popular recommendations than Item-Item.
+  SVD and ItemItem both show a trend of decreasing popularity as more ratings are entered, suggesting that these algorithms will make more nuanced, less popular recommendations for users they know more about.
+  %Interestingly, these algorithms each appear to be converging to their own separate level of popularity, which suggests that this ordering may be independent of the number of ratings.
 
@@
   The second graph in \ref{fig:pop} shows a similar trend to the popularity trend.
-  Again, we expect that the baseline, which provides the least diverse recommendations, would be precised as not very diverse by users.
-  Quite possibly due to the preference for obscure movies, User-User seems to produce the most diverse lists of any algorithm.
-  We expect that the User-User recommendation lists might be perceived as too diverse.
-  Unlike popularity, SVD and Item-Item seem to perform equivalently in terms of diversity.
+  Again, we expect that the baseline, which provides the least diverse recommendations, would be perceived as not very diverse by users.
+  Quite possibly due to the preference for obscure movies, UserUser seems to produce the most diverse lists of any algorithm.
+  We expect that the UserUser recommendation lists might be too diverse.
+  Unlike popularity, SVD and ItemItem seem to perform equivalently in terms of diversity.
   We also see that all three collaborative filtering algorithms show a trend of increasing diversity (decreasing  inter-list similarity) as more ratings are entered.
@@
   \paragraph{Spread@20}
-  The third graph in \ref{fig:pop} shows the spread of the recommendations from each algorithm.
-  As expected the baseline performs worst on spread, as it recommends the same items to each user (unless the user has already rated one of these items).
+  The third graph in figure \ref{fig:pop} shows the spread of the recommendations from each algorithm.
+  As expected the baseline performs worst on spread, as it recommends essentially the same items to each user.
   SVD performs mildly better than the baseline, and shows definite improvement as more ratings are added.
-  User-User and Item-Item start at around the same value as each other, however, after only two ratings they have crossed, as Item-Item quickly increases to have the highest spread of all the algorithms, and User-User quickly decreases into second place.
-  Interestingly, only User-User shows a trend to have less spread ratings as it learns more about the user, Item-Item and SVD both show the expected trend of increasing Spread as they learn more.
+  UserUser and ItemItem start at around the same value as each other, however, after only two ratings they have crossed, as ItemItem quickly increases to have the highest spread of all the algorithms, and UserUser quickly decreases into second place.
+  Interestingly, only UserUser shows a trend to have less spread as it learns more about the user, ItemItem and SVD both show the expected trend of increasing Spread as they learn more.
   This could be some form of a regression towards the mean, where as it learns more about users its recommendations become less personalized to the users individual quirks.
@@
       Algorithm & Prediction accuracy & Recommendation quality      & Other properties \\\hline
-      Item-Item  & Poor                & Poor                        & Good             \\\hline
-      User-User  & Good                & Inconclusive                & Poor             \\\hline
+      ItemItem  & Poor                & Poor                        & Good             \\\hline
+      UserUser  & Good                & Inconclusive                & Poor             \\\hline
       SVD       & Good                & Good                        & Good             \\\hline
@@
   SVD was the only algorithm that consistently outperformed the baseline algorithms on prediction and recommendation tasks.
-  Item-Item had arguably better results on the average popularity and spread metrics, indicating that it was able to find more novel items, and make a larger range of recommendations to users.
-  Unfortunately, Item-Item also performed quite poorly at both prediction and recommendation, which may make it less suitable for new users.
+  ItemItem had arguably better results on the average popularity and spread metrics, indicating that it was able to find more novel items, and make a larger range of recommendations to users.
+  Unfortunately, ItemItem also performed quite poorly at both prediction and recommendation, which may make it less suitable for new users.
 
-  User-User performed quite well are prediction, but tended to favor unpopular items in its recommendations.
-  While some novelty is good, we believe that User-User goes too far into the obscure movies in its recommendations, and therefore does not make good recommendations.
-  Therefore we believe that User-User is not well suited for new users unless only its predictions are used.
+  UserUser performed quite well are prediction, but tended to favor unpopular items in its recommendations.
+  While some novelty is good, we believe that UserUser goes too far into the obscure movies in its recommendations, and therefore does not make good recommendations.
+  Therefore we believe that UserUser is not well suited for new users unless only its predictions are used.
   Exploring corrections for this issue is left as future work.
@@
   One possible reason that the SVD algorithm performs as well as it does is the regularization built into the algorithm.
-  User-User and Item-Item algorithms have no built in method for damping their predictions towards the mean when they don't have much data.
+  UserUser and ItemItem algorithms have no built in method for damping their predictions towards the mean when they don't have much data.
   The regularization in the cost function of the SVD model makes it so its predictions will gradually shift away from the baseline predictions as more relevant ratings are added.

### 2014-05-07 paper/main.tex
finished third draft

@@
     \end{tabular}
-    \caption{Summary of algorithm quality for new users}
+    \caption{Summary of algorithm behavior for new users}
     \label{tbl:results}
@@
   ItemItem had arguably better results on the average popularity and spread metrics, indicating that it was able to find more novel items, and make a larger range of recommendations to users.
-  Unfortunately, ItemItem also performed quite poorly at both prediction and recommendation, which may make it less suitable for new users.
+  Unfortunately, ItemItem also performed quite poorly at both prediction and recommendation, which makes it less suitable for new users.
 
-  UserUser performed quite well are prediction, but tended to favor unpopular items in its recommendations.
-  While some novelty is good, we believe that UserUser goes too far into the obscure movies in its recommendations, and therefore does not make good recommendations.
-  Therefore we believe that UserUser is not well suited for new users unless only its predictions are used.
+  UserUser performed quite well at prediction, but tended to favor unpopular items in its recommendations.
+  While some novelty is good, we believe that UserUser goes too far with obscure movies in its recommendations, and therefore does not make good recommendations.
+  Therefore we conclude that UserUser is not well suited for new users unless only its predictions are used.
   Exploring corrections for this issue is left as future work.
@@
     % non-traditional metrics can tell you more about what is going on, these seem to be more properties of the class of algo than of the data, so its worth looking at these when choosing algorithms, to find one with properties that are good for your task.
-  Interestingly, however, we see that for very small numbers of ratings the UserItemBaseline seems to be our strongest algorithm, both in terms of accuracy (as measured by RMSE), and recommendation quality (as measured by MeanRating@20).
-  This suggests a switching recommendation strategy for live systems.
-  Instead of recommending from one recommender right away, a system could provide baseline predictions and recommendations until the user has enough ratings and then switching to a better recommender.
-  This has the added avantage that it might help users notice the changes in popularity and diversity, helping them appreciate these properties more.
-  For switching recommendation strategies like this, our methodology provides a way to find the correct number of ratings at which to switch algorithms.
+  Interestingly, we repeatedly see that for very small numbers of ratings the UserItemBaseline is our strongest algorithm, both in terms of accuracy, and recommendation quality.
+  This suggests a switching strategy for live systems, in which the UserItemBaseline is deployed until a user has enough ratings, after which they are switched to a better recommender.
+  The switching point between algorithms can be determined with our methodology.
+  The only downside of this approach is that the baseline performed poorly on our nonmonotonic metrics, which may effect users as they first join the system.
 
@@
   The downside of this, however, could be the higher popularity and lower spread seen by the SVD algorithm.
-  The same regularization that helps avoid making mistakes due to too little data, may also prevents the algorithm from taking occasional risks on more obscure movies.
+  The same regularization that helps avoid making mistakes due to too little data, may also prevent the algorithm from taking occasional risks on more obscure movies.
 
@@
   Future work should replicate our experiment on other datasets from different recommender domains.
-  By doing this we will find out which of our results are true properties of the algorithms, and which results are instead related to properties of our dataset.
+  By doing this we will find out which of our results are true properties of the algorithms, and which results are specific to our dataset.
 
@@
   This is a limitation of our methodology, to maintain a constant test set throughout the evaluation we are not able to evaluate users who have less ratings than the maximum amount we wish to study.
-  More work needs to be done to understand users who only ever make one or two ratings to understand why they leave, and how they differ from other users of our systems.
+  More work needs to be done to understand users who only  make one or two ratings to discover why they leave, and how they differ from other users of our systems.
 
-  Finally, we feel that much more work should be done using this methodology to explore different algorithms.
-  There are many algorithms beyond the three e evaluated.
-  Some of these algorithms are designed specifically for cold start.
-  Others are designed to leverage extra, non-rating information to improve quality.
+  Finally, we feel that more work is needed to explore different algorithms.
+  We only explored three of the dozens of known recommendation algorithms.
+  Some of these algorithms are designed specifically for cold start, others are designed to leverage extra, non-rating information to improve quality.
   Work should be done to understand the differences between these algorithms, and what effect extra information has on the recommendation process.
-  Using our methodology we might be able to come up with an understanding of the value that extra information about the user has on recommendation, and to compare it the value of extra ratings for new users.
+  Using our methodology should might be possible to measure the effect that extra information about the user has on recommendation, and to compare it the effect of extra initial ratings.
     % algorithm limitations
@@
 
-  Our work shows that a multiple metric approach to understanding recommender behavior can provide useful insights into the differences between algorithms.
-  Our methodology presents a list of metrics that can be reused in the future for new multiple metric evaluations, and our methodology shows how this can be applied to understand new users.
+  Our work shows that a multiple metric approach to understanding recommender behavior can provide useful insights into the differences between algorithms for new users.
+  Our methodology presents a list of metrics, and an evaluation framework that can be used to perform further evaluations in the future.
   We find that recommending for new users is hard and none of our algorithms can reliably beat a simple baseline for users with very few ratings.
   Once the user has a small number of ratings we find that the SVD algorithm performs best.
-  This result can be useful both in designing new recommender systems, and in designing new recommendation algorithms.
+  This result can be useful both in designing new recommender systems, and in designing novel recommendation algorithms for new users.
 

### 2014-05-07 paper/main.tex
almost done

@@
   Our work focuses on how different algorithms behave for new users.
-  We describe a methodology that we use to compare representative of three common families of algorithms along eleven different metrics.
-  We find for first few ratings a baseline algorithm performs better than three common collaborative filtering algorithms.
+  We describe a methodology that we use to compare representatives of three common families of algorithms along eleven different metrics.
+  We find that for the first few ratings a baseline algorithm performs better than three common collaborative filtering algorithms.
   Once we have a few ratings, we find that Funk's SVD algorithm has the best overall performance.
+  We also find that ItemItem, a very commonly deployed algorithm, performs very poorly for new users.
   Our results can inform the design of interfaces and algorithms for new users.
@@
 % A category with the (minimum) three required fields
-\category{H.3.3}{nformation Search and Retrieval}{Information filtering}
+\category{H.3.3}{Information Search and Retrieval}{Information filtering}
 
@@
   If the system cannot provide a good user experience, then new users are likely to leave and not return.
-  Even if the system can make accurate recommendations accuracy alone is not enough to make a good first impression.
+  Even if the system can make accurate recommendations, accuracy alone is not enough to make a good first impression.
   If the system only recommends items the user has seen, then the user may see no value in the system and leave.
@@
 
-  The natural way to improve the system for new users is to have access to some extra information.
-  The best studied strategy is for new users to go through a new user rating survey.
-  Systems that take this practice do not allow their users to receive recommendations until some fixed number of ratings have been entered.
+  The most studied strategy for gathering more information for new users is to include a new user rating survey.
+  Systems that utilize this practice do not allow their users to receive recommendations until some fixed number of ratings have been entered.
   The idea is that after enough ratings have been entered, the predictions will be `good enough'.
-  This idea has been explored by many authors, most consider a system in which the user is presented with a list of items, and asked to rate items until they have enough ratings.
+  This idea has been explored by many authors; most consider a system in which the user is presented with a list of items and asked to rate items until they have enough ratings.
   For a good discussion of prior work in this direction see \cite{adaptiveBootstrapping}.
@@
   The new user survey work sidesteps this issue by assuming that the user will receive no personalization until the survey is complete.
-  Even for systems with a new user survey carefully choosing algorithms for new users may be very important.
+  Even for systems with a new user survey, carefully choosing algorithms for new users may be very important.
   Users fresh out of the new user survey will still have far fewer ratings than the average user of the recommender system.
@@
   Fortunately, there has been a growing trend in recommender system research to perform multiple metric evaluations of recommender systems.
-  This work comes out of a pushback against early recommendation evaluations which only evaluated recommenders in terms of their predictive accuracy \cite{accurateIsNotEnough}.
+  This work comes out of a pushback against early recommender evaluations which only evaluated recommenders in terms of their predictive accuracy \cite{accurateIsNotEnough}.
   The response of this pushback has been the generating of many different metrics which allow us to more carefully understand the separate aspects of a recommender's output \cite{LathiaTemporal, CremonesiTopN, zieglerDiversity}.
@@
 
-  We combine these two types of research, new user experience research, and multiple metric algorithm evaluation.
+  In this work we combine these two streams of research: new user experience research and multiple metric algorithm evaluation.
   Our goal is to understand the differences between algorithms for new users, especially how these algorithms respond to different amounts of information.
-  by using multiple metrics we can understand not just what algorithm is best in terms of absolute performance, but also gain a more nuanced understanding of the qualities of different algorithm's recommendations.
+  By using multiple metrics we can understand not just what algorithm is best in terms of absolute performance, but also gain a more nuanced understanding of the qualities of different algorithms' recommendations.
   %Furthermore, we do not seek to find the algorithm that scores best on some metric of choice, instead, we seek to form a complete understanding of the differences between algorithms for cold start use.
 
-  We structure our work around the following research question.
+  We structure our work around the following research question:
   \begin{itemize}
@@
   Algorithm behavior is a very broad subject.
-  We will look at three distinct sub questions to help capture algorithm behavior.
+  We will look at three distinct sub-questions to help capture algorithm behavior.
   %We will look at three distinct ways to quantify the behavior of a recommender system. each of which will lead to multiple metrics.
@@
   \item RQ1.b: How well can different algorithms rank and recommend good items to new users?
-  \item RQ1.c: How do algorithms behave as measured by other metrics such as popularity and diversity for new users?
+  \item RQ1.c: How do algorithms behave as measured by other metrics, such as popularity and diversity, for new users?
   \end{itemize}
@@
 
-  To answer our research questions we will evaluate algorithms in an offline simulation designed to simulate users as they join the system.
-  We do this by adopting the standard train and test methodology to simulate users with a given number of ratings.
-  We run this methodology for various profile sizes, and then plot the results.
+  To answer our research questions we will evaluate algorithms in an offline evaluation designed to simulate users as they join the system.
+  We do this by adapting the standard train and test methodology to test existing users with a specific number of ratings.
+  We run this approach for various profile sizes, and then plot the results.
   This lets us understand both how the algorithms compare on these metrics, and how the algorithm performance changes with the number of ratings provided.
@@
 
-  We seek to develop an understanding of how a range of different standard algorithms perform at for new users.
+  We seek to develop an understanding of how a range of different standard algorithms perform for new users.
   Rather than picking advanced algorithms that are tuned for new user performance, we decided to first understand how representatives of three common types of algorithms perform.
-  Therefore we will compare UserUser\cite{resnick1994grouplens}, ItemItem\cite{sarwar2001item}, and Simon Funk's SVD\cite{funk_netflix_2006} algorithms against two simple baseline algorithms.
+  Therefore we will compare UserUser\cite{resnick1994grouplens}, ItemItem\cite{sarwar2001item}, and Simon Funk's SVD\cite{funk_netflix_2006} against two simple baseline algorithms.
   More information about these algorithms can be found in table \ref{tbl:algo}.
@@
       ItemBaseline       & Item's average rating, with mild bayesian damping towards the global mean \cite{funk_netflix_2006}. \\\hline
-      UserItem\-Baseline & ItemBaseline adjusted by the user's average offset from the ItemBaseline. Mild mean damping is applied \cite{funk_netflix_2006}. \\\hline
+      UserItem\-Baseline & ItemBaseline adjusted by the user's average offset from the ItemBaseline. Mild bayesian damping is applied \cite{funk_netflix_2006}. \\\hline
       UserUser           & User based nearest neighbor collaborative filtering \cite{resnick1994grouplens} with a neighborhood size of 30 and ratings normalized by subtracting the UserItemBaseline score.
@@
   % citations and lenskit info
-  Each algorithm was configured and tuned similarly to Ekstrand's 2012 evaluation \cite{ekstrand2012recommenders}.
+  Each algorithm was tuned similarly to Ekstrand's 2012 evaluation \cite{ekstrand2012recommenders}.
   Since this evaluation was on a different dataset (MovieLens 10M vs. MovieLens 1M), we performed minor tuning from these parameters.
@@
   The metrics are listed in table \ref{tbl:metrics}.
+  We split our metrics into three groups roughly along our three sub research questions.
+
   \begin{table*}[ht!]
     \centering
-    \begin{tabular}{|p{7em}|p{20em}|p{4em}|p{18em}|}
+    \begin{tabular}{|p{7.5em}|p{21em}|p{7em}|p{12.5em}|}
+      \hline
+      Metric & Description & Range & Preferred direction\\\hline
       \hline
       \multicolumn{4}{|c|}{{\bf Accuracy and Ranking Metrics}} \\\hline
-      RMSE                   & Standard prediction accuracy metric \cite{handbook}. & 0 to 5 & Smaller is better. \\\hline
+      RMSE                   & Standard prediction accuracy metric \cite{handbook}. & 0 to 5 stars & Smaller is better. \\\hline
       nDCG                   & Standard ranking quality metric \cite{handbook}.     & 0 to 1       & Larger is better.  \\\hline
@@
       Precision@20           & Count of recommended items in the test set and rated 4.0 or higher\footnotemark[1] \cite{handbook}. & 0 to 1       & Larger is better.  \\\hline
-      MAP@20                 & Average of precision at each rank one through twenty \cite{manning2008introduction}. & 0 to 1 & Larger is better.\\\hline
-      Fallout@20             & Count of recommended items in the test set and rated 2.0 or lower\footnotemark[1] \cite{handbook}. & -1 to 1 & Smaller is better. \\\hline
-      MeanRating@20          & Average rating of recommended items that are in the test set.& 0 to 5 & Larger is better. \\\hline
-      RMSE@20                & RMSE computed only over those items that were recommended.& 0 to 5 & Smaller is better. \\\hline
+      MAP@20                 & Average of precision at each rank one through twenty \cite{manning2008introduction}.                & 0 to 1       & Larger is better.  \\\hline
+      Fallout@20             & Count of recommended items in the test set and rated 2.0 or lower\footnotemark[1] \cite{handbook}.  & 0 to 1       & Smaller is better. \\\hline
+      MeanRating@20          & Average rating of recommended items that are in the test set.                                       & 0 to 5 stars & Larger is better.  \\\hline
+      RMSE@20                & RMSE computed only over those items that were recommended.                                          & 0 to 5 stars & Smaller is better. \\\hline
       \hline
-      \multicolumn{4}{|c|}{{\bf Nonmonotonic Recommender Metrics}} \\\hline
-      SeenItems@20           & Count of recommended items in the test set. & 0 to 20 & Not well understood. \newline Larger is \emph{probably} better within reason. \\\hline
-      Average\-Popularity@20 & The average number of users in the training set that have seen recommended items \cite{zieglerDiversity}.& 1 to 6000 & Not well understood.
-\newline Smaller is \emph{probably} better within reason.
-\\\hline
-      AILS@20                & The average pairwise similarity between recommended items \cite{zieglerDiversity}.
-& 0 to 1 & Not well understood.
-\newline  Smaller is \emph{probably} better within reason.
-\\\hline
-      Spread@20              & The Shannon's entropy of the distribution of recommended items for users in the test set.
-& 0 to 12 & Not well understood.
-\newline Larger is \emph{probably} better within reason.
-\\\hline
+      \multicolumn{4}{|c|}{{\bf Non-monotonic Recommender Metrics}} \\\hline
+      SeenItems@20           & Count of recommended items in the test set. & 0 to 20 items & Moderate values are better. \\\hline
+      Average\-Popularity@20 & The average number of users in the training set that have seen recommended items \cite{zieglerDiversity}.& 1 to 6000 users & Moderate values are better. \\\hline
+      AILS@20                & The average pairwise similarity between recommended items \cite{zieglerDiversity}.                       & -1 to 1         & Moderate values are better. \\\hline
+      Spread@20              & The Shannon's entropy of the distribution of recommended items for users in the test set.                & 0 to 12 bits    & Moderate values are better. \\\hline
     \end{tabular}
@@
   \end{table*}
+
+  \paragraph{Accuracy and Ranking Metrics}
+  {\bf RMSE} and {\bf nDCG} evaluate the recommender based on its prediction for each item.
+  These evaluate the accuracy of the predictions and the ability of the recommender to rank the list of all items.
+  We find that these two metrics perform quite similarly in our evaluation.
+
   \footnotetext[1]{We also tried 5.0 and 1.0 as cutoffs respectively, and found no meaningful difference.}\addtocounter{footnote}{1}
 
-  We split our metrics into three groups roughly along our three sub research questions.
-
-  \paragraph{Accuracy and Ranking Metrics}
-  The first two metrics evaluate the recommender based on its predictions for each item.
-  These evaluate, the accuracy of the predictions, and the ability of the recommender to rank the list of all items.
-  We find that these two metrics perform quite similarly in our evaluation.
-
   \paragraph{Monotonic Recommendation Metrics}
+  %TODO: kenny comment: is 20 standard size for these?
+  %TODO: daniel comment: check for consistency with the user vs. users.
   These metrics evaluate the top 20 recommendations produced by the recommender.
-  As the UserItemBaseline and ItemBaseline make the same recommendations for all users, we will only report results for the ItemBaseline here.
   The choice of 20 items for the evaluations was arbitrary, testing with other values showed similar results, except on one metric.
@@
   These metrics are useful because they have a well defined relation to the satisfaction of the user.
-  As the metric grows larger, or smaller (depending on metric), we expect the recommendations to satisfy users more.
+  As the metric grows larger (or smaller, depending on metric) we expect the recommendations to satisfy users more.
   This makes these metrics easy to apply.
 
-  Our first three metrics (Precision, MAP, and Fallout) are standard information retrieval metrics.
+  {\bf Precision@20}, {\bf MAP@20}, and {\bf Fallout@20} are standard information retrieval metrics.
   As we show later, we found these metrics to not be useful for this evaluation.
-  Therefore, to help understand the quality of our recommendations, we also include MeanRating@20, and RMSE@20.
+  Therefore, to help understand the quality of our recommendations, we also include {\bf MeanRating@20}, and {\bf RMSE@20}.
 
-  MeanRatings@20, and RMSE@20 both try to establish how much the user likes the recommendations.
-  MeanRatings@20 tells us directly how much users tend to like recommended movies.
-  RMSE@20 accompanies this and lets us know if the recommended items are likely to be incorrectly predicted, which can explain why users may or may not like their recommendations.
-  As it helps us interpret these two metrics we will also discuss SeenItems@20 with these metrics, despite SeenItems@20 belonging to the next group of metrics.
+  {\bf MeanRating@20} tries to establish how much users like their recommendations by measuring users average satisfaction with recommended items based on withheld ratings.
+  {\bf RMSE@20} accompanies this metric and lets us know if the recommended items are likely to be incorrectly predicted, which can explain why users may or may not like their recommendations.
 
-  \paragraph{Nonmonotonic Recommendation Metrics}
-  Our final four metrics are all nonmonotonic measures of the user's recommendations.
+  \paragraph{Non-monotonic Recommendation Metrics}
+  {\bf SeenItems@20}, {\bf AveragePopularity@20}, {\bf AILS@20}, and {\bf Spread@20} are all non-monotonic measures of the user's recommendations.
   These metrics are harder to apply than the monotonic measures of recommendation quality.
@@
   We use these metrics as they can help us understand important properties of how users perceive their recommendations.
-  For example, as mentioned in the introduction how many of the recommended items the user has seen (estimated by SeenItems@20) can have a major impact on how much users trust the recommendations.
+  As mentioned in the introduction the number of recommended items that the user has seen (estimated by {\bf SeenItems@20}) can have a major impact on how much users trust the recommendations.
   If an algorithm recommends too few items that the user has seen then we expect the user will have trouble evaluating if the recommendations are trustworthy.
-  That said, if too many items in the recommendations have already been seen, the user may feel that the recommendations are not interesting or useful.
+  Too many, and the user may feel that the recommendations are not interesting or useful.
   Unfortunately, without more research on these metrics to establish a mapping between user satisfaction and metric values we will have to rely on educated guesses to establish which values for these metrics are too extreme.
 
-  The average popularity metric is a crude metric of how novel the recommendations are to the user \cite{zieglerDiversity}.
+  {\bf AveragePopularity@20} is a crude metric of how novel the recommendations are to the user \cite{zieglerDiversity}.
   Broadly speaking, we expect users will prefer recommendation lists containing more novel (less popular) items.
@@
   At some point, a list cannot become more diverse without including items that do not interest the user.
-  We will measure diversity with the AILS@20 metric.
+  We will measure diversity with the {\bf AILS@20} metric.
 
-  The AILS@20 metric is based on the ILS metric introduced by ziegler et al \cite{zieglerDiversity}.
+  The {\bf AILS@20} metric is based on the Intra-List Similarity (ILS) metric introduced by ziegler et al \cite{zieglerDiversity}.
   AILS@N is simply a rescaling of ILS to be scale free with regards to N, this was done to make the metric easier to interpret, by putting it on the same scale as similarity values.
@@
 
-  The Spread@20 metric is designed to measure how well the recommender spreads its attention across many items.
+  The {\bf Spread@20} metric is designed to measure how well the recommender spreads its attention across many items.
   We expect that algorithms with a good understanding of its users will be able to recommend different items to different users.
@@
   Let $P(I) = C(I) / \sum_I C(I)$ be the probability that a random recommendation is for item $I$.
-  Finally we define spread as the Shannon's entropy of $P(I)$, $Spread = -\sum_I P(I) \log(P(I))$.
+  We define spread as the Shannon's entropy of $P(I)$, $Spread = -\sum_I P(I) \log(P(I))$.
 
@@
   The UserItemBaseline should only differ from the ItemBaseline for metrics that use the prediction value.
-  For all other metrics these two should provide the same responses.
+  For all other metrics these two provide the same responses, and we will only report results for the ItemBaseline.
   As the UserItemBaseline is still quite a simple baseline, we expect that the non-baseline algorithm will beat this baseline as well.
-  That said, for small rating values its possible that the improvements over this baseline will be very small.
+  That said, for small rating values it is possible that the improvements over this baseline will be very small.
 
@@
   To the best of our knowledge, at the time this data was collected the new user survey preferred items that were very popular.
-  Therefore will randomize the order of the ratings to try to avoid measuring the effect of these initial ratings, and instead focus on the algorithm's performance on normal ratings.
+  Therefore we will randomize the order of the ratings to try to avoid measuring the effect of these initial ratings, and instead focus on the algorithm's performance on normal ratings.
 
@@
   % describe normal crossfold solution
-  The natural way to perform this evaluation is with a modified user crossfolding strategy.
-  The normal 5 fold cross validation strategy splits the users into five groups.
-  In each of five folds all ratings from four of the groups will be kept in the training set, and then either some constant number or percent of ratings from each user in the fifth group will be set aside for the test set.
+  The natural way to perform this evaluation is with a modified user crossfold strategy.
+  The normal five-fold cross validation strategy splits the users into five groups.
+  In each of five folds all ratings from four of the groups will be kept in the training set, and then some subset of the ratings from each user in the fifth group will be set aside for the test set.
   The selection of a test set can be random to avoid any possible ordering effects in the data due to a new user rating strategy.
   To understand how metrics change with a different number of users we can simply generate separate crossfolds for each simulated profile size.
-  At each crossfold, rather than only taking a constant number of ratings for the test set, we can instead only keep a constant number of ratings for the training set.
+  At each crossfold we include only a constant number of ratings for each test user into the training set, ensuring that all test users have the specified profiles size.
 
   % describe issues with solution
-  Unfortunately, we found that this strategy can causes biases in our evaluation.
-  In particular, this strategy led to the size of the test sets to change as a function of the simulated profile size.
+  Unfortunately, we found that this strategy causes biases in our evaluation.
+  In particular, this strategy led to the size of the test sets changing as a function of the simulated profile size.
   As we kept more ratings for the training set, we would have fewer ratings in the test set.
-  This turned out to bias our results, as some metrics, such as nDCG, proved to be sensitive to the number of items in the test set.
+  This turned out to bias our results; some metrics, such as nDCG, proved to be sensitive to the number of items in the test set.
 
   % describe our crossfold then subsample solution.
-  Our solution to this bias is to subsample an existing crossfold.
-  Therefore, to generate training and test sets for different number of ratings from 1 to 19, we first generate a set of test and training splits so that test users have 19 training items using the crossfold strategy outlined above.
+  Our solution to this bias is to subsample from an existing crossfold.
+  To generate training and test sets for different number of ratings from 1 to 19, we first generate a set of test and training splits so that test users have 19 training items using the crossfold strategy outlined above.
   Then, to generate any other number of ratings, we randomly downsample the 19 item training set.
@@
 
-  Using the methodology described in section \ref{sec:methodology} we generate plots for each metric showing how the five algorithms perform new users, versus simulated profile size.
+  Using the methodology described in section \ref{sec:methodology} we generate plots for each metric showing how the five algorithms perform versus simulated profile size.
 
 \subsection*{Accuracy and Ranking Metrics}
+
+\vspace{-1.5em}
 
@@
   UserUser, and SVD both behaved essentially as expected.
-  SVD shows no improvement with only one rating, and UserUser actually gets worse with the first rating, but other than that, these algorithms outperform ItemBaseline.
+  SVD shows no improvement with only one rating, and UserUser actually gets worse with the first rating.
+  Other than that, these algorithms outperform ItemBaseline.
   Interestingly, UserItemBaseline outperforms both algorithms until around 8 ratings, at which point both algorithms start to perform better.
@@
   This trend is a result of the limited coverage of the ItemItem algorithm.
-  If you consider the accuracy of RMSE on only those items where it makes a personalized prediction it does show a monotonic increase in accuracy similar to the other algorithms.
+  Consider the accuracy of RMSE on only those items where it makes a personalized prediction, it shows a monotonic increase in accuracy similar to the other algorithms.
   However, for small rating counts ItemItem can only predict for some percent of the item (around 60\% for users with two ratings, and around 75\% for users with four ratings).
@@
 
+\vspace{-1.5em}
+
 \begin{figure*}[ht!]
@@
 \end{figure*}
+%
 \paragraph{MAP@20, Precision@20, Fallout@20}
-  Figure \ref{fig:map} shows MAP@20, precision@20, and fallout@20.
+  Figure \ref{fig:map} shows MAP@20, Precision@20, and Fallout@20.
   All three metrics show essentially the same trend, UserUser gets the lowest score, the baseline get the highest score.
-  On both MAP@20 and precision@20 SVD gets a higher score than ItemItem throughout, performing almost as highly as the baseline.
-  On fallout@20 ItemItem and SVD perform about as well as each other, with SVD showing a very slight lead.
+  On both MAP@20 and Precision@20, SVD gets a higher score than ItemItem throughout, performing almost as well as the baseline.
+  On Fallout@20 ItemItem and SVD perform about as well as each other, with SVD showing a very slight lead.
   We also tested recall and mean reciprocal rank metrics, but found them to exhibit the same trend, so they were not explored further.
 
-  This result is rather strange as MAP and Precision are metrics where larger scores imply better performance, whereas fallout is a metric where larger scores imply poor performance.
+  This result is rather strange as MAP and Precision are metrics where larger scores imply better performance, whereas Fallout is a metric where larger scores imply poor performance.
   Therefore we would not expect to see rank equivalent behavior from these metrics.
@@
   Looking forward to the average popularity (figure \ref{fig:pop}) of recommended items, we see that the average popularity of the recommendations closely replicates the previous three plots.
-  Therefore we will discard these metrics, and consider less biased metrics to help us analyze the quality of the recommendations.
+  As a result of this, we discard these metrics, and consider less biased metrics to help us analyze the quality of the recommendations.
 
@@
   \paragraph{SeenItems@20}
+  We address this non-monotonic metric out of turn here as it provides important context for the next two metrics.
   The first graph in figure \ref{fig:topN.rmse} shows the average number of items in the top 20 recommendations for each algorithm that were also in the test set.
-  We address this metric out of turn here as it provides important context for the next two metrics.
   The average number of test set items in recommendations shows a similar trend to the MAP and precision metrics, with UserUser getting by far the fewest seen items, and the baseline getting the most.
   As UserUser gets less than one seen item on average we found that UserUser was not appropriate for the MeanRating@20 and RMSE@20 metrics.
-  In particular, we found that the UserUser results for these metrics was highly sensitive to recommendation list size.
-  Because of this issue we will not include UserUser in our next evaluations and consider the next two metrics inconclusive for UserUser.
-  That said, we consider the small number of seen items itself a bad result for UserUser.
+  In particular, we found that the UserUser results for these metrics are highly sensitive to recommendation list size.
+  Because of this issue we will not include UserUser in our next two metrics and consider the next two metrics inconclusive for UserUser.
+  That said, we consider the small number of seen items to be a bad result for UserUser.
   Put simply, if we cannot reliably estimate if the recommendations are reasonable, how can we expect a user to do so?
@@
 
-\subsection*{Nonmonotonic Recommender Quality Metrics}
+\subsection*{Non-monotonic Recommender Quality Metrics}
   % Other recommender metrics
-
+\vspace{-1.5em}
 \begin{figure*}[ht!]
@@
   \includegraphics[width=\linewidth]{../lenskit/output/ekstrandTuned20/popdiv.pdf}
-  \caption{Nonmonotonic Recommendation Quality Metrics}
+  \caption{Non-monotonic Recommendation Quality Metrics}
   \label{fig:pop}
 \end{figure*}
-
+%
   \paragraph{AveragePopularity@20}
@@
 
-  The high level results of our evaluation are summarized in table \ref{tbl:results}
+  The high level results of our evaluation are summarized in table \ref{tbl:results}.
   Overall we see that SVD seems to be the best performing algorithm for cold start use.
@@
   While some novelty is good, we believe that UserUser goes too far with obscure movies in its recommendations, and therefore does not make good recommendations.
-  Therefore we conclude that UserUser is not well suited for new users unless only its predictions are used.
-  Exploring corrections for this issue is left as future work.
+  Therefore we conclude that UserUser alone is not well suited for new users.
+  It is possible that careful tuning or hybridizing with another algorithm may help UserUser perform well for new users.
+  Exploring this issue further is left as future work.
 
@@
   The switching point between algorithms can be determined with our methodology.
-  The only downside of this approach is that the baseline performed poorly on our nonmonotonic metrics, which may effect users as they first join the system.
+  The only downside of this approach is that the baseline performed poorly on our non-monotonic metrics, which may effect users as they first join the system.
 
@@
   The downside of this, however, could be the higher popularity and lower spread seen by the SVD algorithm.
-  The same regularization that helps avoid making mistakes due to too little data, may also prevent the algorithm from taking occasional risks on more obscure movies.
+  The same regularization that helps avoid making mistakes due to too little data may also prevent the algorithm from taking occasional risks on more obscure movies.
 
@@
   Unfortunately, this means that we do not know if there are differences between users who do and do not survive in the system.
-  This is a limitation of our methodology, to maintain a constant test set throughout the evaluation we are not able to evaluate users who have less ratings than the maximum amount we wish to study.
+  This is a limitation of our methodology, to maintain a constant test set throughout the evaluation we are not able to evaluate users who have less ratings than the maximum amount of ratings we wish to study.
   More work needs to be done to understand users who only  make one or two ratings to discover why they leave, and how they differ from other users of our systems.
@@
   Work should be done to understand the differences between these algorithms, and what effect extra information has on the recommendation process.
-  Using our methodology should might be possible to measure the effect that extra information about the user has on recommendation, and to compare it the effect of extra initial ratings.
+  Using our methodology it should be possible to measure the effect that extra information about the user has on recommendation, and to compare it against the effect of extra initial ratings.
     % algorithm limitations
@@
   We would like to specially that Michael Ekstrand for his  important work on the Lenskit recommendation framework.
-
-  This work is supported by NSF grant TODO:
+  This work is supported by NSF grants IIS1017697 and IIS0808692.
+  %netfishing and volunteers
+  %check with Joe about netfishing.
 

### 2014-05-07 lenskit/src/eval/chart.R

@@ -1,11 +1,14 @@
 # Create a chart comparing the algorithms
 
-cp = c("#e41a1c","#377eb8","#4daf4a","#984ea3","#ff7f00")
+cp = c("#4daf4a","#377eb8","#e41a1c","#984ea3","#ff7f00")
 
 library("ggplot2")
 library(reshape2)
 
 all.data = read.csv("eval-user.csv")
+
+nonMetrics = c("Algorithm","Retain", "DataSet", "User", "Partition", "NGood", "TestEvents", "TrainEvents", "NAttempted", "RatingEntropy")
+metrics = setdiff(names(all.data),nonMetrics)
 
 all.data.melt = melt(all.data, c("Algorithm", "Retain", "User"), metrics)
 data.melt = aggregate(value~Algorithm+Retain+variable, data=all.data.melt, mean)
@@ -28,8 +31,8 @@ data.melt$Algorithm = ordered(data.melt$Algorithm, c("ItemItem", "UserUser", "sv
 vars = c("RMSE", "nDCG")
 data.melt.sub = subset(data.melt, variable %in% vars)
 data.melt.sub = subset(data.melt.sub, Algorithm != "UserItemBaseline" | variable == "RMSE")
-plot = ggplot(data.melt.sub, aes(x=Retain, y=value, color=Algorithm, shape=Algorithm))+geom_point()+geom_line()+scale_x_continuous(breaks=4*(0:5))+facet_wrap(~variable, scales="free_y", nrow=1)+scale_color_manual(values=cp)+theme_minimal()+xlab("Simulated Profile Size")+theme(legend.position="bottom")
-pdf("accuracy.pdf", width=8.5*2/2.5, height=3.5)
+plot = ggplot(data.melt.sub, aes(x=Retain, y=value, color=Algorithm, shape=Algorithm))+geom_point()+geom_line()+scale_x_continuous(breaks=4*(0:5))+facet_wrap(~variable, scales="free_y", nrow=1)+scale_color_manual(values=cp)+theme_minimal()+xlab("Simulated Profile Size")+theme(legend.position="bottom") + theme(plot.margin = unit(c(0,0,0,0), "cm"))
+pdf("accuracy.pdf", width=8.5*2/2.5, height=3.2)
 print(plot)
 dev.off()
 
@@ -37,8 +40,8 @@ dev.off()
 vars = c("Precision@20", "MAP@20", "Fallout@20")
 data.melt.sub = subset(data.melt, variable %in% vars)
[…]

### 2014-05-07 paper/main.tex
final submission

@@
 \pdfpageheight = 11in
+
+\widowpenalty=3000
 
@@
   As we show later, we found these metrics to not be useful for this evaluation.
-  Therefore, to help understand the quality of our recommendations, we also include {\bf MeanRating@20}, and {\bf RMSE@20}.
+  Therefore, to help understand the quality of our recommendations, we also include MeanRating@20, and RMSE@20.
 
@@
 
+\footnotetext[3]{We do not plot UserItemBaseline here as it is equivalent to ItemBaseline on this metric.}
+
   % Recommendation quality metrics
@@
   Again, we find that ItemItem performs poorly until quite a few ratings are added, and does worse than baseline throughout.
+
+\footnotetext[4]{We do not plot UserUser here as it cannot be reliably measured for this metric.}
 
@@
   The authors would like to express their gratitude towards our colleagues at GroupLens research for their feedback and support while developing this work.
-  We would like to specially that Michael Ekstrand for his  important work on the Lenskit recommendation framework.
+  We would like to specially thank Michael Ekstrand for his  important work on the Lenskit recommendation framework.
   This work is supported by NSF grants IIS1017697 and IIS0808692.

### 2014-05-07 lenskit/src/eval/chart.R

@@ -15,20 +15,22 @@ data.melt = aggregate(value~Algorithm+Retain+variable, data=all.data.melt, mean)
 
 # redo some names
 data.melt$variable = as.character(data.melt$variable)
-data.melt$variable[data.melt$variable == "MAP"] = "MAP@20"
-data.melt$variable[data.melt$variable == "Precision"] = "Precision@20"
-data.melt$variable[data.melt$variable == "Precision.fallout"] = "Fallout@20"
-data.melt$variable[data.melt$variable == "TopN.RMSE.seenItems"] = "SeenItems@20"
+data.melt$variable[data.melt$variable == "MAP"] = "MAP@20\nUserItemBaseline is equivalent to ItemBaseline on this metric, and is not plotted"
+data.melt$variable[data.melt$variable == "Precision"] = "Precision@20\nUserItemBaseline is equivalent to ItemBaseline on this metric, and is not plotted"
+data.melt$variable[data.melt$variable == "Precision.fallout"] = "Fallout@20\nUserItemBaseline is equivalent to ItemBaseline on this metric, and is not plotted"
+data.melt$variable[data.melt$variable == "TopN.RMSE.seenItems"] = "SeenItems@20\nUserItemBaseline is equivalent to ItemBaseline on this metric, and is not plotted"
 data.melt$variable[data.melt$variable == "TopN.RMSE"] = "RMSE@20"
-data.melt$variable[data.melt$variable == "TopN.Ave.Rat"] = "MeanRating@20"
-data.melt$variable[data.melt$variable == "TopN.MeanPopularity"] = "AveragePopularity@20"
-data.melt$variable[data.melt$variable == "diversity"] = "AILS@20"
-data.melt$variable = ordered(data.melt$variable, levels = c("Coverage", "RMSE", "nDCG", "Precision@20", "MAP@20", "Fallout@20", "SeenItems@20", "MeanRating@20", "RMSE@20", "AveragePopularity@20", "AILS@20", "TestTime","MAE","MutualInformation","PredicitonEntropy","ItemScorer.NAttempted","ItemScorer.NGood","ItemScorer.Coverage","TopN.ActualLength", "Recall.fallout", "Recall"))
+data.melt$variable[data.melt$variable == "TopN.Ave.Rat"] = "MeanRating@20\nUserItemBaseline is equivalent to ItemBaseline on this metric, and is not plotted"
[…]

### 2014-05-07 lenskit/src/eval/chart.R
oops, forgot to save a file

@@ -15,22 +15,21 @@ data.melt = aggregate(value~Algorithm+Retain+variable, data=all.data.melt, mean)
 
 # redo some names
 data.melt$variable = as.character(data.melt$variable)
-data.melt$variable[data.melt$variable == "MAP"] = "MAP@20\nUserItemBaseline is equivalent to ItemBaseline on this metric, and is not plotted"
-data.melt$variable[data.melt$variable == "Precision"] = "Precision@20\nUserItemBaseline is equivalent to ItemBaseline on this metric, and is not plotted"
-data.melt$variable[data.melt$variable == "Precision.fallout"] = "Fallout@20\nUserItemBaseline is equivalent to ItemBaseline on this metric, and is not plotted"
-data.melt$variable[data.melt$variable == "TopN.RMSE.seenItems"] = "SeenItems@20\nUserItemBaseline is equivalent to ItemBaseline on this metric, and is not plotted"
-data.melt$variable[data.melt$variable == "TopN.RMSE"] = "RMSE@20"
-data.melt$variable[data.melt$variable == "TopN.Ave.Rat"] = "MeanRating@20\nUserItemBaseline is equivalent to ItemBaseline on this metric, and is not plotted"
-data.melt$variable[data.melt$variable == "TopN.MeanPopularity"] = "AveragePopularity@20\nUserItemBaseline is equivalent to ItemBaseline on this metric, and is not plotted"
-data.melt$variable[data.melt$variable == "diversity"] = "AILS@20\nUserItemBaseline is equivalent to ItemBaseline on this metric, and is not plotted"
-data.melt$variable[data.melt$variable == "nDCG"] = "nDCG\nUserItemBaseline is equivalent to ItemBaseline on this metric, and is not plotted"
-
[…]

### 2014-07-21 paper/main.tex
first pass for photo-ready. Still needs work

@@
   While this makes our methodology unsuited for applying temporal metrics, it allows us to separate an algorithm's behavior from how ratings are collected for new users.
+
 Therefore, we feel that our methodology is more suitable to answer our research question.
@@
   Therefore we will compare UserUser\cite{resnick1994grouplens}, ItemItem\cite{sarwar2001item}, and Simon Funk's SVD\cite{funk_netflix_2006} against two simple baseline algorithms.
+
+  Due to the low profile sizes involved in our evaluation, not every algorithm was able to make predictions for every user-item pair.
+  In these cases, we use the UserItemBaseline prediction as a fallback.
+  This ensured that our prediction based metrics were always evaluated over the same number items.
+  We found this unnecessary for recommendations as each algorithm was able to make a complete recommendation with only one rating.
+  Therefore our algorithms recommend only items that they can predict.
   More information about these algorithms can be found in table \ref{tbl:algo}.
@@
   % citations and lenskit info
-  Each algorithm was tuned similarly to Ekstrand's 2012 evaluation \cite{ekstrand2012recommenders}.
-  Since this evaluation was on a different dataset (MovieLens 10M vs. MovieLens 1M), we performed minor tuning from these parameters.
-  Most notably, we found that the previous damping parameter used for the baselines was too large.
+  Each algorithm was tuned based off of Ekstrand's 2012 evaluation \cite{ekstrand2012recommenders}.
+  Since this evaluation is on a different dataset (MovieLens 10M vs. MovieLens 1M), we performed minor tuning using the Ekstrand's values as a starting point.
+  We tuned based on RMSE performance.
+  While tuning  we found that the previous damping parameter used for the baselines to be too large.
   We found that a value of 5 gave much better results for new user recommendation.
   Our key results are not sensitive to minor changes of algorithm parameters.
-  %TODO: carefully run this wording by joe.
 
   \subsection*{Metrics}
-
-  We use a total of eleven metrics.
-  In our evaluation we found that some of these metrics are redundant, and future work should be able to use a much smaller list of metrics.
-  The metrics are listed in table \ref{tbl:metrics}.
-  We split our metrics into three groups roughly along our three sub research questions.
 
@@
       MAP@20                 & Average of precision at each rank one through twenty \cite{manning2008introduction}.                  & 0 to 1       & Larger is better.  \\\hline
-      Fallout@20             & Count of recommended items in the test set and rated 2.0 or lower\footnotemark[1] \cite{handbook}.  & 0 to 1       & Smaller is better. \\\hline
+      Fallout@20             & Count of recommended items in the test set and rated 2.0 or lower\footnotemark[1] \cite{TenIsEnough}. & 0 to 1       & Smaller is better. \\\hline
       MeanRating@20          & Average rating of recommended items that are in the test set.                                         & 0 to 5 stars & Larger is better.  \\\hline
@@
 
+  We present results from eleven metrics.
+  As has been seen in previous studies \cite{herlockerEvaluating} we find that some of are metrics are redundant.
+  Future work should be able to use a smaller list of metrics.
+  The metrics are listed in table \ref{tbl:metrics}.
+  We split our metrics into three groups roughly along our three sub research questions.
+
   \paragraph{Accuracy and Ranking Metrics}
@@
   \paragraph{Monotonic Recommendation Metrics}
-  %TODO: kenny comment: is 20 standard size for these?
   %TODO: daniel comment: check for consistency with the user vs. users.
@@
   This issue will be addressed later in the results section.
+  We refer to these as monotonic metrics as there is a clear good direction for the metric.
 
-  These metrics are useful because they have a well defined relation to the satisfaction of the user.
-  As the metric grows larger (or smaller, depending on metric) we expect the recommendations to satisfy users more.
-  This makes these metrics easy to apply.
-
-  {\bf Precision@20}, {\bf MAP@20}, and {\bf Fallout@20} are standard information retrieval metrics.
-  As we show later, we found these metrics to not be useful for this evaluation.
+  {\bf Precision@20}, {\bf MAP@20}, and {\bf Fallout@20} are standard information retrieval metrics which measure the quality of the recommendations.
+  As we show later, these metrics were not useful for our evaluation.
   Therefore, to help understand the quality of our recommendations, we also include MeanRating@20, and RMSE@20.
@@
   {\bf MeanRating@20} tries to establish how much users like their recommendations by measuring users average satisfaction with recommended items based on withheld ratings.
-  {\bf RMSE@20} accompanies this metric and lets us know if the recommended items are likely to be incorrectly predicted, which can explain why users may or may not like their recommendations.
+  {\bf RMSE@20} accompanies this metric and lets us know if the recommended items are likely to be incorrectly predicted. 
+  This can help us understand why an algorithm scores well at MeanRatings@20.
 
@@
   These metrics are harder to apply than the monotonic measures of recommendation quality.
-  Unlike the previous metrics, we expect that both of the extremes of these metrics relate to a negative user experience.
+  Unlike the previous metrics, we expect that users will be most satisfied by moderate values of these metrics, and that both extremes can lead to a negative user experience.
 
@@
   Broadly speaking, we expect users will prefer recommendation lists containing more novel (less popular) items.
-  However, if the recommended items are too unpopular we risk that the user will not have heard of the items, which could make the user doubt the recommendations, hurting the user experience.
+  However, if the recommended items are too novel the user is unlikely to have heard of the items and will have no point of reference in understanding their recommendations.
 
@@
 
-  The {\bf AILS@20} metric is based on the Intra-List Similarity (ILS) metric introduced by ziegler et al \cite{zieglerDiversity}.
+  The {\bf AILS@20} metric is based on the Intra-List Similarity (ILS) metric introduced by Ziegler et al \cite{zieglerDiversity}.
   AILS@N is simply a rescaling of ILS to be scale free with regards to N, this was done to make the metric easier to interpret, by putting it on the same scale as similarity values.
@@
   When Spread is small, the recommender is focusing on a small set of items which it recommends to every user.
+  In this way spread is related to item coverage metrics.
+  Spread however looks at how often each item is recommended, not just whether or not it is recommended.
 
@@
   Since the ItemBaseline is such a simple algorithm we expect that any reasonable algorithm should be able to do better than it.
-  Furthermore, we expect the improvement over the ItemBaseline to be gradual, with larger improvements only happening with more data.
+  That said, we expect the improvement over the ItemBaseline to be gradual, with larger improvements only happening with more data.
 
@@
   For all other metrics these two provide the same responses, and we will only report results for the ItemBaseline.
-  As the UserItemBaseline is still quite a simple baseline, we expect that the non-baseline algorithm will beat this baseline as well.
-  That said, for small rating values it is possible that the improvements over this baseline will be very small.
+  As the UserItemBaseline is still quite a simple baseline, we expect that the non-baseline algorithm will beat this baseline as well given enough ratings.
 
   We expect that the baseline algorithms will provide the most popular recommendations, with the least diversity, and the least spread.
-  We expect that any personalized algorithm should be able to make more nuanced recommendations than simply recommending the best rated items, which should translate as gains on these metrics.
-  Furthermore, we expect that as the algorithm learns more about the user it should decrease in popularity, and increase in diversity, and spread, as it becomes able to make more personal and nuanced recommendations.
+  Any personalized algorithm should be able to make more nuanced recommendations than simply recommending the best rated items, which should translate as gains on these metrics.
+  Furthermore, we expect that as an algorithm learns more about the user it should decrease in popularity, and increase in diversity, and spread, as it becomes able to make more personal and nuanced recommendations.
 
@@
       \hline
-      Algorithm & Prediction accuracy & Recommendation quality      & Other properties \\\hline
+      {\bf Algorithm} & {\bf Prediction accuracy} & {\bf Recommendation quality} & {\bf Other properties} \\\hline
       ItemItem        & Poor                      & Poor                         & Good                   \\\hline
@@
 \section{Conclusion}
-
-  Our work shows that a multiple metric approach to understanding recommender behavior can provide useful insights into the differences between algorithms for new users.
-  Our methodology presents a list of metrics, and an evaluation framework that can be used to perform further evaluations in the future.
   We find that recommending for new users is hard and none of our algorithms can reliably beat a simple baseline for users with very few ratings.
@@
   This result can be useful both in designing new recommender systems, and in designing novel recommendation algorithms for new users.
+  We also present a methodology and list of metrics for understanding recommender behavior that can be used to extend our results to other algorithms and datasets.
 

### 2014-07-22 paper/main.tex
post second pass.

@@
 
-  While past work has covered the new user rating survey from many different angles, it has not carefully evaluated which algorithm is best at recommending to new users.
+  While past work has carefully covered the new user rating survey, it has not carefully evaluated which algorithm is best at recommending to new users.
   The new user survey work sidesteps this issue by assuming that the user will receive no personalization until the survey is complete.
@@
 
-  Fortunately, there has been a growing trend in recommender system research to perform multiple metric evaluations of recommender systems.
+  The ideal way to study this issue would be a live study with real people who are new to a recommender system.
+  Users could be randomly assigned an algorithm and then surveyed after a random number of ratings to establish what they thought of the new user experience. 
+  Unfortunately, there are many algorithms and new users are a scarce resource.
+
+  Fortunately, there has been a growing trend in recommender system research to perform multiple metric offline evaluations of recommender systems.
   This work comes out of a pushback against early recommender evaluations which only evaluated recommenders in terms of their predictive accuracy \cite{accurateIsNotEnough}.
@@
 
-  To answer our research questions we will evaluate algorithms in an offline evaluation designed to simulate users as they join the system.
-  We do this by adapting the standard train and test methodology to test existing users with a specific number of ratings.
-  We run this approach for various profile sizes, and then plot the results.
+  Ideally we would measure new user experience with a live survey on users who are just joining a recommender service.
+  Because new users are a scarce resource we opt to first explore the new user experience in an offline setting.
+  This allows us to test many more profile sizes and algorithms much faster than an online study.
+  We will adapt the standard train and test methodology so that test users have only a specified number of training ratings.
+  By running this approach for various profile sizes we can generate plots of performance against profile size.
   This lets us understand both how the algorithms compare on these metrics, and how the algorithm performance changes with the number of ratings provided.
@@
   While this makes our methodology unsuited for applying temporal metrics, it allows us to separate an algorithm's behavior from how ratings are collected for new users.
-
-Therefore, we feel that our methodology is more suitable to answer our research question.
+  We feel that this is better given the limitations of our dataset, and our focus on the algorithm, rather than how ratings are collected.
 
@@
   In these cases, we use the UserItemBaseline prediction as a fallback.
-  This ensured that our prediction based metrics were always evaluated over the same number items.
+  This ensured that our prediction based metrics were always evaluated over the same number of items.
   We found this unnecessary for recommendations as each algorithm was able to make a complete recommendation with only one rating.
-  Therefore our algorithms recommend only items that they can predict.
+  Therefore we form recommendations as the top 20 items that the algorithm was able to predict which were not in the training set.
   More information about these algorithms can be found in table \ref{tbl:algo}.
@@
   \end{table}
-
-  % evaluation and code to run these algortihms available at.
-  % citations and lenskit info
-  Each algorithm was tuned based off of Ekstrand's 2012 evaluation \cite{ekstrand2012recommenders}.
-  Since this evaluation is on a different dataset (MovieLens 10M vs. MovieLens 1M), we performed minor tuning using the Ekstrand's values as a starting point.
-  We tuned based on RMSE performance.
-  While tuning  we found that the previous damping parameter used for the baselines to be too large.
-  We found that a value of 5 gave much better results for new user recommendation.
-  Our key results are not sensitive to minor changes of algorithm parameters.
-
-  \subsection*{Metrics}
 
@@
 
+  % evaluation and code to run these algortihms available at.
+  % citations and lenskit info
+  Each algorithm was tuned based off of Ekstrand's 2012 evaluation \cite{ekstrand2012recommenders}.
+  Since this evaluation is on a different dataset (MovieLens 10M vs. MovieLens 1M), we performed minor tuning using the Ekstrand's values as a starting point.
+  We tuned based on RMSE performance.
+  While tuning, we found that the previous damping parameter used for the baselines to be too large.
+  We found that a value of 5 gave much better results for new user recommendation.
+  Our key results are not sensitive to minor changes of algorithm parameters.
+
+  \subsection*{Metrics}
+
   We present results from eleven metrics.
@@
   Future work should be able to use a smaller list of metrics.
-  The metrics are listed in table \ref{tbl:metrics}.
+  The metrics are described in table \ref{tbl:metrics}.
   We split our metrics into three groups roughly along our three sub research questions.
@@
   {\bf RMSE@20} accompanies this metric and lets us know if the recommended items are likely to be incorrectly predicted. 
-  This can help us understand why an algorithm scores well at MeanRatings@20.
+  This can help us understand why an algorithm scores well on MeanRatings@20.
 
@@
   To the best of our knowledge, at the time this data was collected the new user survey preferred items that were very popular.
-  Therefore we will randomize the order of the ratings to try to avoid measuring the effect of these initial ratings, and instead focus on the algorithm's performance on normal ratings.
+  We will randomize the order of the ratings to try to avoid measuring the effect of these initial ratings, and instead focus on the algorithm's performance on normal ratings.
 
   Each user in the dataset has at least 20 ratings.
-  Therefore we will evaluate new users with up to 19 ratings.
+  Therefore we will evaluate user profiles with up to 19 ratings.
   This will allow us to ensure that each user in the dataset has at least one test rating.
@@
   The selection of a test set can be random to avoid any possible ordering effects in the data due to a new user rating strategy.
-  To understand how metrics change with a different number of users we can simply generate separate crossfolds for each simulated profile size.
-  At each crossfold we include only a constant number of ratings for each test user into the training set, ensuring that all test users have the specified profiles size.
+  To understand how metrics change with a different number of users we can simply generate separate random crossfolds for each simulated profile size.
+  At each crossfold we include only a constant number of training ratings for each test user, ensuring that all test users have the specified profiles size.
 
@@
   As we kept more ratings for the training set, we would have fewer ratings in the test set.
-  This turned out to bias our results; some metrics, such as nDCG, proved to be sensitive to the number of items in the test set.
+  This biased our results; some metrics such as nDCG proved to be sensitive to the size of the test set.
+  As we compared different profile sizes we would see trends that were solely a property of the changing test size.
 
@@
   Then, to generate any other number of ratings, we randomly downsample the 19 item training set.
-  This allows us to generate training sets with a given number of ratings for test users, such that profile size is evaluated with the same test set.
+  This allows us to generate training sets with a given number of ratings for test users, such that each profile size is evaluated with the same test set.
+
+ % Prediction accuracy
+\begin{figure*}[ht!]
+  \centering
+  \includegraphics[width=0.75\linewidth]{../lenskit/output/ekstrandTuned20/accuracy.pdf}
+  \caption{Accuracy and Ranking Algorithms}
+  \label{fig:rmse}
+  \label{fig:ndcg}
+\end{figure*}
 
@@
 
- % Prediction accuracy
-\begin{figure*}[ht!]
-  \centering
-  \includegraphics[width=0.75\linewidth]{../lenskit/output/ekstrandTuned20/accuracy.pdf}
-  \caption{Accuracy and Ranking Algorithms}
-  \label{fig:rmse}
-  \label{fig:ndcg}
-\end{figure*}
     % RMSE plot
@@
   This trend is a result of the limited coverage of the ItemItem algorithm.
-  Consider the accuracy of RMSE on only those items where it makes a personalized prediction, it shows a monotonic increase in accuracy similar to the other algorithms.
-  However, for small rating counts ItemItem can only predict for some percent of the item (around 60\% for users with two ratings, and around 75\% for users with four ratings).
+  Considering the accuracy of RMSE on only those items where it makes a personalized prediction, it shows a monotonic increase in accuracy similar to the other algorithms.
+  However, for small rating counts ItemItem can only predict for some percent of the items (around 60\% for users with two ratings, and around 75\% for users with four ratings).
   Because of this, many of the predictions come from the better performing UserItemBaseline.
@@
   The first thing to note about this graph is that all of these algorithms are doing a reasonable job of recommending items the user might like to watch, with all algorithms averaging above 4 stars (implying that the recommended items are, at least on average, enjoyable).
-  SVD performs as expected, starting around the baseline for few ratings, and then steadily improving to be the best algorithm after 3 ratings.
+  SVD performs as expected, starting around the baseline for a few ratings, and then steadily improving to be the best algorithm after 3 ratings.
   Again, we find that ItemItem performs poorly until quite a few ratings are added, and does worse than baseline throughout.
@@
 
-  One of the major limitations of our work is that we only evaluated one dataset.
-  Future work should replicate our experiment on other datasets from different recommender domains.
-  By doing this we will find out which of our results are true properties of the algorithms, and which results are specific to our dataset.
+  One of the limitations of our work is that we only evaluated on one dataset.
+  Future work should replicate our experiment on other datasets from different domains, or where ratings are collected in different ways.
+  This will establish which of our results are true properties of the algorithms, and which results are specific to our dataset.
 
-  One important property that needs to be considered is that our dataset only contained users that naturally used the system enough to collect 20 ratings.
+  One important issue that our methodology cannot address is that we can only compare users who have made some minimum number of ratings (20 for our dataset)
   Unfortunately, this means that we do not know if there are differences between users who do and do not survive in the system.
-  This is a limitation of our methodology, to maintain a constant test set throughout the evaluation we are not able to evaluate users who have less ratings than the maximum amount of ratings we wish to study.
-  More work needs to be done to understand users who only  make one or two ratings to discover why they leave, and how they differ from other users of our systems.
+  This is an unfortunate side effect of insisting on a constant test set throughout the evaluation.
+  More work needs to be done to understand users who only make one or two ratings to understand how they are different and discover why they leave the system.
 
-  Finally, we feel that more work is needed to explore different algorithms.
-  We only explored three of the dozens of known recommendation algorithms.
-  Some of these algorithms are designed specifically for cold start, others are designed to leverage extra, non-rating information to improve quality.
-  Work should be done to understand the differences between these algorithms, and what effect extra information has on the recommendation process.
-  Using our methodology it should be possible to measure the effect that extra information about the user has on recommendation, and to compare it against the effect of extra initial ratings.
+  Nonetheless, we feel that our methodology will allow offline comparisons of many different cold start experiences.
+  Our work has only scratched the surface of recommender algorithms.
+  Future work can be done comparing these standard algorithms against newer algorithms such as those designed for cold start, or those designed to leverage extra non-rating information.
+  More work should be done to understand how extra information about a user can effect cold start ratings.
+  Our methodology can be used to measure this effect, and even establish a tradeoff between the extra information or extra initial ratings.
     % algorithm limitations

### 2014-07-22 paper/main.tex
pass for some typos

@@
   Figure \ref{fig:map} shows MAP@20, Precision@20, and Fallout@20.
-  All three metrics show essentially the same trend, UserUser gets the lowest score, the baseline get the highest score.
+  All three metrics show essentially the same trend, UserUser gets the lowest score, the baseline gets the highest score.
   On both MAP@20 and Precision@20, SVD gets a higher score than ItemItem throughout, performing almost as well as the baseline.
@@
   One possible reason that the SVD algorithm performs as well as it does is the regularization built into the algorithm.
-  UserUser and ItemItem algorithms have no built in method for damping their predictions towards the mean when they don't have much data.
+  UserUser and ItemItem algorithms have no built in method for damping their predictions towards the mean when they do not have much data.
   The regularization in the cost function of the SVD model makes it so its predictions will gradually shift away from the baseline predictions as more relevant ratings are added.

### 2014-07-23 paper/main.tex
another proofreading and formatting pass

@@
 
-  The ideal way to study this issue would be a live study with real people who are new to a recommender system.
+  The ideal way to study this issue would be a live study conducted on people who are new to a recommender system.
   Users could be randomly assigned an algorithm and then surveyed after a random number of ratings to establish what they thought of the new user experience. 
@@
   This work comes out of a pushback against early recommender evaluations which only evaluated recommenders in terms of their predictive accuracy \cite{accurateIsNotEnough}.
-  The response of this pushback has been the generating of many different metrics which allow us to more carefully understand the separate aspects of a recommender's output \cite{LathiaTemporal, CremonesiTopN, zieglerDiversity}.
+  The response of this pushback has been the generation of many different metrics which allow us to more carefully understand the separate aspects of a recommender's output \cite{LathiaTemporal, CremonesiTopN, zieglerDiversity}.
   %TODO: I want to cite more papers here, I don't think I have good ones either. Joe would probably know.
@@
   Because new users are a scarce resource we opt to first explore the new user experience in an offline setting.
-  This allows us to test many more profile sizes and algorithms much faster than an online study.
+  This allows us to test many more profile sizes and algorithms than an online study.
   We will adapt the standard train and test methodology so that test users have only a specified number of training ratings.
@@
   We found this unnecessary for recommendations as each algorithm was able to make a complete recommendation with only one rating.
-  Therefore we form recommendations as the top 20 items that the algorithm was able to predict which were not in the training set.
+  Therefore we form recommendations as the top 20 items with predictions from the algorithm which were not in the training set. 
   More information about these algorithms can be found in table \ref{tbl:algo}.
@@
   Each algorithm was tuned based off of Ekstrand's 2012 evaluation \cite{ekstrand2012recommenders}.
-  Since this evaluation is on a different dataset (MovieLens 10M vs. MovieLens 1M), we performed minor tuning using the Ekstrand's values as a starting point.
+  Since this evaluation is on a different dataset (MovieLens 10M vs. MovieLens 1M), we performed minor tuning using Ekstrand's values as a starting point.
   We tuned based on RMSE performance.
@@
 
+  \vspace{-0.5em}
   \paragraph{Accuracy and Ranking Metrics}
@@
 
+  \vspace{-0.5em}
   \paragraph{Monotonic Recommendation Metrics}
@@
 
+  \vspace{-0.5em}
   \paragraph{Non-monotonic Recommendation Metrics}
@@
 
-  The {\bf AILS@20} metric is based on the Intra-List Similarity (ILS) metric introduced by Ziegler et al \cite{zieglerDiversity}.
+  The {\bf AILS@20} metric is based on the Intra-List Similarity (ILS) metric introduced by Ziegler et al. \cite{zieglerDiversity}.
   AILS@N is simply a rescaling of ILS to be scale free with regards to N, this was done to make the metric easier to interpret, by putting it on the same scale as similarity values.
@@
   When Spread is small, the recommender is focusing on a small set of items which it recommends to every user.
-  In this way spread is related to item coverage metrics.
-  Spread however looks at how often each item is recommended, not just whether or not it is recommended.
+  In this way spread is similar to an item coverage metric, but provides a more nuanced measure of item coverage  as it considers how often each item is recommended.
 
@@
 
+  \vspace{-0.5em}
   \paragraph{Expectations about algorithm performance}
@@
   This biased our results; some metrics such as nDCG proved to be sensitive to the size of the test set.
-  As we compared different profile sizes we would see trends that were solely a property of the changing test size.
+  As we compared different profile sizes we would see trends that were solely a property of the changing test set size.
 
@@
 
+\pagebreak
 \section{Results}
@@
     % NDCG
+  \vspace{-0.5em}
   \paragraph{nDCG}
@@
 
+  \vspace{-0.5em}
   \paragraph{SeenItems@20}
@@
 
+  \vspace{-0.5em}
   \paragraph{MeanRating@20}
@@
 
+  \vspace{-0.5em}
   \paragraph{RMSE@20}
@@
     % diversity plot
+  \vspace{-0.5em}
   \paragraph{AILS@20}
@@
   % spread plot
+  \vspace{-0.5em}
   \paragraph{Spread@20}
@@
       \hline
-      {\bf Algorithm} & {\bf Prediction accuracy} & {\bf Recommendation quality} & {\bf Other properties} \\\hline
+      { Algorithm} & { Prediction accuracy} & { Recommendation quality} & { Other properties} \\\hline
       ItemItem        & Poor                      & Poor                         & Good                   \\\hline

### 2014-07-23 paper/main.tex
updated with the acm text

@@
 
-\widowpenalty=3000
+\newfont{\mycrnotice}{ptmr8t at 7pt}
+\newfont{\myconfname}{ptmri8t at 7pt}
+\let\crnotice\mycrnotice%
+\let\confname\myconfname%
+
+\permission{Permission to make digital or hard copies of all or part of this work for personal or classroom use is granted without fee provided that copies are not made or distributed for profit or corecsysercial advantage and that copies bear this notice and the full citation on the first page.
+Copyrights for components of this work owned by others than the author(s) must be honored.
+Abstracting with credit is permitted.
+To copy otherwise, or republish, to post on servers or to redistribute to lists, requires prior specific permission and/or a fee.
+Request permissions from permissions@acm.org.}
+\conferenceinfo{RecSys'14,}{October 6--10, 2014, Foster City, Silicon Valley, CA, USA. \\
+{\mycrnotice{Copyright is held by the owner/author(s). Publication rights licensed to ACM.}}}
+\copyrightetc{ACM \the\acmcopyr}
+\crdata{978-1-4503-2668-1/14/10\ ...\$15.00.\\
+http://dx.doi.org/10.1145/2645710.2645742 }
+
+\clubpenalty=10000 
+\widowpenalty = 10000
 
@@
 \begin{document}
-\conferenceinfo{RecSys '14}{October 6--10, 2014, Foster City, Silicon Valley, USA}
-%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
-%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
 

### 2014-07-24 paper/main.tex
one last little change

@@
   Too many, and the user may feel that the recommendations are not interesting or useful.
-  Unfortunately, without more research on these metrics to establish a mapping between user satisfaction and metric values we will have to rely on educated guesses to establish which values for these metrics are too extreme.
+%  Unfortunately, without more research on these metrics to establish a mapping between user satisfaction and metric values we will have to rely on educated guesses to establish which values for these metrics are too extreme.
+  Unfortunately, the optimal value for these metrics will vary by domain, user, and context.
+  That said, we feel it is useful to understand how these metrics vary across algorithms for new users, as this knowledge can help inform default algorithm selection within a particular domain and context.
 
@@
   We would like to specially thank Michael Ekstrand for his  important work on the Lenskit recommendation framework.
-  This work is supported by NSF grants IIS1017697 and IIS0808692.
+This research was supported by the National Science foundation under grants IIS 08-08692 and IIS 10-17697,
+
   %netfishing and volunteers

### 2014-07-24 paper/embedFonts.sh
finally found a reliable command for embedding fonts

@@ -1,0 +1,1 @@
+gs -q -dNOPAUSE -dBATCH -dPDFSETTINGS=/prepress -sDEVICE=pdfwrite -sOutputFile=output.pdf main.pdf
